\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Static variables}

First let me talk about static variables in general (which has nothing
to do with classes.)

Static variables are declared before the program executes, and they stay
in memory until the program halts. Functions are static, because the
memory allocated for them is declared before the start of the program,
and is only reclaimed once the program ends. Global variables are also
static for the same reason.

Try the following program:

\begin{consolethree}[escapeinside=||]
#include <iostream>

void f()
{
    static int x = 0;
    int y = 0;
    x++; y++;
    std::cout << x << ' ' << y << std::endl;
}

int main()
{
    f(); f(); f(); f();
    return 0;
}
\end{consolethree}

What's happening??? Well ...

\begin{consolethree}[escapeinside=||]
void f()
{
    static int x = 0; // this is executed ONCE when
                      // the program starts up
                      // and x STAYS AROUND even when
                      // you return from f(). x goes
                      // away and only when the
                      // program ends. So x is NOT
                      // automatic.
    ...
}
\end{consolethree}

\begin{ex}
Write a function \texttt{mybankaccount()} such that the
first time you call \texttt{mybackaccount()}, it prints 0. If you call it
the second time with 10, the function prints 10. If you call it the
third time with 20, it prints 30. If you call it the fourth time with
-5, it prints 25. In other words, it records the cumulative sum of the
arguments sent into the function. Note that the first function call does
not pass in a value while the remaining ones do. That tells you that
\texttt{mybankaccount()} has a parameter with a default value. For
instance you can use -9999 as the default value to mean ``reset the
cumulative sum to 0''.
\end{ex}


