\begin{python0}
from solutions import *; clear()
\end{python0}
\input{../exercises/42-10-for-loops-part1/question.tex}\input{../exercises/43-10-for-loops-part1/question.tex}\sectionthree{Using for-loop to compute products}

\input{../exercises/44-10-for-loops-part1/question.tex}
What is the output (or is there an error)? Now run the program with your
C++ compiler and verify.

By the way, the product of all positive integer from 1 to 5 (i.e.
1x2x3x4x5) is called the \EMPHASIZE{factorial} of 5. The factorial of 10 is
just
\begin{center}
1 x 2 x 3 x 4 x 5 x 6 x 7 x 8 x 9 x 10
\end{center}
In general, the factorial of n is such an important quantity that we
have a shorthand notation for it: The factorial of n is written n!. So
\begin{center}
10! = 1 x 2 x 3 x 4 x 5 x 6 x 7 x 8 x 9 x 10
\end{center}
Note that the factorial of 0 is defined to be 1.
\begin{center}
0! = 1
\end{center}
Study the above code. Do you see that it computes the factorial of 5?

The factorial appears in Computer Science, Math, Physics, Chemistry,
Finance, etc. So you bet it's important.

\input{../exercises/45-10-for-loops-part1/question.tex}\input{../exercises/46-10-for-loops-part1/question.tex}
\input{../exercises/47-10-for-loops-part1/question.tex}
\input{../exercises/48-10-for-loops-part1/question.tex}
\input{../exercises/49-10-for-loops-part1/question.tex}
\input{../exercises/50-10-for-loops-part1/question.tex}\input{../exercises/51-10-for-loops-part1/question.tex}\begin{center}
n! / (r! (n - r)!)
\end{center}
n-choose-r is the total numbers of ways to choose r things from n
distinct things. For instance, suppose you are an IT manager and have a
team of 40 programmers, and you have to form a team of 5 software
engineers. The total numbers of ways to form this team is 40-choose-5,
i.e.,
\begin{center}
40! / (5! x (40 - 5)!) = 40! / (5! x 35!)
\end{center}
However 40! is too huge to compute. Use your program to compute
5-choose-2. Verify the correctness of your program by doing this
computation by hand. Here are 5 symbols:

\begin{center}
a, b, c, d, e
\end{center}
Write down all possible selections of 2 symbols. Let me begin for you:
\begin{align*}
\text{a, b}\\
\text{a, c}
\end{align*}

Note that the selection of a, b is considered the same as the selection of b, a. So do not include b, a. Likewise only include b, e and not e, b.

