\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Using for-Loop to compute min/max}

Recall that if you have int variables a, b, c, d, and you want to
compute the minimum, you can do this (pseudocode only!):
\begin{console}
min = a
if b < min:
   min = b
if c < min:
   min = c
if d < min:
   min = d
\end{console}
In this case min is an integer variable. (Re-study your notes on the if
statement if you don't recall this!!!)

The idea of course works for doubles too. Of course the min variable
must be declared to be a double variable. Make sure you understand the
above!

The idea for the computation of the maximum is similar.

Now the above sure looks like a loop. If the values are not stored in
variables a, b, c, d but for instance is supplied by the user (so we
need to do std::cin), then we have a very compact way to compute the
minimum of 4 values supplied by the user. Instead of doing
\begin{console}
prompt user for a
min = a

prompt user for b
if b < min:
   min = b

prompt user for c
if c < min:
   min = c

prompt user for d
if d < min:
   min = d

print min
\end{console}
we can do this:

\begin{console}
prompt user for x
min = x

// 3 more values
for i = 0, 1, 2:
    prompt user for x
    if x < min:
       min = x
       
print min
\end{console}

(user inputs are all stored using x).

The C++ code is then:

\begin{console}
int x;
int min;

std::cin >> x;
min = x;

// 3 more values, so i = 0, 1, 2 works.
for (int i = 0; i < 3; i++)
{   
    std::cin >> x;
    if (x < min)
    {
        min = x;
    }
}
std::cout << min << std::endl; 
\end{console}
Not only that, this code can be EASILY modified to compute the minimum
of 10 values:

\begin{console}
int x;
int min;

std::cin >> x;
min = x;

// i = 0, 1, 2, 3, 4, 5, 6, 7, 8 --- 9 more values
for (int i = 0; i < 9; i++)
{   
    std::cin >> x;
    if (x < min)
    {
        min = x;
    }
}
std::cout << min << std::endl;
\end{console}

The old method without loops would take a lot more time to write!

Not only that, the above method works when you allow the user to specify
how many values the user wants to input!!

\begin{console}
int x;
int min;
int n; // number of values for the min computation

std::cin >> n;
std::cin >> x;
min = x;

for (int i = 0; i < n - 1; i++)
{   
    std::cin >> x;
    if (x < min)
    {
        min = x;
    }
}

std::cout << min << std::endl;
\end{console}

As you can see, the loop is indispensable!!!
\input{temp/10-for-loops-part1/exercises/52-10-for-loops-part1/question.tex}\input{temp/10-for-loops-part1/exercises/53-10-for-loops-part1/question.tex}\begin{center}
$|42| = 42$

$|-24| = 24$
\end{center}

The absolute value functions were mentioned in the set of notes on
doubles. For your convenience, here's the information
again. For C++, if you need the absolute value of an integer value this
is how yo do it:
\begin{console}
#include <iostream>
#include <cstdlib>

int main()
{   
    int x = 42, y = -24;
    std::cout << abs(42) << '\n';
    std::cout << abs(-42)<< '\n'; 
    std::cout << abs(x) << '\n'; 
    std::cout << abs(y) << '\n';
    return 0;
}
\end{console}

If you need to compute the absolute value of doubles or floats you do
this:
\begin{console}
#include <iostream>
#include <cmath>
int main()
{   
    int x = 4.2, y = -2.4;
    std::cout << fabs(42) << '\n';
    std::cout << fabs(-42)<< '\n'; 
    std::cout << fabs(x) << '\n'; 
    std::cout << fabs(y) << '\n';
}
return 0;
\end{console}


