\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Increment and decrement operators}\\\\
You have already seen the +, -, *, /, \% operators for numeric values.
Now for a few more.
\input{exercises/0-10-for-loops-part1/main.tex}Change the initial value of \textbf{i} to any other number and run the program again.

The ++ and -- are the increment and decrement operators respectively. When you apply ++ \textbf{in front} of the variable, you are using the \EMPHASIZE{pre-increment} operator:\\
\begin{center}
  \texttt{++i;}
\end{center}

And when you use the ++ \textbf{after} the variable, you are using the \textbf{post-increment} operator.
\begin{center}
  \texttt{i++;}
\end{center}

Although they both add \textbf{1} to \texttt{i}, there is a difference. First of all, when you apply ++ to a variable, not only is the value of the variable changed, a copy of the value is returned. The next two exercises will make
everything clear.
\input{exercises/1-10-for-loops-part1/main.tex}
\input{exercises/2-10-for-loops-part1/main.tex}Make sure to compare the above two programs!!!
In other words, the difference is in the \EMPHASIZE{order} of incrementing and giving the value of the variable

The same applies to the pre- and the post-decrement operators.
\input{exercises/3-10-for-loops-part1/main.tex}\input{exercises/4-10-for-loops-part1/main.tex}
So now you know three different ways to add 1 to variable i:
\begin{center}
 \texttt{i = i + 1;}        \texttt{i++;}        \texttt{++i;}
\end{center}

Choices are nice, right? Well, except for people who cant make up their minds. Here's a hint: the increment operators are faster.
Another thing. Do not try to confuse yourself (or others) by using the increment and decrement operators in an expression. For instance the following is \EMPHASIZE{BADDDDDDD} although it does run:
\begin{center}
\texttt{a = (++i) + (k--) * (++j);}
\end{center}
This is the same as:
\begin{align*}
&\texttt{++i;}\\
&\texttt{++j;}\\
&\texttt{a = i + k * j;}\\
&\texttt{--k;}\\
\end{align*}
which is much easier to read.

\input{exercises/5-10-for-loops-part1/main.tex}
Now verify with your C++ compiler. Explain to yourself what happens when
the pre-increment operator is applied multiple times to the same
variable in an expression. Now do the same with this:
\begin{console}
int i = 0;
int j = (i++) + (i++);
std::cout << i << ' ' << j << '\n';
\end{console}

\input{exercises/6-10-for-loops-part1/main.tex}Let me summarize what we now know about the pre- and post-increment
operators:
\begin{flushleft}
\texttt{++i}\tab[3em]{\text{increments the value of \texttt{i}}}\\
\texttt{i++}\tab[3em]{\text{increments the value of \texttt{i}}}\\
\texttt{j=(++i)}\tab[1em]{\text{increments the value of \texttt{i} and then assign new value of \texttt{i} to \texttt{j}}}\\
\texttt{j=(i++)}\tab[1em]{\text{give the value of \texttt{i} to \texttt{j} and then increments the value of \texttt{i}}}\\
\end{flushleft}
and here' s the summary for pre- and post-decrement
operators:
\begin{flushleft}
\texttt{--i}\tab[3em]{\text{decrements the value of \texttt{i}}}\\
\texttt{i--}\tab[3em]{\text{decrements the value of \texttt{i}}}\\
\texttt{j=(--i)}\tab[1em]{\text{decrements the value of \texttt{i} and then assign new value of \texttt{i} to \texttt{j}}}\\
\texttt{j=(i--)}\tab[1em]{\text{give the value of \texttt{i} to \texttt{j} and then decrements the value of \texttt{i}}}\\
\end{flushleft}
