\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Augmented assignment operators}

\input{exercises/7-10-for-loops-part1/main.tex}

\input{exercises/8-10-for-loops-part1/main.tex}
\input{exercises/9-10-for-loops-part1/main.tex}
In general, suppose \emph{\texttt{[op]}} is an operator such as +, --, *, /,
\%, then
\begin{center}
\texttt{x [op]= y}
\end{center}
is the same as
\begin{center}
\texttt{x = x [op] y}
\end{center}
For instance,
\begin{align*}
\texttt{x \%= y + 5} && \text{is the same as} && \texttt{x = x \% (y + 5)}\\
\texttt{y /= c} && \text{is the same as} && \texttt{y = y / c}\\
\end{align*}
etc. Note that the left hand version (using the augmented assignment
operators) is faster than using an operator and then assignment.

\input{exercises/10-10-for-loops-part1/main.tex}
Now verify with your C++ compiler.

\input{exercises/11-10-for-loops-part1/main.tex}
Verify with your compiler.


\input{exercises/12-10-for-loops-part1/main.tex}
Verify with your compiler.


\input{exercises/13-10-for-loops-part1/main.tex}
By the way, you now have \emph{\textbf{four}} different ways to add one
to variable \emph{i}:
\begin{align*}
&\texttt{++i;} &&\texttt{i++;} &&\texttt{i += 1;} &&\texttt{i = i + 1;}
\end{align*}
(C++ programming is terrible for people who can't decide. Right?) In general, the operators on the left are faster than the ones to their right.

\input{exercises/14-10-for-loops-part1/main.tex}

\input{exercises/15-10-for-loops-part1/main.tex}

\input{exercises/16-10-for-loops-part1/main.tex}
\input{exercises/17-10-for-loops-part1/main.tex}
\input{exercises/18-10-for-loops-part1/main.tex}Now verify with your C++ compiler. You should \emph{\textbf{never}} write such an expression in actual programs. And if you join a company where programmers write such code ... well ... good luck to you!!!

\input{exercises/19-10-for-loops-part1/main.tex}
\input{exercises/20-10-for-loops-part1/main.tex}
\input{exercises/21-10-for-loops-part1/main.tex}
\input{exercises/22-10-for-loops-part1/main.tex}
