\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Containers}

We're already talked about arrays (arrays of ints, arrays of doubles, etc.) Usually we use all the slots in the arrays. However an array can be used to model \EMPHASIZE{containers}. What do I mean by that? For instance suppose you want to write a program to contain student IDs of current students who are at least 7 feet tall. (Not a big list I suppose \ldots) In fall 2009 the list might contain:

\begin{center}
3145673, 4135778, 2356317
\end{center}

in spring 2009 it might be

\begin{center}
3145673, 4135778, 2356317, 5678246
\end{center}

and in fall 2010 it might be

\begin{center}
2356317, 5678246
\end{center}

As you can see the number of things in the list changes.

An array can be used to model a container. It can be done in two ways (at least!) You can keep an extra variable to indicate the number of things modeled by the array. For instance try this program that prompts the user for ids to be added to the id array and terminates when the user enters -1.

\begin{consolethree}[escapeinside=||]
int length = 0;
int id[10];

while (1)
{
    if (length == 10)
    {
        std::cout << "oops ... I'm full!\n";
        break;
    }
    int x;
    std::cin >> x;
    if (x == -1) break;
    id[length] = x;
    length++;
    for (int i = 0; i < length; i++)
    {
        std::cout << id[i] << ' ';
    }
    std::cout << '\n';
}
\end{consolethree}

In this case \texttt{\EMPHASIZE{length}} is used to indicate that the actual student ids are

\begin{center}
\textit{id[0], id[1], \ldots, id[length-1]}
\end{center}

and the rest

\begin{center}
\textit{id[length], id[length+1], \ldots, id[999]}
\end{center}

are unused.

A second way to model a container is to use a special \EMPHASIZE{sentinel} value to denote \EMPHASIZE{``end-of-data''}. Try this program that does the same thing as above except that I'm using -9999 to mark the end-of-data.

\begin{consolethree}[escapeinside=||]
int id[10] = {-9999};

while (1)
{
    // find length
    int length = 0;
    for (int i = 0; i < 10; i++)
    {
        if (id[i] == -9999)
        {
            length = i;
            break;
        }
    }
    if (length == 9)
    {
        std::cout << "oops ... I'm full!\n";
        break;
    }
    int x;
    std::cin >> x;
    if (x == -1) break;
    id[length] = x;
    id[length + 1] = -9999;
    // print values in container
    int i = 0;
    while (id[i] != -9999)
    {
        std::cout << id[i] << ' ';
        i++;
    }
    std::cout << '\n';
}
\end{consolethree}

Note that the array begins with

\begin{center}
-9999, \ldots{}
\end{center}

(the \ldots{} after the -9999) If I entered 42, the array becomes

\begin{center}
42, -9999, \ldots{}
\end{center}

And if I entered 135 as the next number, the array becomes

\begin{center}
42, 135, -9999, \ldots{}
\end{center}

Note that when using a sentinel to implement a container

\begin{itemize}
\item
When adding the end of the container you always need to search for the sentinel value (more or less to compute the length of the container)
\item
The number of values you can put into the container is always \EMPHASIZE{one less than the actual size} of the container. Why? Because the sentinel value takes up on of the spots in the array!
\end{itemize}

We'll be talking about C-strings in this set of notes. You'll see that the C-string uses the second method to describe characters in a string.


