\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Other C-string functions}

In fact there are MANY standard string functions that come with all C compilers. Here are some of the most basic ones. Experiment with this code and figure out what the string functions do:

\begin{consolethree}[escapeinside=||]
char s[100] = "aaa";
char t[100] = "bbb";
char u[100];

// string copy function
// overwrite the string in u with the string in
// s
strcpy(u, s);
std::cout << u << std::endl;

// string concatenation function
// put the string in t to the end of the string
// in u
strcat(u, t);
std::cout << u << std::endl;

// WARNING: You cannot do strcat(u, u), i.e., the
// two arguments of the function must be different.

// string comparison function
// 0 is returned when the strings are
// the same. Otherwise a nonzero value is
// returned.
int i = strcmp(u, s);
std::cout << i << std::endl;
\end{consolethree}

You might wonder what nonzero value is returned by \texttt{strcmp()} when the strings are different.

\begin{consolethree}[escapeinside=||]
std::cout << strcmp("a", "b") << std::endl;
std::cout << strcmp("a", "c") << std::endl;
std::cout << strcmp("a", "d") << std::endl;
std::cout << strcmp("a", "z") << std::endl;
std::cout << strcmp("a", "!") << std::endl;
std::cout << strcmp("a", "ba") << std::endl;
std::cout << strcmp("b", "a") << std::endl;
std::cout << strcmp("c", "a") << std::endl;
std::cout << strcmp("d", "a") << std::endl;
std::cout << strcmp("ba", "a") << std::endl;
\end{consolethree}

See it yet? (Hint: Check an ASCII table.)

There are other C-string functions but you will need to know a little more in order to understand how to use them.


