\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{The \texttt{strnlen(), strncmp(), strncat()} functions}

All the string functions we talked about run into problems if the character arrays passed into these function do not contain \texttt{'\textbackslash 0'} resulting in buffer overflow.

In fact this is one scenario of the so-called \EMPHASIZE{buffer overflow attack}, a technique used in many virus software (malware).

Accompanying these functions are similar functions that allow you to specify a stopping point. For instance in the case of \verb!strlen()! (the string length function) there is a function called \verb!strnlen()!

Besides passing in a string, you also pass in a maximum possible length
for the string.

\begin{consolethree}[escapeinside=||]
char s[100] = ""; // s[0] = '\0';
std::cin >> s;
std::cout << strnlen(s, 99) << std::endl;
\end{consolethree}

Note that I pass in 99 and not 100. If s has a \EMPHASIZE{size of 100} the \EMPHASIZE{maximum possible length is 99} since in this case the \EMPHASIZE{last character is used by '\textbackslash 0'}. Therefore \texttt{s} (of size 100) has a maximum length of 99.

\input{../exercises/12-18-characters-and-cstrings/question.tex}
The accompanying safer functions for strcmp() and strcat() are strncmp() and strncat(). So we have the following \textbf{\underline{pairs}} of functions:

\begin{tabular}{ll}
strlen() & strnlen() \\
strcmp() & strncmp() \\
strcat() & strncat() \\
\end{tabular}

Try this program:

\begin{consolethree}[escapeinside=||]
char s[100] = "aaa";
char t[100] = "bbb";
char u[100];

// string copy function
// In this case at most 100 chars are copied
// from s to u
strncpy(u, s, 100);
std::cout << u << std::endl;

// string concatenation function
// In this case at most 100 chars are
// concatenated from t to u
strncat(u, t, 100);
std::cout << u << std::endl;

// string comparison function
// In this case at most 100 characters are
// compared.
int i = strncmp(u, s, 100);
std::cout << i << std::endl;
\end{consolethree}

The only problem in the above example is the call to strncat() function. You see if \texttt{u} already has a string of length 10, then only 90 characters are left for concatenation. If \texttt{t} has length greater than that, the program might crash. Therefore it's safe to write this:

\begin{consolethree}[escapeinside=||]
// string concatenation function
// In this case at most 100 chars are
// concatenated from t to u
strncat(u, t, 100 - strnlen(u, 100));
std::cout << u << std::endl;
\end{consolethree}

But in this case some character might not be copied from \texttt{t} to \texttt{u}. Depending on what your program is supposed to do, you might want to print an error message and halt the program.


