%-*-latex-*-
\input{myquizpreamble.tex}
\input{ciss240.tex}
\input{yliow}
\renewcommand\TITLE{Quiz q0101}

\begin{document}
\topmatter

This is a closed book quiz. 
You have 5 minutes (that is from beginning to the end including
the submission of your work.)

If you are using Microsoft Visual Studio, you must clear all the
auto-generated code and write the program on your own.

Q1. The goal is to write a \cpp\ program.
The name of your cpp file must be \verb!q0101.cpp!.

The FIRST FEW LINES of your \cpp\ source file must look like this:
\begin{console}
// Name: John Doe
// File: q0101.cpp

#include <iostream>
\end{console}
with \lq\lq John Doe''replaced by your name (of course).

Write a \cpp\ program that produces the following output in the console
window (when you execute the program):
\begin{console}
hello master ...
\end{console}
Your output must following the above \textit{exactly} as given.
For instance,
your program is \textit{wrong} if the output is
\begin{console}
Hello master ...
\end{console}
or
\begin{console}
hello master.
\end{console}
or
\begin{console}
hello    master ...
\end{console}
or any other variation.

Your coding style must following the coding style as used in class.
That includes the spacing, the blank lines, etc.

\textsc{Grading.}
\begin{tightlist}
\item If your program was not received in time: 0/2
\item If your program contains error(s) and does not run: 0/2
\item If your program is error-free, does run, but produces no output: 0/2
\item If your program is error-free, does run, produces an output, but the output does not match the given output: 0/2
\item If your program is error-free, does run, and produces an output that matches
the given output: 2/2
\item After the points from above,
point(s) will be taken off for incorrect coding style.
\end{tightlist}
\end{document}
