\newpage\EMPHASIZE{91. Namespace}

\textsc{Objectives}

\begin{itemize}
\item
  Create a namespace
\item
  Place identifiers in a namespace
\end{itemize}

\newpage\EMPHASIZE{Namespace}

Run this:
\begin{console}
#include <iostream>
namespace yliow
{    
     int x;
     void f()
     {    
          std::cout << "hello world" <<
          std::endl;
     }
     
     class C
     {};
}

int main()
{   
    yliow::x = 42;
    yliow::f();
    yliow::C obj;
    return 0;
}
 
\end{console}

A \EMPHASIZE{namespace} is just a container of names (i.e., variables, functions, structs, classes.)

Why?

Allows you to put names into a namespace to avoid naming conflicts.

For instance suppose John and Mary work on the same project. They can choose different namespaces for their own things:

\begin{itemize}
\item
  \verb!john::print_helloworld()!
\item
  \verb!mary::print_helloworld()!
\end{itemize}

\newpage\EMPHASIZE{Openness of namespace}

Namespaces are \EMPHASIZE{open}: you can open it after it's closed.
\begin{console}
namespace yliow
{    
     class C
     {};
}

namespace yliow // open again!!!
{    
     class D
     {};
}
int main()
{   
     yliow::C obj1;
     yliow::D obj2;
     return 0;
}
 
\end{console}

The most common usage of namespace is probably to hold classes and function prototypes:

\begin{console}
// GameLib.h
namespace GameLib
{    
     class vec2d { ... };
     class graphics { ... };
     class sound { ... }; 
     class physics { ... };
}
               
\end{console}

If you' re writing games and scientific simulations. You
can split the above.

\begin{console}
// Physics.h
namespace GameLib
{    
     class vec2d { ... };
     class physics { ... };
} 
\end{console}

\begin{console}
// GameLib.h
#include "Physics.h"
namespace GameLib
{    
     class graphics { ... };
     class sound { ... };
} 
\end{console}

In this way, when you're writing a text-based sci simulations, you don't need to compile with the graphics and sound class.

\newpage\EMPHASIZE{Using}

You can avoid typing a namespace when you use it:

\begin{console} 
namespace yliow
{    
     class X {};
     class Y {};
}

using namespace yliow;
int main()
{    
     X obj; // do not need yliow::X
     return 0;
}
\end{console}
        
