
For CISS245, I also emphasize rigorous software testing. Much of the testing
strategies that you will learn in this class (although in a small scale) is
actually being practiced in the real software engineering world. Specifically,
for this and future assignments I will emphasize the testing of the smallest
units of a software such as functions. These are called \textbf{unit tests}.


\newpage
\textsc{Structure Variables}

A structure variable is just a variable that contains variables (refer to your
notes). Run this:
\begin{Verbatim}[frame=single,fontsize=\small]
#include <iostream>
#include <iomanip>

struct Time
{
    int hour;
    int min;
    int sec;
}

int main()
{
    Time t0;
    t0.hour = 5;
    t0.min = 18;
    t0.sec = 0;

    std::cout << std::setw(2) << std::setfill('0') << t0.hour
              << ':'
              << std::setw(2) << std::setfill('0') << t0.min
              << ':'
              << std::setw(2) << std::setfill('0') << t0.sec
              << std::endl;

    return 0;
}
\end{Verbatim}

In this case, I want to work with time which is represented by hour, minute,
and second. However, I frequently want to view hour, minute, and second not as
three things but one. So I create this new type:
\begin{console}[fontsize=\small]
struct Time
{
    int hour;
    int min;
    int sec;
}
\end{console}

I can then create a \verb!Time! (structure) variable like this:
\begin{console}[fontsize=\small]
Time t0;
\end{console}

And now \verb!t0! contains three variables. \verb!t0! has:
\begin{tightlist}
  \li \verb!t0.hour! which is an \verb!int! variable.
  \li \verb!t0.min! which is an \verb!int! variable.
  \li \verb!t0.sec! which is an \verb!int! variable.
\end{tightlist}

The variables inside a structure variable are usually called
\textbf{member variables}.

\textsc{Memory model}

Here's the memory model of \verb!main()! for the variable \verb!t0! in the
above example:
\begin{python}
from latextool_basic import *
p = Plot()
t0_label = Rect(x0=-0.5, y0=0.5, x1=0.5, y1=1.5, linewidth=0.0, label=r"\texttt{t0}")
t0 = array(x0=0.5, y0=0.5, width=2, height=1, xs=[5,18,0])
main_label = Rect(x0=-0.5, y0=3.0, x1=0.5, y1=3.5, linewidth=0.0, label=r"\texttt{main}")
main = Rect(x0=-0.5, y0=-1.0, x1=7.0, y1=3.0, linewidth=0.05)
p.add(main_label)
p.add(main)
p.add(t0_label)
p.add(t0)

print(p)
\end{python}

(Like I said, \verb!t0! is just a variable containing variables.)

Here's another example of a structure type:
\begin{console}
struct Student
{
    int id;
    double gpa;
}
\end{console}

In this case, a \verb!Student! variable contains $2$ variables. For instance,
after this declaration:
\begin{console}
Student jdoe;
\end{console}
the variable \verb!jdoe! contains:
\begin{tightlist}
  \li \verb!jdoe.id!, an \verb!int! variable and
  \li \verb!jdoe.gpa!, a \verb!double! variable.
\end{tightlist}

Note that the variables in a structure can have different types. In the
\verb!Student! case, there is one \verb!int! variable and one \verb!double!
variable.

Note also that to go into a variable inside a structure variable, you use the
dot. For instance, to go to the hour of \verb!t0!, you use:
\begin{console}
t0.hour
\end{console}

The variables inside a structure variable are just like any regular variable so
you do know how to work with them (input, output, operators etc.) For instance,
back in the first \verb!Time! example, we assign values to the variable
\verb!t0!:
\begin{console}
t0.hour = 5;
t0.min = 18;
t0.sec = 0;
\end{console}

We also read the values in \verb!t0! and print them:
\begin{console}
std::cout << std::setw(2) << std::setfill('0') << t0.hour
          << ':'
          << std::setw(2) << std::setfill('0') << t0.min
          << ':'
          << std::setw(2) << std::setfill('0') << t0.sec
          << std::endl;
\end{console}

(By the way, although a structure variable is sort of like an array, you see
from the above that there are differences. For instance, an array can only
contain values of the same type. You can't have an array of $2$ integers and
$3$ doubles. An array is an \verb!int! array or a \verb!double! array. You
can't have it both ways! To go into an array, you have the bracket \verb![]!
with an index value. You get into a structure using the dot and a name for the
variable inside the structure that you're trying to access.)

\textsc{Initialization}

Instead of doing this:
\begin{console}
Time t0;
t0.hour = 5;
t0.min = 18;
t0.sec = 0;
\end{console}
to initialize structure variables, you can use the same notation as in array
initialization:
\begin{console}
Time t0 = {5, 18, 0};
\end{console}




\textsc{Functions: struct parameters}

Just like variables of basic types, it's not surprising that you can pass
structure variables into a function. Run this:
\begin{console}[commandchars=\\\@\$]
#include <iostream>
#include <iomanip>

struct Time
{
    int hour;
    int min;
    int sec;
};

\textbf@void print(Time t)$
\textbf@{$
    \textbf@std::cout << std::setw(2) << std::setfill('0') << t.hour$
              \textbf@<< ':'$
              \textbf@<< std::setw(2) << std::setfill('0') << t.min$
              \textbf@<< ':'$
              \textbf@<< std::setw(2) << std::setfill('0') << t.sec$
              \textbf@<< std::endl;$
\textbf@}$

int main()
{
    Time t0 = {5, 18, 0};
    \textbf@print(t0);$

    return 0;
}
\end{console}

Structure variables in function calls are by default pass--by--value. This
means that the variable inside the function cannot change the variable in the
calling function. Try this:
\begin{console}[commandchars=\\\@\$]
...

\textbf@void inc(Time t)$
\textbf@{$
    \textbf@t.sec++;$
\textbf@}$

...

int main()
{
    ...
    \textbf@inc(t0);$
    \textbf@print(t0); // t0.sec is the same!!!$

    return 0;
}
\end{console}

The variable \verb!t0! is not changed when you return from the function call to
\verb!inc()!. So, \textbf{if you do want to change \texttt{t0}}, you have to force a
pass--by--reference like this:
\begin{console}[commandchars=\\\@\$]
...

void inc(Time \underline@\textbf@&$$ t)
{
    t.sec++;
}

...

int main()
{
    ...
    inc(t0);
    print(t0); // t0.sec has changed!!!

    return 0;
}
\end{console}
So, in terms of parameter--passing, structure variables are like \verb!int! and
\verb!double! parameters which are also pass--by--value.




\textsc{Function: Reference--to--struct parameters}

For performance reasons (i.e., speed), we force all \textbf{structure variables
to be passed by reference}. Make this change and run your program again:
\begin{console}[commandchars=\\\@\$]
void print(Time \textbf@\underline@&$$ t)
{
    std::cout << std::setw(2) << std::setfill('0') << t.hour
              << ':'
              << std::setw(2) << std::setfill('0') << t.min
              << ':'
              << std::setw(2) << std::setfill('0') << t.sec
              << std::endl;
}
\end{console}
This avoids creating memory for \verb!t!; \verb!t! simply references the
variable in the calling function.

If a \verb!struct! parameter is a reference, the function can change the
variable in the calling function:
\begin{console}[commandchars=\\\@\$]
void print(Time \textbf@\underline@&$$ t)
{
    std::cout << std::setw(2) << std::setfill('0') << t.hour
              << ':'
              << std::setw(2) << std::setfill('0') << t.min
              << ':'
              << std::setw(2) << std::setfill('0') << t.sec
              << std::endl;
    \textbf@// This will change the Time variable that t references!!!$
    \textbf@t.sec = 0;$
}
\end{console}

This means that the \verb!print()! function now has the ability to change the
variable in the calling function. To avoid accidentally changing this variable,
we force the parameter \verb!t! to be constant:
\begin{Verbatim}[frame=single, commandchars=\\\@\$]
void print(\textbf@\underline@const$$ Time \textbf@\underline@&$$ t)
{
    std::cout << std::setw(2) << std::setfill('0') << t.hour
              << ':'
              << std::setw(2) << std::setfill('0') << t.min
              << ':'
              << std::setw(2) << std::setfill('0') << t.sec
              << std::endl;
}
\end{Verbatim}

So, let me repeat myself:

All stucture variables must be passed by reference (it's also possible to pass
the address of the \verb!struct! variable to a pointer receiving the address --
see later.) If a function wants to change the variable in the calling function,
there's nothing else to do. If a function should not change the variable in the
calling function, make the parameter \verb!const! as well.




\textsc{Functions: struct return value}

You can return a structure value:
\begin{console}[commandchars=~!@]
...

~textbf!Time addOneHour(const Time & t)@
~textbf!{@
~textbf!    Time newtime = t;@
~textbf!    newtime.hour++;@

~textbf!    return newtime;@
~textbf!}@

int main()
{
    ...

    ~textbf!Time t1 = addOneHour(t0);@
    ~textbf!print(t1);@

    return 0;
}
\end{console}

(So, structure variables are also different from arrays because you can return
them. Recall that you cannot return arrays.)




\textsc{Functions: pointer--to--struct parameters}

As mentioned earlier, it's also possible to pass the address of a
\verb!struct! variable. Try this version:
\begin{console}[commandchars=~!@]
void print(~textbf!~underline!const@@ Time * t)
{
    std::cout << std::setw(2) << std::setfill('0') << t->hour 
              << ':'
              << std::setw(2) << std::setfill('0') << t->min 
              << ':'
              << std::setw(2) << std::setfill('0') << t->sec
              << std::endl;
}

... 

int main()
{
    Time t0;
    t0.hour = 5;
    t0.min = 18;
    t0.sec = 0;
    print(~textbf!~underline!&@@t0);

    return 0;
}
\end{console}
where \verb!t->hour! is really just \verb!(*t).hour!.

The corresponding \verb!addOneHour()! function looks like this:
\begin{Verbatim}[frame=single, commandchars=~!@]
Time addOneHour(const Time * t)
{
    Time newtime = *t;
    newtime.hour++;

    return newtime;
}

int main()
{
    ...
    Time t1 = addOneHour(~textbf!~underline!&@@t0);

    return 0;
}
\end{Verbatim}

Let me summarize:
\begin{itemize}
  \li If you want a function to work with a \verb!struct! value, instead of
pass--by--value, you should use either pass--by--reference or pass the address,
i.e., the receiving parameter in the function must be a reference or a pointer.
  \li If the parameter in the function should not change the \verb!struct!
value from the calling function, then the parameter in the function must be a
constant reference or a pointer to a \verb!const! value.
\end{itemize}




\newpage
\textsc{Now for the assignment}

The goal is to build a simple library for computing fractions and test it
rigorously.

To make it easy for you, I have written it in parts.

The questions build upon each other. This means that subsequent questions are
going to rely significantly on previous questions.
Once you are done with a question, you make a copy of that folder
for the next question and continue work on the next folder.
For instance if you are done with question 3, you make a copy of the
folder for question 3 and rename it appropriately for question 4 and
continue work on the folder for question 4. Etc.



\newpage
\textsc{The Given Code Base}

Note that C\texttt{++}
does not natively support fractions. (Doubles are not fractions.)
In this assignment we will be developing useful functions and operators to
support the use of fractions. I'll give you the structure definition:
\begin{console}
struct Fraction
{
    int n; // numerator
    int d; // denominator
};
\end{console}

This means of course that if you want to model the mathematical fraction
\[x = \frac{3}{4}\]
you would do this in your program:
\begin{console}
Fraction x;
x.n = 3;
x.d = 4;
\end{console}
or better:
\begin{console}
Fraction x = {3, 4};
\end{console}

You will need to write three files:
\begin{itemize}
  \li \verb!test_fraction.cpp!: This contains a program to test the
      \verb!Fraction struct! and its supporting functions and operators.
  \li \verb!Fraction.h!: This is the header function containing the
      \verb!Fraction struct! and its prototypes.
  \li \verb!Fraction.cpp!: This file contains the definition of the prototypes
      in \verb!Fraction.h!.
\end{itemize}

The following are the skeleton files (remember that skeleton files
are incomplete and might contain errors):
{\small
\begin{Verbatim}[frame=single]
// Author: 
// Date  : 
// File  : test_fraction.cpp

#include <iostream>
#include "Fraction.h"


void test_print()
{
    int n = 0, d = 0;
    std::cin >> n >> d;
    Fraction f = {n, d};
    std::cout << f << std::endl;
}


int main()
{
    int option;
    std::cin >> option;
 
    switch (option)
    {
        case 1:
            test_print();
            break;
    }

    return 0; 
}
\end{Verbatim}
}

(Note: the tester uses option $1$ to test the print feature.)

\begin{Verbatim}[frame=single]
// Author: 
// Date  : 
// File  : Fraction.h

#ifndef FRACTION_H
#define FRACTION_H

#include <iostream>

struct Fraction
{
    int n; // numerator
    int d; // denominator
};


std::ostream & operator<<(std::ostream &, const Fraction &);

#endif
\end{Verbatim}

\begin{Verbatim}[frame=single]
// Author: 
// Date  : 
// File  : Fraction.cpp

#include <iostream>
#include "Fraction.h"


std::ostream & operator<<(std::ostream & cout, const Fraction & r)
{
    cout << r.n << '/' << r.d;
    return cout;
}
\end{Verbatim}

You should compile (correcting any errors if necessary)
and run the program 
to make sure that it works before
continuing. Note that in \verb!Fraction.cpp! the \verb!operator<<! is defined.
Right now, the only thing you need to know is that in the code for
\verb!operator<<!, if you want to print \verb!r.n! (say), call
\begin{console}
cout << r.n
\end{console}
instead of
\begin{console}
std::cout << r.n
\end{console}
i.e.:
\begin{Verbatim}[frame=single, commandchars=~!@]
// Author: 
// Date  : 
// File  : Fraction.cpp

#include <iostream>
#include "Fraction.h"


std::ostream & operator<<(std::ostream & cout, const Fraction & r)
{
    ~textbf!~underline!cout@@ << r.n << '/' << r.d;
    return cout;
}
\end{Verbatim}

This is how you should print inside this \lq\lq function" (technically speaking
this is an operator, not a function). Also, do not modify the prototype of this
\lq\lq function" nor remove the last statement in the body:
\begin{Verbatim}[frame=single, commandchars=~!@]
std::ostream & operator<<(std::ostream & cout, const Fraction & r)
{
    ...
    ~textbf!~underline!return cout;@@
}
\end{Verbatim}

In more detail, in \verb!main()!, suppose you have \verb!Time! variable called
\verb!r!, if you execute
\begin{console}
std::cout << r;
\end{console}
C\texttt{++} actually executes
\begin{console}
operator<< (std::cout, r);
\end{console}
i.e., it calls the \lq\lq function" (or rather the operator) \verb!operator<<!
so that in this \lq\lq function" the parameter \verb!cout! references
\verb!std::cout! and the reference variable \verb!r! references the \verb!r! in
\verb!main()!.

\newpage
\textsc{Compiling and Automating Test Inputs}

The following instructions are for those using Fedora.

Be neat! Have a directory just for this program and do not have your source
files cluttered with other things! To compile all your source files you do
\begin{console}
g++ *.cpp -o prog
\end{console}

This will take all your cpp files in the current directory and compile an
executable named \verb!prog!. (If you don't like \verb!prog! you can
choose any other name.)

You can automate the testing by entering test data into a file, say you called
it \verb!stdin.txt! (you can use any name you like):
\begin{console}
0 1 2
0 -1 2
0 4 7
0 15 10
\end{console}

To send the input from \verb!stdin.txt! to your program (instead of wasting
time typing the test data again and again), do this:
\begin{console}
./prog < stdin.txt
\end{console}

You can also send the output not to your terminal window but to a file like
this:
\begin{console}
./prog < stdin.txt > stdout.txt
\end{console}

You can now open \verb!stdout.txt! to see the output.

Yet another level of software testing automation is to figure out the correct
output and type the correct output to another file, say \verb!correct.txt!. You
can then get Fedora to test if \verb!stdout.txt! is the same as
\verb!correct.txt! using this command:
\begin{console}
diff correct.txt stdout.txt
\end{console}
If both files are the same, nothing is printed to your terminal window.

Hence your code--test cycle involves the following linux commands:
\begin{console}
g++ *.cpp -o prog
./prog < stdin.txt > stdout.txt
diff correct.txt stdout.txt
\end{console}

(Of course you have to continually add test inputs to \verb!stdin.txt! and the
correct output to \verb!correct.txt!.)




\newpage
\textsc{Operators}

Operators are easy to understand. They are just functions except that you call
them in a different way. Let me give you an example. First run the following
example:
\begin{Verbatim}[frame=single]
#include <iostream>


struct Blah
{
    int x;
    int y;
};


double funnyOperator(const Blah & i, const Blah & j)
{
    double d;
    d = (double)i.x / i.y + (double)j.x / j.y;
    return d;
}

  
int main()
{
    Blah b = {2, 3};
    Blah c = {5, 9};
    double d = funnyOperator(b, c);
    std::cout << d << std::endl;
    return 0;
}
\end{Verbatim}

(The actual body of the function is not important. Focus on how the function
is called.) There are no surprises here. Now try this:
\begin{Verbatim}[frame=single, commandchars=~!@]
#include <iostream>


struct Blah
{
    int x;
    int y;
};


double ~textbf!~underline!operator+@@(const Blah & i, const Blah & j)
{
    double d;
    d = (double) i.x / i.y + (double) j.x / j.y;
    return d;
}


int main()
{
    Blah b = {2, 3};
    Blah c = {5, 9};
    double d = ~textbf!~underline!operator+@@(b, c);
    std::cout << d << std::endl;
    return 0;
}
\end{Verbatim}
The name of the function, \verb!funnyOperator! is changed to \verb!operator+!.

And finally run this program:
\begin{Verbatim}[frame=single, commandchars=~!@]
#include <iostream>


struct Blah
{
    int x;
    int y;
};


double operator+(const Blah & i, const Blah & j)
{
    double d;
    d = (double) i.x / i.y + (double) j.x / j.y;
    return d;
}

  
int main()
{
    Blah b = {2, 3};
    Blah c = {5, 9};
    double d = ~textbf!~underline!b + c@@;
    std::cout << d << std::endl;
    return 0;
}
\end{Verbatim}

As you can see, in the above program:
\begin{console}
b + c
\end{console}
is really a call to the \lq\lq function" \verb!operator+!:
\begin{console}
 operator+(b, c)
\end{console}
except that the \lq\lq function" \verb!operator+!, in this context, is really
called an \textbf{\large operator}. Once again, the following are actually the
same:
\begin{Verbatim}
                       b + c            operator+(b, c)
\end{Verbatim}

You can define all kinds of operators in C\texttt{++}, including
\begin{console}
 >>      <<      +      -      *      /
\end{console}
etc.

A simple \verb!operator<<! is already defined for you which is why you can
print a \verb!Fraction! variable \verb!r! with
\begin{console}
std::cout << r;
\end{console}
