%-*-latex-*-
\sectionthree{Applications}
\begin{python0}
from solutions import *; clear()
\end{python0}

There are many applications of linked lists (singly or doubly linked), stacks, and
queues.
(Don't forget that stacks and queues can be implemented using lists so when
you use stacks/queues
you're might be using linked lists.)

I'm just going to talk about just a couple of them.
(Don't be fooled by this short list -- linked lists are really very
important.)

Linked lists are definitely used in language computation.
We have to start somewhere so I'm going to talk about
the processing of mathematical languages.
Natural language prcoessing (NLP) as in processing human language
(example: English) is a lot more complex because the English
language is not as precise as the mathematical language.
Pure abstract mathematical languages do not have humor, puns,
sarcasm, etc.

We'll look at the simplest sentences in mathematical languages.
The first few will be on evaluating mathematical expressions,
something that you
have been doing since elementary school.
Besides the standard mathematical expressions such as
\[
\text{\texttt{1 + 2 - 3 * 5}} 
\]
we will also look at expressions using a different mathemaical language where a
sentence in this language looks like this:
\[
\texttt{+ 3 5}
\]
The operator is placed \textit{in front}.
Then there's this mathematical language where the operator is placed at the back,
like this:
\[
\texttt{3 5 +}
\]
You'll see that the stack and queue can be used as memory devices while you
evaluate expressions.

I'll also talk about using linked list to represent
common quantities used in computer science and math (example:
long integers, matrices).

I'll briefly talk about using linked list to model sparse data structures
such as sparse long integers, sparse matrices, and
graphs (which are just dots joined by lines) especially when
each dot is not joined to too many other dots -- again something
that's \lq\lq sparse''.

We'll briefly look at some search problems.
This is a huge topic. We'll come back to search again later.

But let's start with something that looks naively simple ...
