%-*-latex-*-
\sectionthree{Delete and extract-root}
\begin{python0}
  from solutions import *; clear()
\end{python0}

Suppose I want to delete
\texttt{8}
from this maxheap:

\input{stdout35.tex}

In other words I want to delete the value at index 2
in the array implementation of the above heap.

I'll do it this way:
I look for the rightmost node of the last level --
this corresponds to the last value in the array implementation.
In the above case, that's the value \texttt{5}.
I then overwrite \texttt{8} with \texttt{5}:

\input{stdout36.tex}

to get this:

\input{stdout37.tex}

Now this is not a maxheap. Do you see why?

I look at the children of \texttt{5}: \texttt{0} and \texttt{7}.
I swap \texttt{5} with the max of the children which is \texttt{7}

\input{stdout38.tex}

and get this:

\input{stdout39.tex}

Clearly, we keep pushing \texttt{5} down as much as possible until
we get a maxheap again (i.e., heapify-down).
It might take more than one swap.

This above method works as long as \texttt{5} is a descendent of the
value to be removed.

Usually the value to be delete is in fact the root of the heap.
When the root is to be deleted, then the operation is called
\defone{extract-max} for the case of maxheap
and \defone{extract-min} for the case of
minheap.
I'm going to call both of them \defone{extract-root}.

Let's do another example.
Let's delete \texttt{20} (the root) from this maxheap:

\input{stdout40.tex}


Again, I overwrite \texttt{20} with \texttt{5}

\input{stdout41.tex}

(\texttt{5} is chosen, again because it's the rightmost
in the last level, or equivalently, it's the last in the
corresponding array implementation.)
I get this:

\input{stdout42.tex}

I do the same again as above: I pick the larger of the children
of \texttt{5}, which in this case is \texttt{10} and swap with \texttt{5}.
I get this:

\input{stdout43.tex}

It's not a maxheap yet.
I look at the children of \texttt{5} and choose the largest,
which would be \texttt{9}, and swap it with \texttt{5}.
Here's what I get:

\input{stdout44.tex}

Ahhh ... at this point I have a maxheap.
I'm done!

\begin{console}[commandchars=\\\{\}]
ALGORITHM: heap_delete (for maxheap)
INPUT: x - an array representing a maxheap
       n - the length of the heap in x (pass by reference)
       i - index of value to be removed (note usually this is 0)

x[i] = x[n - 1]
n = n - 1
heapify_down(x, i)
\end{console}

(Note that because of the shape of the tree -- a complete tree --
a node cannot have a right child but no left child.)

Recall what I just said: usually the delete operation for maxheap
occurs at index 0, i.e., you're removing the maximum value in the maxheap.
In that case the operation is also called extract-max or delete-max.
You'll see why when we use this delete operation to perform heapsort
and when we use this for priority queues.

Note that the runtime is
\[
  O(\log n)
\]
since the heapify-down operation basically moves
a value in the tree down to the leaf level, possibly stopping
early.
This means that the worse runtime must be
the height of the tree which is $O(\log n)$.

\input{exercises/delete-0/main.tex}

\input{exercises/delete-2/main.tex}

\input{exercises/delete-3/main.tex}

\input{exercises/delete-4/main.tex}
% choose the right because the tree might be one step shorter on the right
% and therefore the heapify down might terminate earlier by 1 step
%
% Example:
%     5
%    1 1
%   3

\input{exercises/delete-5/main.tex}
%NO
%    1
%  2   3
%
%    2
%  3
%
%    2      --->      1
%  3   1            3   2
%
