\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Dynamic \& Static Binding}

In the earlier sections, we saw that there are two ways to choose the \verb!f()! to execute, either \verb!P::f()! or \verb!C::f()!.

\EMPHASIZE{Dynamic binding} refers to the choice of method (between the parent method and the child method with the same name) where the method is virtual. This is also called \EMPHASIZE{late binding.}

\EMPHASIZE{Static binding} refers to the choice of method (between the parent method and the child method with the same name) where the method is not virtual. This is also known as \EMPHASIZE{early binding.}

\tab[5em]{\verb!C c;!}\\
\tab[5em]{\verb!P * p = c;!}\\

Why is it called dynamic (or late) or static (or early)?

Make sure you run the following really important example \ldots

Suppose \verb!C0! and \verb!C1! are both subclass of \verb!P!. There are virtual functions \verb!f! defined in \verb!P!, \verb!C0!, \verb!C1!.

\begin{console}
#include < iostream>

class P
{
public:
     virtual void f() { std::cout << "P::f\ n"; }
};

class C: public P
{
public:
     virtual void f() { std::cout << "C::f\ n"; }
};

class C1: public P
{
public:
     virtual void f() { std::cout << "C1::f\ n"; }
};

int main()
{   
    P * p;
    C c;
    C1 c1;
    int i;
    std::cout << "Do you want C or C1? (0 or 1) ";
    std::cin >> i;
    if (i == 0)
    {    
         p = &c;
    }
    else
    {
         p = &c1;
    }
    p-> f();
    
    return 0;
}
\end{console}

In the above, the decision of which \verb!f()! (\verb!C::f()! or \verb!C1::f()!) to use is \EMPHASIZE{not known during compilation} but is known \EMPHASIZE{only during runtime},. Therefore the binding is \EMPHASIZE{late} and is \EMPHASIZE{dynamic} (depending on runtime situation).

On the other hand, if \verb!f()! is not virtual -- remove the virtual keyword -- and no matter what you enter, you always execute \verb!P::f()!. Therefore which \verb!f()! is used is known during compilation (it's \verb!P::f()!) -- it' s known even before you execute the program since it' s already known during compile time. Therefore the decision is \EMPHASIZE{early}. After the program is compiled, you \EMPHASIZE{cannot} change the \verb!f()! -- the choice of \verb!f()! is \EMPHASIZE{static} and \EMPHASIZE{not dynamic.}
\newpage
\tab[6em]{\large\verb!P * p;!}\\
\tab[6em]{\large\verb!C c;!}\\
\tab[6em]{\large\verb!p = &c;!}
\begin{python}
from latextool_basic import *
from latexcircuit import *
p = Plot()

#p += Grid(x0=-5, y0=-5, x1=5, y1=5)

p += Rect(1, 2, 9, 5, s=r"If \texttt{f()} is not virtual, the decision of which \texttt{f()} to use is based on the pointer's type P.\\ \textbf{Static/early binding.}", linewidth=0.1, linecolor='red', innersep=0.3)

p += Rect(1, -2, 9, 1, s=r"If \texttt{f()} not virtual, the decision of which \texttt{f()} to use is based on the object's class C.\\ \textbf{Dynamic/late binding.}", linewidth=0.1, linecolor='red', innersep=0.3)
p += Rect(-3.5,2.5,-2.5,4.5,linewidth=0.05)
p += Line(points=[(-2.5,3.5),(1,3.5)], linestyle='dashed',linecolor='red',linewidth=0.1)
p += Line(points=[(-2.5,-0.5),(1,-0.5)], linestyle='dashed',linecolor='red',linewidth=0.1)
p += Line(points=[(-3,3.5),(-3,0.5)], endstyle='>',linewidth=0.05)
p += ellipse(center=(-3,-0.5), height=2,width=3,linewidth=0.05)
X = POINT(x=-3.9, y=3.5, r=0, label='pointer to object')
Y = POINT(x=-4.9, y=-0.5, r=0, label='object')
p += str(X)
p += str(Y)
print(p)
\end{python}
WARNING: Static here refers to static \EMPHASIZE{binding}, i.e., which \verb!f()! is the p-> f() bound to. Do not be confused with static members!!!

\textbf{Make sure you do the following exercises!}

\input{../exercises/1-81-polymorphism/question.tex}\input{../exercises/2-81-polymorphism/question.tex}\input{../exercises/3-81-polymorphism/question.tex}
