%-*-latex-*-
\section{Heapify-up}

The following two operations are the basic tools for min- and maxheaps.

Look at this:

\begin{python}
from latextool_basic import *
from drawheap import *  
p = Plot()
edges={'10':['5','8'],
       '5':['2', '1'],
       '8':['3', '7'],
       '2':['0', '9'],
       }
drawheap(p, edges)
print(p)
\end{python}

It's almost a maxheap except for the \verb!9!.

What is the simplest way to make it a maxheap?
I compare \verb!9! with it's parent, i.e., \verb!2!.
Since this is a maxheap, something is wrong:
parents should be larger than children.
So I swap \verb!9! and \verb!2! to get:

\begin{python}
from latextool_basic import *
from drawheap import *  
p = Plot()
edges={'10':['5','8'],
       '5':['9', '1'],
       '8':['3', '7'],
       '9':['0', '2'],
       }
drawheap(p, edges)
print(p)
\end{python}

I then compare \verb!9! and \verb!5! and decide to swap again to get:

\begin{python}
from latextool_basic import *
from drawheap import *  
p = Plot()
edges={'10':['9','8'],
       '9':['5', '1'],
       '8':['3', '7'],
       '5':['0', '2'],
       }
drawheap(p, edges)
print(p)
\end{python}

Now it's a maxheap.
Basically, you continually swap the value in question with
its parent if the value is larger than the parent.

This is called
\defone{heapify-up}.
It's also called
\defone{bubble up} or
\defone{percolate up}.


In general, if \verb!x[0..n]! is an array
and I can talk about heapify-up on array \verb!x!
treating \verb!x! as a maxheap,
starting at index \verb!i! and going up
along its path to the root (i.e., index 0).

You can also talk about heapify-up on an array
treating it as a minheap,
starting at some index.

For a maxheap, to heapify-up at an index \verb!i!, you compare
\verb!x[i]! and \verb!x[parent(i)]!
and swap them if necessary, i.e. if \verb!x[i]! $>$ \verb!x[parent(i)]!
and you keep following that value up to the root.
If \verb!x[i]! $\leq$ \verb!x[parent(i)]!, you stop.
That's it.

\begin{Verbatim}[frame=single]
ALGORITHM: heapify-up (for maxheap)
INPUTS: x[0..n-1] -- array
        i         -- index of x

while 1:
    if parent(i) < 0: break
    if x[i] > x[parent(i)]:
        swap x[i], x[parent(i)]
        i = parent(i)
\end{Verbatim}

With a tiny of optimization:
\begin{Verbatim}[frame=single]
ALGORITHM: heapify-up (for maxheap)
INPUTS: x[0..n-1] -- array
        i         -- index of x

v = x[i]
while 1:
    p = parent(i)
    if p < 0: break
    if v <= x[p]:
        x[i] = v
        break
    else:
        x[i] = x[p]
        i = p
\end{Verbatim}

The same idea works for minheap.
