% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\author{}
\date{}

\begin{document}

Class hour planning

5 minQuiz on previous class meeting

25 minLecture

15 minQuick lab on current class

5 minQuiz on current class meeting

Grading

Day -7: Work study students create accounts on gandalf for students. Use
CC email addresses.

\textbf{Track 1:} Students with little or no programming experience

Track 2: Students with extensive programming experience

Turning in homework:

\begin{itemize}
\tightlist
\item
  Hardcopy with honor statement
\item
  Upload to bilbo?
\end{itemize}

Day 1

\begin{itemize}
\tightlist
\item
  Distribute course information
\item
  General advice:
\item
  talk about my teaching style for 240
\item
  emphasize that 240 is made up of (a) learning a language and (b)
  problem solving. (a) requires repetition.
\item
  learn by doing
\item
  read, study notes. Type and run all programs several times.
\item
  slow typers must come up to speed this week. Best way is to type the
  textbook especially every problem.
\item
  Frequent quizzes, labs start maybe week 3, weekly assignments.
\item
  Distribute MS Studio .NET CDs and installation instructions. Students
  must install by end of day. Talk to instructor/workstudy students if
  help is needed. Return CDs next class hour.
\item
  Distribute linux lab hours {[}NOT DONE{]}
\item
  Demo of using MS Studio .NET. Students must be able to recreate
  helloworld in 2 minutes by tomorrow's lab hour -- quiz. Closed book.
\item
  Readings: Gaddis chap 1. Finish reading before next class.
\end{itemize}

Ws meeting:

\begin{itemize}
\tightlist
\item
  lab hours
\item
  unix accounts. Student name, login, password csv file emailed by
  tomorrow 8AM.
\item
  question for notes
\item
  plone migration update
\end{itemize}

Day 2

\begin{itemize}
\tightlist
\item
  Quiz 1: hello world (2 min)
\end{itemize}

\begin{itemize}
\tightlist
\item
  Distribute unix accounts (Day 2?)
\item
  Demo on using WinSCP (Day 2?). Organization of their home directory.
\item
  Assignment: winscp submit hello world (Day 2?)
\item
  Readings: Gaddis 2.1-2.6
\end{itemize}

\end{document}
