\tinysidebar{\debug{exercises/{test/question.tex}}}
\begin{ex}
Write a while-loop that prompts the user for a row \texttt{r}, column \texttt{c} and value \texttt{v} and set \texttt{x[r][c]} to \texttt{v}. Print the whole array after each change to the array. If the user enters a negative value for \texttt{r} or \texttt{c}, exit the while-loop

\begin{consolethree}[escapeinside=||]
const int ROW_SIZE = 3;

const int COL_SIZE = 2;

int x[ROW_SIZE][COL_SIZE];

for (int row = 0; row < ROW_SIZE; row++)
{
     for (int col = 0; col < COL_SIZE; col++)
     {
          x[row][col] = row + col;
     }
}

for (int row = 0; row < ROW_SIZE; row++)
{
     for (int col = 0; col < COL_SIZE; col++)
     {
       std::cout << x[row][col] << ' ';
     }
     std::cout << std::endl;
}

while (1)
{

// Prompt for r. If r < 0, break the loop.

// Prompt for c. If c < 0, break the loop.

// Prompt for v and set x[r][c] to v

// Print the array

}
\end{consolethree}

Test your program.

\begin{ex}
Continuing with the above example, initialize the 3-by-3 tic-tac-toe board to

\begin{align*}
' ' & 'O' & ' ' \\
' ' & 'O' & ' ' \\
'X' & 'X' & 'X' \\
\end{align*}

and write code to check that 'X' has a winning row. The code should detect a winning row for row 2. Now add code to detect either a winning row or winning column. The code should detect a winning column for this

\begin{align*}
' ' & 'O' & 'X' \\
' ' & 'O' & 'X' \\
' ' & ' ' & 'X' \\
\end{align*}

\begin{ex}
Declare an array of 5-by-10 integers. Assign random integers from 1 to 10 to the array. Print the array in a 5-by-10 grid of values separating values in each row by a space and separating each row with a newline. Compute and print the average of the values in the whole array.
\end{ex}
