\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Recursive computation on arrays: Part 1}

Here's another example. You already know that you can
write a loop to compute the max value in an array. Let's
look at this problem recursively. Suppose I write x is an array of
value 2,3,5,6,4,1 and we want a function that looks like this:

\verb!max(x)!

If you split the array x into the first value of the array

\verb!2!

and the rest say we call it y of value 3,5,6,4,1, then we have this
relation:

\verb!max(x) = maximum of 2 and max(y)!

You see that the max function occurs on both sides. This is a recursive
relation. Furthermore, we need the function max to accept index values.
Then

\begin{consolethree}[escapeinside=||]
// Returns max of x[start], x[1], x[2], ...,
// x[end - 1]
int max(int x[], int start, int end)
{
    if (start == end - 1)
    {
        return x[start];
    }
    else
    {
        if (x[start] < max(x, start + 1, end))
        {
            return max(x, start + 1, end);
        }
        else
        {
            return x[start];
        }
    }
}
\end{consolethree}

Note that \texttt{max(x, start + 1, end)} is computed twice. To save
computations, we can do this:

\begin{consolethree}[escapeinside=||]
int max(int x[], int start, int end)
{
    if (start == end - 1)
    {
        return x[start];
    }
    else
    {
        int y = max(x, start + 1, end);
        if (x[start] < y)
        {
            return y;
        }
        else
        {
            return x[start];
        }
    }
}
\end{consolethree}

Essentially you break the problem of finding the max of
x[0],x[1],...,x[n-1] to

\textbf{maximum between x[0] and max of x[1],x[2],...,x[n-1]}

<<<<<<< HEAD
\input{../exercises/33-29-recursion/question.tex}
\input{../exercises/34-29-recursion/question.tex}
\input{../exercises/35-29-recursion/question.tex}
\input{../exercises/36-29-recursion/question.tex}
\input{../exercises/37-29-recursion/question.tex}
=======
\begin{ex}
Write a recursive version of a minimum function

\verb!int min(int x[], int start, int end);!

that returns the minimum of x[start], \ldots, x[end - 1].
\end{ex}

\begin{ex}
Write a recursive version of a linear search function

\verb!int linearsearch(int x[], int start, int end, int target);!

that finds the first index where target is found in the array
x[start], x[start + 1], \ldots, x[end - 1].
\end{ex}

\begin{ex}
Write a recursive version of a count function

\verb!int count(int x[], int start, int end, int target);!

that returns the number of times target occurs in x[start], x[start + 1], x[start + 2], \ldots, x[end - 1].
\end{ex}

\begin{ex}
Write a recursive version of a sum function

\verb!int sum(int x[], int start, int end);!

that returns x[start] + x[start + 1] + x[start + 2] +
\ldots{} + x[end - 1].
\end{ex}

\begin{ex}
A string is a palindrome if it's the same when you read it left-to-right and right-to-left. For instance here's one:

\verb!madam!

Here's another one: What does Adam say to Eve?

\verb!Madam, I'm Adam.!

If you remove spaces and punctuations and change everything to lowercase, you get

\verb!madamimadam!

Write a recursive function with the prototypes

\verb!bool is_palindrome(char s[], start, end);!

so that \verb!is_palindrome("madamimadam", 0, 11)! returns true.
\end{ex}

>>>>>>> parent of 3043c95 (.)

