\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Sum from 1 to n}

Suppose we write

sum(n)

to denote ``the sum of all integers from 0 to n'', i.e.,

sum(n) = 0 + 1 + 2 + 3 + \ldots{} + n

Of course we know how to sum from 1 to n without using recursion but
using a loop:

\begin{consolethree}[escapeinside=||]
int sum(int n)
{
    int s = 0;
    for (int i = 0; i < n; ++i)
    {
        s += i;
    }
    return s;
}
\end{consolethree}

In fact the \ldots{} in sum(n) = 1 + 2 + 3 + \ldots{} + n more or less described a loop.

Let me think of the sum(n) computation recursively \ldots{} let's try
some examples to get a feel for the problem. Here's an
example:

sum(5) = 0 + 1 + 2 + 3 + 4 + 5

Note that

sum(4) = 0 + 1 + 2 + 3 + 4

Therefore you have this:

sum(5) = 0 + 1 + 2 + 3 + 4 + 5

= (0 + 1 + 2 + 3 + 4) + 5

= sum(4) + 5

Clearly we also have

sum(8) = 0 + 1 + 2 + 3 + 4 + 5 + 6 + 7 + 8

= (0 + 1 + 2 + 3 + 4 + 5 + 6 + 7) + 8

= sum(7) + 8

See you should quickly see that in general if n is a positive integer
greater than 0, then

sum(n) = sum(n - 1) + n

You see that sum(n) is related to sum(n - 1). Also note that

sum(0) = 0

So if we define sum(n) to be ``the sum of all integers from 0 to n'',
then we get

\[
\text{sum}(n) = \begin{cases}
0 & \text{if } n = 0 \\
\text{sum}(n - 1) + n & \text{if } n > 0
\end{cases}
\]

The relationship (when n > 0)

sum(n) = sum(n - 1) + n

is called a \EMPHASIZE{recursion} because the sum function called
\EMPHASIZE{itself}. The case when n = 0

sum(0) = 0

is called the \EMPHASIZE{base case} of the recursion.

Recursion is an \EMPHASIZE{extremely} important phenomenon in math,
computer science, \ldots, in nature. So listen up.

First of all, if I ask you to give me sum(5), you would say, well it's
just

sum(5) = 1 + 2 + 3 + 4 + 5

and then you evaluate the express to get 15. But you can also use

\[
\text{sum}(n) = \begin{cases}
0 & \text{if } n = 0 \text{ (B)} \\
\text{sum}(n - 1) + n & \text{if } n > 0 \text{ (R)}
\end{cases}
\]

to get the same value. (Note that I've labeled the two cases.) Here's
how you use (B) and (R) to compute sum(5):

Since 5 > 0, from (R) we get

sum(5) = sum(5 - 1) + 5

So??? Well if we simplify that we get

sum(5) = sum(4) + 5

That doesn't give us a value of sum(5). But wait, sum(4)
can be written in a different way: Because 4 > 0, by (R):

sum(4) = sum(4 - 1) + 4

which simplifies to

sum(4) = sum(3) + 4

So altogether we have this:

sum(5) = sum(4) + 5

sum(4) = sum(3) + 4

That still does NOT tell us what the value of sum(5) is!!! But hang on,
with the same reasoning (using (R)) we get three more facts:

sum(3) = sum(2) + 3

sum(2) = sum(1) + 2

sum(1) = sum(0) + 1

So altogether we have

sum(5) = sum(4) + 5

sum(4) = sum(3) + 4

sum(3) = sum(2) + 3

sum(2) = sum(1) + 2

sum(1) = sum(0) + 1

Now you might be tempted to write

sum(0) = sum(-1) + 0

That's wrong!!! Because the relationship (R)

(R): sum(n) = sum(n - 1) + n

applies only when n > 0. You cannot apply it to the case of
n = 0. But wait \ldots{} for the case of n = 0 we have this:

(B): sum(0) = 0

Therefore altogether we have

sum(5) = sum(4) + 5

sum(4) = sum(3) + 4

sum(3) = sum(2) + 3

sum(2) = sum(1) + 2

sum(1) = sum(0) + 1

sum(0) = 0

So what? Well, with these we can compute sum(5)!!! Let me show you how.
First let me label the facts:

(5): sum(5) = sum(4) + 5

(4): sum(4) = sum(3) + 4

(3): sum(3) = sum(2) + 3

(2): sum(2) = sum(1) + 2

(1): sum(1) = sum(0) + 1

(0): sum(0) = 0

From equation (0), we know that sum(0) = 0. Substituting this into
equation (1) we get

sum(1) = sum(0) + 1 = 0 + 1

And if we substitute this into equation (2) we get

sum(2) = sum(1) + 2 = 0 + 1+ 2

Substituting this into equation (3) we get

sum(3) = sum(2) + 3 = 0 + 1+ 2 + 3

And when this is put into equation (4) we get

sum(4) = sum(3) + 4 = 0 + 1+ 2 + 3 + 4

And of course finally we get

sum(5) = sum(4) + 5 = 0 + 1+ 2 + 3 + 4 + 5

So whether you define sum(n) as

sum(n) = 1 + 2 + \ldots{} + n

(which looks more like a loop) or

\[
\text{sum}(n) = \begin{cases}
0 & \text{if } n = 0 \\
\text{sum}(n - 1) + n & \text{if } n > 0
\end{cases}
\]

you get the same value for sum(n).

\input{temp/29-recursion/exercises/0-29-recursion/question.tex}
The two ways of looking at sum(n), using a loop and using recursion,
gives us the same result. The really important question to ask is this:

Why should

\[
\text{sum}(n) = \begin{cases}
0 & \text{if } n = 0 \\
\text{sum}(n - 1) + n & \text{if } n > 0
\end{cases}
\]

be a good alternative view of

sum(n) = 0 + 1 + 2 + \ldots{} + n?

The reason is because of this \ldots{} pay attention! \ldots{} many
problems in nature present themselves naturally in a \EMPHASIZE{recursive
way}. For more complex problems, the recursive way is the way that leads
to a simpler solution. And simple mean less bugs.

(For those of you who have taken Physics and Differential Equations, you
know that many problems in Physics requires the description of a
function or behavior of a physical phenomenon, but the function takes
part in a relationship of a special form -- the function appears in a
differential equation. For CS, many problems gives rise to functions
which are recursive.)

Of course we still need code \ldots{} but don't worry: We can easily
translate the recursive sum(n)

\[
\text{sum}(n) = \begin{cases}
0 & \text{if } n = 0 \\
\text{sum}(n - 1) + n & \text{if } n > 0
\end{cases}
\]

into a C++ function:

\begin{consolethree}[escapeinside=||]
int sum(int n)
{
    if (n == 0)
    {
        return 0;
    }
    else
    {
        return sum(n - 1) + n;
    }
}
\end{consolethree}

Note that the \EMPHASIZE{recursive version does not have loops}. Make sure
you study and test it carefully.

\input{temp/29-recursion/exercises/1-29-recursion/question.tex}
In this case, the sum-from-1-to-n is easy so recursion is not really
necessary. The point of the section is to show you that
it's possible to write a recursion function for
sum-from-1-to-n. In the real world many problems are described more
naturally in a recursive form and solving such problems using loops is
extremely untidy and painful.

\input{temp/29-recursion/exercises/2-29-recursion/question.tex}
\input{temp/29-recursion/exercises/3-29-recursion/question.tex}
\input{temp/29-recursion/exercises/4-29-recursion/question.tex}
\input{temp/29-recursion/exercises/5-29-recursion/question.tex}
\input{temp/29-recursion/exercises/6-29-recursion/question.tex}
\input{temp/29-recursion/exercises/7-29-recursion/question.tex}

