\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{How to discover recursion: look for smaller subproblems inside a problem}

This is really important \ldots{} !!!

Let's step back and look at the recursive version of the
computation of ``sum from 0 to n'' function and the ``factorial of n''.

Here's the sum(n) (i.e., sum from 0 to n):

$sum(0) = 0$

$sum(n) = n + sum(n - 1)$ if $n > 0$

and here's the factorial (which I'll
just call f):

$f(0) = 1$

$f(n) = n * f(n - 1)$ if $n > 0$

How does one recognize the recursive form of a function? The crucial
thing is to see a smaller subproblem within a bigger problem. You should
also look at specific concrete examples. This is what I mean \ldots{}

Let's look at the sum(n) and say you
don't know how to write down the recursive form yet. You
try a concrete example, say the problem of computing sum(5). It looks
like this:

$sum(5) = 0 + 1 + 2 + 3 + 4 + 5$

Next, you ask yourself if you can see another sum(n) within sum(5). You
see immediately that sum(4) appears within sum(5):

$sum(5) = 0 + 1 + 2 + 3 + 4 + 5$

$= (0 + 1 + 2 + 3 + 4) + 5$

$sum(4) = 0 + 1 + 2 + 3 + 4$

i.e.,

$sum(5) = sum(4) + 5$

You then try a few more examples and you should see that in general

$sum(n) = sum(n - 1) + n$

Of course this is true for n > 0. You are pretty much done
except for sum(0). You then ask yourself what is the meaning of sum(0).
You see that it must be 0. The base case is not as abstract since
it's just one case and is usually the easier to figure
out.

It's really the same for the factorial too. You look at
a concrete example like f(5):

$f(5) = 5 * 4 * 3 * 2 * 1$

and you try to see another factorial on the right. You see that it must
be:

$f(5) = 5 * (4 * 3 * 2 * 1) = 5 * f(4)$

in other words

$f(n) = n * f(n - 1)$ \hspace{1cm} $(+)$

If you don't see what f(0) is, you just try some
examples using

$f(n) = n * f(n - 1)$ \hspace{1cm} $(+)$

Let's try f(5).

$f(5) = 5 * f(4)$ using (+) with n = 5

$= 5 * (4 * f(3))$ using (+) with n = 4

$= 5 * (4 * (3 * f(2)))$ using (+) with n = 3

$= 5 * (4 * (3 * (2 * f(1))))$ using (+) with n = 2

$= 5 * (4 * (3 * (2 * (1 * f(0)))))$ using (+) with n = 1

But you know concretely that

$f(5) = 1 * 2 * 3 * 4 * 5$

which means that

$1 * 2 * 3 * 4 * 5 = 5 * (4 * (3 * (2 * (1 * f(0)))))$

i.e., (removing the annoying parentheses):

$1 * 2 * 3 * 4 * 5 = 5 * 4 * 3 * 2 * 1 * f(0)$

which means, after canceling terms,

$1 = f(0)$

Remember this extremely important principle: To design a recursive
formula, you must see a \EMPHASIZE{smaller subproblem(s) within a bigger
problem}.

For instance you must see sum(n - 1) within sum(n) and you must see f(n - 1) inside f(n).

There are times when instead of ``n - 1'' in the smaller problem you
might see the smaller subproblem of g(n - 2) within g(n). Or you might
see h(n - 1) and h(n - 2) with h(n). Etc.


