\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Recursive function of two parameters}

The \EMPHASIZE{number of ways to choose r objects from n} is a very common
problem in math and computer science. It occurs so frequently that a
special symbol has been invented for it:

$\binom{n}{r}$

(We read this as ``n choose r''). Let's do a simple
example. Suppose we have 5 symbols A, B, C, D, E and we want to choose
two symbols from this set of symbols. These are all the possible
selections:

\textbf{AB, AC, AD, AE, BC, BD, BE, CD, CE, DE}

In other words ``5 choose 2'' is 10. Note that AB is the same as BA
since BA is ``choosing B and A''. We are not interested in the order of
choosing the symbols; choosing A then B is the same as choosing B then
A.

In fact there's a formula for ``n choose r'' in terms of
factorials. It's given by this:

\[
\binom{n}{r} = \frac{n!}{r!(n-r)!}
\]

So for ``5 choose 2'' is

\[
\frac{5!}{2!(5-2)!} = \frac{5!}{2!3!} = \frac{5 \cdot 4 \cdot 3 \cdot 2 \cdot 1}{(2 \cdot 1)(3 \cdot 2 \cdot 1)} = 10
\]

and you can see easily that that is indeed 10. (I'm not
going to explain why the above formula works. Take MATH225 if you want
to find out more.)

But there's another way to compute ``n choose r'' that
avoids the factorial ...

Let's write C(n,r) for ``n choose r''.
Here's a relation:

$C(n, r) = C(n - 1, r) + C(n - 1, r - 1)$

If you use this repeatedly to compute C(5, 2) (``5 choose 2'') you get
this:

$C(5, 2) = C(4, 2) + C(4, 1)$

$= (C(3, 2) + C(3, 1)) + (C(3, 1) + C(3, 0))$

This formula is obviously helpful:

$C(n, 0) = 1$

Continuing the computation:

$C(5, 2) = C(4, 2) + C(4, 1)$

$= (C(3, 2) + C(3, 1)) + (C(3, 1) + 1)$

$= ((C(2, 2) + C(2, 1)) + (C(2,1) + C(2,0))$

$+ ((C(2, 1) + C(2, 0)) + 1)$

At this point it's useful to know this:

$C(n, n) = 1$

Continuing the computation:

$C(5, 2) = ((1 + C(2, 1)) + (C(2,1) + 1)$

$+ ((C(2, 1) + 1) + 1)$

$= ((1 + (C(1, 1) + C(1, 0)))) + ((C(1, 1) + C(1, 0)) + 1)$

$+ (((C(1, 1) + C(1, 0)) + 1) + 1)$

$= 10$

In summary we have these relations:

$C(n, r) = C(n - 1, r) + C(n - 1, r - 1)$ if $r > 0$ and $r < n$

$C(n, 0) = 1$

$C(n, n) = 1$

Here's the code implementing the above mathematical
facts about C(n,r):

\begin{consolethree}[escapeinside=||]
int c(int n, int r)
{
    if (n == r || r == 0)
    {
        return 1;
    }
    else
    {
        return c(n - 1, r) + c(n - 1, r - 1);
    }
}
\end{consolethree}

Note that the recursion is different from previous examples because c()
calls itself \EMPHASIZE{twice:}

\verb!return c(n - 1, r) + c(n - 1, r - 1);!

For instance \texttt{c(5,2)} calls \texttt{c(4,2)} and \texttt{c(4,1)}, adds
the return values, and return this sum. The mental picture is therefore
slightly more complex. This is part of the diagram:

\begin{python}
from latextool_basic import *
p = Plot()

p += Rect(0, 0, 3, 1.5, label=r'\texttt{c(4,2)}', linewidth=0.05)
p += Rect(5.5, 0, 8.5, 1.5, label=r'\texttt{c(4,1)}', linewidth=0.05)
p += Rect(2.75, -3, 5.75, -1.5, label=r'\texttt{c(5,2)}', linewidth=0.05)

p += Line(points=[(2.5, 0), (3.75, -1.5)], endstyle='>', linewidth=0.075)
p += Line(points=[(3.25, -1.5), (2, 0)], endstyle='>', linewidth=0.075)

p += Line(points=[(6, 0), (5, -1.5)], endstyle='>', linewidth=0.075)
p += Line(points=[(5.5, -1.5), (6.5, 0)], endstyle='>', linewidth=0.075)

p += Line(points=[(4.25, -4.5), (4.25, -3)], endstyle='>', linewidth=0.075)
print(p)
\end{python}

The whole picture looks like this. I will only draw a line for each
function call and omit the line for return:

\begin{python}
from latextool_basic import *
p = Plot()

bw, bh = 1.5, 0.8

labels = [
    ['c(5, 2)'],
    ['c(4, 2)', 'c(4, 1)'],
    ['c(3, 2)', 'c(3, 1)', 'c(3, 1)', 'c(3, 0)'],
    ['c(2, 2)', 'c(2, 1)', 'c(2, 1)', 'c(2, 0)', 'c(2, 1)', 'c(2, 0)'],
    ['c(1, 1)', 'c(1, 0)', 'c(1, 1)', 'c(1, 0)', 'c(1, 1)', 'c(1, 0)']
]

xpos = [
    [0],
    [-4, 4],
    [-6, -2, 2, 6],
    [-7, -5, -3, -1, 1, 3],
    [-7, -5, -3.5, -1.5, 0.5, 2.5]
]

ypos = [0, 1.5, 3, 4.5, 6]

pos = []
for lvl in range(5):
    lvl_pos = []
    for i, (x, lbl) in enumerate(zip(xpos[lvl], labels[lvl])):
        y = ypos[lvl]
        p += Rect(x - bw/2, y, x + bw/2, y + bh, innersep=0.1, s=lbl, align='c', linewidth=0.05)
        lvl_pos.append((x, y))
    pos.append(lvl_pos)

# Level 0 -> 1
p += Line(points=[(pos[0][0][0], pos[0][0][1] + bh), (pos[1][0][0], pos[1][0][1])], linewidth=0.05)
p += Line(points=[(pos[0][0][0], pos[0][0][1] + bh), (pos[1][1][0], pos[1][1][1])], linewidth=0.05)

# Level 1 -> 2
p += Line(points=[(pos[1][0][0], pos[1][0][1] + bh), (pos[2][0][0], pos[2][0][1])], linewidth=0.05)
p += Line(points=[(pos[1][0][0], pos[1][0][1] + bh), (pos[2][1][0], pos[2][1][1])], linewidth=0.05)
p += Line(points=[(pos[1][1][0], pos[1][1][1] + bh), (pos[2][2][0], pos[2][2][1])], linewidth=0.05)
p += Line(points=[(pos[1][1][0], pos[1][1][1] + bh), (pos[2][3][0], pos[2][3][1])], linewidth=0.05)

# Level 2 -> 3 
p += Line(points=[(pos[2][0][0], pos[2][0][1] + bh), (pos[3][0][0], pos[3][0][1])], linewidth=0.05)
p += Line(points=[(pos[2][0][0], pos[2][0][1] + bh), (pos[3][1][0], pos[3][1][1])], linewidth=0.05)
p += Line(points=[(pos[2][1][0], pos[2][1][1] + bh), (pos[3][2][0], pos[3][2][1])], linewidth=0.05)
p += Line(points=[(pos[2][1][0], pos[2][1][1] + bh), (pos[3][3][0], pos[3][3][1])], linewidth=0.05)
p += Line(points=[(pos[2][2][0], pos[2][2][1] + bh), (pos[3][4][0], pos[3][4][1])], linewidth=0.05)
p += Line(points=[(pos[2][2][0], pos[2][2][1] + bh), (pos[3][5][0], pos[3][5][1])], linewidth=0.05)

# Level 3 -> 4 
p += Line(points=[(pos[3][1][0], pos[3][1][1] + bh), (pos[4][0][0], pos[4][0][1])], linewidth=0.05)
p += Line(points=[(pos[3][1][0], pos[3][1][1] + bh), (pos[4][1][0], pos[4][1][1])], linewidth=0.05)
p += Line(points=[(pos[3][2][0], pos[3][2][1] + bh), (pos[4][2][0], pos[4][2][1])], linewidth=0.05)
p += Line(points=[(pos[3][2][0], pos[3][2][1] + bh), (pos[4][3][0], pos[4][3][1])], linewidth=0.05)
p += Line(points=[(pos[3][4][0], pos[3][4][1] + bh), (pos[4][4][0], pos[4][4][1])], linewidth=0.05)
p += Line(points=[(pos[3][4][0], pos[3][4][1] + bh), (pos[4][5][0], pos[4][5][1])], linewidth=0.05)

print(p)
\end{python}

Isn't this beautiful?

Do you notice that this looks like a tree? This structure is in fact
called a \EMPHASIZE{tree}. The box labeled c(5,2) is called the
\EMPHASIZE{root} of the tree. The boxes are called \EMPHASIZE{nodes/vertices}
of the tree. The boxes that corresponds to base cases are called
\EMPHASIZE{leaf nodes/vertices}. You will learn more about this
mathematical structure in your other Math/CS classes (for instance
MATH325, CISS350, CISS358, etc.)

\input{temp/29-recursion/exercises/8-29-recursion/question.tex}
\input{temp/29-recursion/exercises/9-29-recursion/question.tex}
\input{temp/29-recursion/exercises/10-29-recursion/question.tex}

