\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Degree 2 recursion}

The \EMPHASIZE{fibonacci numbers} are defined to be

\textbf{1, 1, 2, 3, 5, 8, 13, 21, ...}

i.e., the first two fibonacci numbers are 1 and a subsequent fibonacci
number is defined to the the sum of the previous two. For instance 5 is
the sum of 2 and 3. Therefore if fib() is the fibonacci function, we
have this:

$fib(n) = fib(n-1) + fib(n-2)$ if $n > 1$

$fib(0) = 1$

$fib(1) = 1$

This is called a \EMPHASIZE{degree two} recursion because fib(n) is in
terms of fib(n-1) and fib(n-2) and the largest difference between the
parameters is 2 (i.e., the largest difference is between n and n - 2 which is 2).

<<<<<<< HEAD
\input{../exercises/14-29-recursion/question.tex}
\input{../exercises/15-29-recursion/question.tex}
=======
\begin{ex}
What is the degree of this recursive definition:

$a(n) = 3a(n-1) + 4a(n-2) + 5a(n-3)$ if $n > 2$

$a(0) = 0$

$a(1) = 1$

$a(2) = 2$

What will happen if I only give you two base cases such as

$a(n) = 3a(n-1) + 4a(n-2) + 5a(n-3)$ if $n > 1$

$a(0) = 0$

$a(1) = 1$
\end{ex}

\begin{ex}
What is the degree of this recursive definition:

$b(n) = 5b(n-2) - 7b(n-4) + 12b(n-7)$

How many base cases should there be?
\end{ex}

>>>>>>> parent of 3043c95 (.)
Let's work out fib(2) using the formula.

$fib(2) = fib(1) + fib(0) = 1 + 1 = 2$

What about f(4)?

$fib(4) = fib(3) + fib(2)$

$= (fib(2) + fib(1)) + (fib(1) + fib(0))$

$= ((fib(1) + fib(0)) + 1) + (1 + 1)$

$= ((1 + 1) + 1) + 2$

$= (2 + 1) + 2$

$= 3 + 2$

$= 5$

Note that previously we worked with degree one recursion. For instance
look at this:

$f(n) = n x f(n-1)$, if $n > 0$

f(0) = 1

The difference between n and n-1 is 1. The factorial is a recursive
function of degree 1.

Back to our recursion. Here's the code:

\begin{consolethree}[escapeinside=||]
int fib(int n)
{
    if (n == 0 || n == 1)
        return 1;
    else
        return fib(n - 1) + fib(n - 2);
}
\end{consolethree}

<<<<<<< HEAD
\input{../exercises/16-29-recursion/question.tex}
\input{../exercises/17-29-recursion/question.tex}
\input{../exercises/18-29-recursion/question.tex}
\input{../exercises/19-29-recursion/question.tex}
=======
\begin{ex}
Draw the function call graph for fib(4) -- include
the return call arrows as well. Remember to label the return arrows with
the return value. How many nodes/vertices are there? Can you figure out
how roughly many nodes/vertices there are for fib(n) in general?
\end{ex}

\begin{ex}
You are given the mathematical description of a
degree 2 recursion:

$a(n) = a(n-1) + 3 * a(n-2)$ if $n > 1$

$a(0) = 1$

$a(1) = 2$

Compute a(2), a(3) by hand. Next, write a degree 2 C++ recursive
function for a(). Test your code by printing a(0), a(1), a(2), and a(3).
Make sure your computation by hand matches the output of your program.
Draw the function call diagram for a(4).
\end{ex}

\begin{ex}
In some books the fibonacci function is defined like
this:

$fib(n) = fib(n-1) + fib(n-2)$ if $n > 1$

$fib(0) = 0$

$fib(1) = 1$

Note that fib(0) is 0 and not 1. Write down the value of fib(n) for n =
0, 1, 2, 3, 4, 5, 6, 7. Implement this fib() function in C++ and print
the values of fib(n) for n = 0, 1, 2, 3, 4, 5, 6, 7 and verify that your
computations were done correctly. What is the relationship between this
fibonacci function and ours earlier.
\end{ex}

\begin{ex}
What's wrong with this recursion?

\begin{consolethree}[escapeinside=||]
int f(int x)
{
    if (x == 0)
        return 3;
    else
        return x * f(x - 2);
}
\end{consolethree}

If you don't see it, try to compute f(5) by hand. Next,
insert some print statements into the function like this:

\begin{consolethree}[escapeinside=||]
int f(int x)
{
    std::cout << "f(" << x << ")" << std::endl;
    if (x == 0)
        return 3;
    else
        return x * f(x - 2);
}
\end{consolethree}

and try to call f(5).
\end{ex}

>>>>>>> parent of 3043c95 (.)

