% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\author{}
\date{}

\begin{document}

C++

v0.7 (2023)

Yihsiang Liow

I hope one day I have notes (openbooks) for all my classes so that
students won't have to spend 100s of dollars on books. This is one of
them. It's still under heavy construction.

Goal

Learn C++ programming, learn some CS stuff, and have fun.

Audience

Anyone who can read, has access to a computer, feels comfortable using a
computer, has some basic math background, and most importantly, wants to
learn.

Disclaimer

I hope the number of errors in this set of notes is close to zero.
Please report bugs to
\href{mailto:yliow@ccis.edu}{\nolinkurl{yliow@ccis.edu}}.

PS

For more info on C++, check out my CISS350, CISS362, and other
miscellaneous C++ notes.

Acknowledgments

I want to thank students from the following classes for taking for this
course.

\begin{itemize}
\tightlist
\item
  Fall 2003
\item
  Spring 2004
\item
  Fall 2004
\item
  Spring 2005
\item
  Fall 2005
\item
  Spring 2006
\item
  Spring 2007
\item
  Spring2008
\item
  Spring 2009
\item
  Spring 2010
\item
  Spring 2011
\item
  Fall 2012
\item
  Fall 2013
\item
  Fall 2014
\item
  Fall 2015
\item
  Fall 2016
\item
  Spring 2017
\item
  Fall 2017
\item
  Spring 2018
\item
  Fall 2018
\item
  Spring 2019
\item
  Fall 2019
\item
  Spring 2020
\item
  Spring 2021
\item
  Spring 2022
\item
  Spring 2023
\item
  Fall 2023
\end{itemize}

The following students have helped extensively in the writing of these
notes:

\begin{itemize}
\tightlist
\item
  Ujjwal Pandey
\item
  Ryan Frappier
\item
  Andrew Woods
\end{itemize}

Many students have pointed out typos including

\begin{itemize}
\tightlist
\item
  Kevin Weston
\item
  Saurav Bhattarai
\item
  Eric Garcia
\item
  Cole Schwandt
\item
  Ryan Harvey
\item
  Etc.
\end{itemize}

The list is not complete. If you did point out an error but is not
mentioned, please let me know. I think maybe in the future I should
include names of contributors in the right margin.

\end{document}
