%-*-latex-*-
\sectionthree{Key and satellite data}
\begin{python0}
from solutions import *; clear()
\end{python0}

In the case of an array, such as an array of integers representing
the digits of a long integer, the index provides a search \lq\lq key''.
For instance \lq\lq the third digit of the long integer
represented by \texttt{x}'' (i.e., the hundreds)
means \verb!x[2]!.
In this case \lq\lq third'' means \lq\lq 2'' which leads us to
\verb!x[2]!.

In many applications, the key can be different (i.e., not an index value).
For instance suppose I have array of students:
\begin{Verbatim}[frame=single, fontsize=\footnotesize]
#include <iostream>
#include <string>

class Student
{
public:
    Student()
    {}
    
    Student(const std::string & id,
            const std::string & firstname,
            const std::string & lastname,
    : id_(id),
      firstname_(firstname),
      lastname_(lastname)
    {}
    
private:
    std::string id_;         // ----> key
    
    std::string firstname_;  // \
    std::string lastname_;   //  +--> satellite data
    // dob, address, etc.    // /
};

int main()
{
    const int CAPACITY = 1000;
    Student student[CAPACITY];
    int n = 0;
    
    // Set student[0], student[1], student[2], ...,
    // student[94].
    // Set n to 95
    // Altogether we have 95 students in the system.

    return 0;
}  
\end{Verbatim}


In this case, search for a particular student might depend on
supplying the system with a student id.
The value of the student id (for each object in the array)
is stored in instance variable
\verb!id_!.

In this case student id is the key.
A field that uniquely identifies a value in a container of values
is called \defone{key}.
The other data which is not used to uniquely identify a
particular entry in the array is called \defone{satellite} data.
In the above, the first name and last name for each
entry in the array is satellite data of that entry.
The value of firstname and lastname does not uniquely identify
a student in the array since it's possible to have two students
named John Doe. 
Right?
In a real information system of students, one would expect
a lot more satellite data than key data.

The key problem is how to design containers so that
when given a key and satellite, you can quickly put the
(key, satellite) into the container,
when given a key, how to find the (key, satellite) in the container, and 
when given a key, how to delete the (key, satellite) from the container
(if it's found at all).
In particular, we will focus mainly on the organization of the keys
in the container so that the operations are as fast as possible.

For large scale data, it's common to store the satellite data
separate from the key data.
In that case a \verb!Student! object might have key data and
some kind of data that leads to its satellite data.
Why?
Because then a huge amount of key data can be loaded into memory
from harddrive so that search can be as fast as possible.
Once the student's id is found, then only this student's
satellite data is read from harddrive. 
Situations like this do occur.
See for instance CISS430 (database).
