\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{\texttt{if}-\texttt{else} statements}

Run this twice by entering 42 and then 24 for input:
\begin{console}
int a = 0;
std::cout << "C++: What's your favorite number?"
          << std::endl << "User: ";
std::cin >> a;

if (a == 42)
   std::cout << "C++: Mine too!" <<std::endl;
if (a != 42)
   std::cout << "C++: boo!" << std::endl;
\end{console}

You want to either do one thing or another based on whether a boolean
condition is true or false (respectively). The above does work ...
\textit{but} it is the \EMPHASIZE{WRONG} way of doing it.

Now change the above to this:
\begin{console}[commandchars=\~\@\$]
int a = 0;
std::cout << "C++: What's your favorite number?"
          << std::endl << "User: ";
std::cin >> a;

if (a == 42)
   std::cout << "C++: Mine too!" << std::endl;
~textbf@else$
   std::cout << "C++: boo!" << std::endl;
\end{console}

You should \EMPHASIZE{NEVER} have a program like the earlier one with two
if's with opposite conditions next to each other:
\begin{console}
if (a == 42)
   std::cout << "C++: Mine too!" <<std::endl;
if (a != 42)      // BAD!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
   std::cout << "C++: boo!" << std::endl;
\end{console}

In other words if you can rewrite a bunch of ifs with if-else, then you
should. Why? Because this avoids evaluating boolean expressions
unnecessarily and hence speed up your program. (There are other
reasons).

The \EMPHASIZE{second} way (the if-else) is the \EMPHASIZE{right} way:

\begin{console}
if (a == 42)
   std::cout << "C++: Mine too!" << std::endl;
else
   std::cout << "C++: boo!" << std::endl;
\end{console}

The format of the if-else statement is
\begin{center}
\EMPHASIZE{if (\texttt{[bool expr]})}\\
\EMPHASIZE{\texttt{[stmt1]}}\\
\EMPHASIZE{else}\\
\EMPHASIZE{\texttt{[stmt2]}}\\
\end{center}

If \textit{\texttt{[bool expr}} evaluates to \texttt{true}, \textit{\texttt{[stmt1}}
executes. When that's completed, the else part is
bypassed. If \textit{\texttt{[bool expr}} evaluates to \texttt{false}
\textit{\texttt{[stmt2}} is executed.

Obviously you can't write an else statement without an if!!!

%\textbf{Exercise.} 
<<<<<<< HEAD
\input{../exercises/0-08-if-part2/question.tex}
=======
\begin{ex}
Here's the first if-else warm up exercise. No sweat required to finish it. First try running this program so that you know what it does:
\begin{console}
int numHeads = 0;
std::cout << "How many heads do you have? ";
std::cin >> numHeads;
if (numHeads == 2)
   std::cout << "Are you Zaphod?" << std::endl;
if (numHeads != 2)
   std::cout << "You are not Zaphod" << std::endl;
\end{console}
\end{ex}

>>>>>>> parent of 3043c95 (.)
Next modify it to use if-else instead of two if's. Run
the program and test it. Here are some tests:
\begin{console}[commandchars=\~\@\$]
How many heads do you have? ~textbf@1$
You are not Zaphod
\end{console}

And here's an execution of the same program:

\begin{console}[commandchars=\~\@\$]
How many heads do you have? ~textbf@2$
Are you Zaphod?
\end{console}

<<<<<<< HEAD
\input{../exercises/1-08-if-part2/question.tex}Rewrite the program using if-else.

\input{../exercises/2-08-if-part2/question.tex}Here's another:
=======
\begin{ex}
Run the following program:
\begin{console}
std::cout << "How much do you have? ";
double savings = 0.0;
std::cin >> savings;

if (savings > 100)
   std::cout << "Can I borrow a ten?" << std::endl;
if (savings <= 100)
   std::cout << "Life is tough" << std::endl;
\end{console}
\end{ex}
Rewrite the program using if-else.

\begin{ex}
Write a program that prompts the user for an integer and tells you if the input is even or odd. Here's an execution:
\begin{console}[commandchars=\~\@\$]
gimme an int ~textbf@~underline@42$$
even!!!
\end{console}
\end{ex}
Here's another:
>>>>>>> parent of 3043c95 (.)
\begin{console}[commandchars=\~\@\$]
gimme an int ~textbf@~underline@35$$
odd!!!
\end{console}

<<<<<<< HEAD
\input{../exercises/3-08-if-part2/question.tex}Here's another
=======
\begin{ex}
No skeleton code here. This is a slight modification
of a previous program. Write a program that prompts the user for his/her
height (in ft) and weight (in lbs) and prints \lq\lq you will live more than
100 years!!!'' if the product of the height and weight is at least
200.45. (This is not supported by any form of research.) Otherwise it
prints \lq\lq you will not live more than 100 years''. Here's
an execution of the program:
\begin{console}[commandchars=\~\@\$]
~textbf@6.1 180.180$
you will live more than 100 years!!!
\end{console}
\end{ex}
Here's another
>>>>>>> parent of 3043c95 (.)
\begin{console}[commandchars=\~\@\$]
~textbf@3.4 23.4$
you will live more than 100 years!!!
\end{console}

\textbf{Exercise.}
<<<<<<< HEAD
\input{../exercises/4-08-if-part2/question.tex}
=======
\begin{ex}
 Write a program that prompts the user for his
employee code which is either 0 or 1. If the employee code is 0, his/her
hourly rate is \$123.45. If the employee code is 1, his/her hourly rate
is \$7.99. Prompt the user for the number of hours he/she worked this
week. Finally print the total amount he/she made this week.
Here's an informal description of the program that you
should follow (obviously you have to translate this to C++!!!):
\begin{console}
Declare constant HOURLY_RATE_0 and initialize it to 123.45.
Declare constant HOURLY_RATE_1 and initialize it to 7.99.

Declare variable employee_code.
Get employee_code from user.

Declare variable hourly_rate.
If employee_code is 0:
   set hourly_rate to HOURLY_RATE_0
Otherwise:
        set hourly_rate to HOURLY_RATE_1
        
Get number_of_hours from user.
Print total pay for this week.
\end{console}
\end{ex}

>>>>>>> parent of 3043c95 (.)
