\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{The dangling else problem}

Here's a trick question (and it's another reason why C/C++/Java programmers use { } even when the block has only one statement). So ...
\begin{center}
\EMPHASIZE{PAY ATTENTION}
\end{center}
First figure out the output (without a compiler) when x has a value of
0.

\begin{console}
if (x < 0)
   if (x == -2) std::cout << "1" << std::endl;
else
   std::cout << "2" << std::endl; 
\end{console}
Now run the program.

Mamamia ... ?!?

This is called the \lq\lq dangling else'' problem.

What actually happens is that an \texttt{else} always follows the closest
\texttt{if}. This means that the above code is better indented this way:
\begin{console}
if (x < 0)
   if (x == -2) std::cout << "1" << std::endl;
   else
       std::cout << "2" << std::endl; 
\end{console}

Of course the C++ compiler doesn't care about
whitespaces. However this version of indentation makes it easier for
human beings to follow the logic.

Now if what I intended was actually something like the previous version,
then I would have to do this:
\begin{console}
if (x < 0)
{
   if (x == -2) std::cout << "1" << std::endl;
}
else
    std::cout << "2" << std::endl; 
\end{console}
In this case, the closest if-part that the else-part can attach to is
the if-part on the first line since the else-part cannot \lq\lq go inside''
the block.

This is the reason why it's a good practice to ALWAYS
use blocks even when the if-part or the else-part has only one line to
execute. It prevents the misleading dangling else problem. So if I write
this:
\begin{console}
if (x < 0)
{
   if (x == -2)
   {
       std::cout << "1" << std::endl;
   }
}
else
{
     std::cout << "2" << std::endl;
}
\end{console}

there will be no confusion.

