\begin{python0}
from solutions import *; clear()
\end{python0}
<<<<<<< HEAD
\input{../exercises/24-08-if-part2/question.tex}\sectionthree{The ternary operator}
=======
\sectionthree{The ternary operator}
>>>>>>> parent of 3043c95 (.)

So far our operators accept one value, for example:
\begin{align*}
&\text{$-$y}\\
&\text{!b}\\
\end{align*}
or two values, for example:
\begin{align*}
& \text{x + y}\\
& \text{b1 \&\& b2}\\
\end{align*}
There is a curious operator that accepts \texttt{\textbf{three}}. Try this
program:
\begin{console}
int x = 5;
std::cout << (x == 5 ? 42 : 0) << std::endl; 
\end{console}

Run this with different values for variable x.

Get the point of the \texttt{(?:)} operator yet? The format looks like
this:
\begin{center}
\EMPHASIZE{(\texttt{[bool expr} ? \texttt{[expr1]} : \texttt{[expr2]})}
\end{center}
Basically, if the \texttt{[bool expr]} (a boolean expression)
evaluates to \texttt{true}, C++ will evaluate \texttt{[expr1]} and give
you its value; otherwise it gives you the value from \texttt{[expr2]}.

Here's another example. First try to guess what the
program does. Next run it to verify.
\begin{console}[commandchars=\~\@\$]
int x = 0, y = 1, z = 2;

~textbf@int a = (x + y  z ? x : y + z);$
std::cout << a << std::endl;
\end{console}

Do you see that it does the same thing as this:
\begin{console}[commandchars=\~\@\$]
int x = 0, y = 1, z = 2;

int a = 0;
~textbf@if (x + y < z)$
~textbf@{$
    ~textbf@a = x;$
~textbf@}$
~textbf@else$
~textbf@{$
    ~textbf@a = y + z;$
~textbf@}$
std::cout << a << std::endl; 
\end{console}

Here's another example. Try this die game:
\begin{console}
int die_roll = 1 + rand() % 6 ;
std::cout << "value of die roll: " <<
          die_roll << '\n';
std::cout << "You " 
          << (die_roll == 6 ? "win!" : "lose")
          << '\n'; 
\end{console}
Get it?

<<<<<<< HEAD
\input{../exercises/25-08-if-part2/question.tex}
\input{../exercises/26-08-if-part2/question.tex}
\input{../exercises/27-08-if-part2/question.tex}
=======
\begin{ex}
Write a dice game. Roll two dice. If the dice both
give you the same number, print \lq\lq win''. Otherwise your program prints
\lq\lq lose''. Your program should also print the values of the dice roll.
Use the ternary operator.
\end{ex}

\begin{ex}
Write a dice-coin game. First, you roll the dice. If
the dice values are the same you gain \$1.50 and the game ends.
Otherwise you pay \$1. However you have a second chance: you toss a
coin. If you get a head (use random integer value of 0) you gain \$1,
otherwise (i.e., for a tail), you gain \$0.50. Your program prints all
the dice values and the result of the coin toss and finally your gain
(which can be positive or negative). Run this program 10 times. What is
you average gain? Use the ternary operator whenever possible.
\end{ex}

\begin{ex}
Rewrite the following code fragment so that it's
cleaner
\begin{console}
int x, y;
std::cin >> x >> y;
if (x >= 42)
{
    std::cout << x << ' ';
    if (y < 10)
    {
        std::cout << y << ' ';
    }
    else
    {
        std::cout << '-' << ' ';
    }
}
else
{
     std::cout << '*' << ' ';
     if (y < 10)
     {
        std::cout << y << ' ';
     }
     else
     {
        std::cout << '-' << ' ';
     }
}
\end{console}
\end{ex}

>>>>>>> parent of 3043c95 (.)
