\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Maximum/minimum computation}

Recall I have already talked about the max/min computation. There are
two ways of doing it.

Here's the max computation that you have already seen:
\begin{console}
int w = 1, x = 3, y = 0, z = 5;
int max;

max = w;
if (max < x)
   max = x;
if (max < y)
   max = y;
if (max < z)
   max = z; 
\end{console}

Notice that other than the first statement, the rest of the code is more
or less a repetition of the same kind of code:
\begin{console}
max = w;
if (max < [variable])
   max = [variable]; 
\end{console}

There's another \EMPHASIZE{TERRIBLE} way to do the above:
\begin{consolethree}
int w = 1, x = 3, y = 0, z = 5;
if (w >= x && w >= y && w >= z)
    max = w;
else
{
     if (x >= w && x >= y && x >= z)
         max = x;
     else
     {
        if (y >= w && y >= x && y >= z)
            max = y;
        else
            max = z;
      }
}
\end{consolethree}

Of course you can try to make it ``neater'':
\begin{console}
int w = 1, x = 3, y = 0, z = 5;

if (w >= x && w >= y && w >= z)
    max = w;
else if (x >= w && x >= y && x >= z)
    max = x;
else if (y >= w && y >= x && y >= z)
    max = y;
else
    max = z;
\end{console}

But you are computing way too many boolean conditions. If z is the
maximum value, how many boolean operators are you evaluating? Compare
this to the first max computation.

Secondly, and this is probably even more important:

\begin{ex}
Suppose you need to add another variable to your w,
x, y, z for max computation:
\begin{console}
int w = 1, x = 3, y = 0, z = 5, a = 7;
int max;

// ... max computation ... 
\end{console}
\end{ex}
Complete this using the first method and then using the second method.
Do I need to explain anymore that the first method is way better?

Part of the problem is that the boolean expression in
\begin{console}
if (w >= x && w >= y && w >= z)
    max = w;  
\end{console}
depends on a huge number of variables.

The earlier better solution is to consider one variable at a time.

In general if the boolean is made up of expressions which \EMPHASIZE{do not depend on each other}, it's usually best to process them \EMPHASIZE{separately}.

