\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{\texttt{if-else} blocks}

It should not be too surprising that you can execute blocks for if-else.
You can have a block controlled by the if-part, or a block controlled by
the else-part, or both.

\begin{console}
std::cout << "Head Surgery Cost Planner \n"
          << "First 3: \$100 each. \n"
          << "After that: \$2 each. \n"
          << "Less than 3: extra 500 charge"
          << std::endl;

int extraHeads = 0;
int cost = 0;
std::cout  "How many extra heads do you want? ";
std::cin >> extraHeads;
if (extraHeads <= 3)
{
        cost = extraHeads * 100 + 500;
        std::cout << "Please go to Cheapskate Room."
                  << std::endl;
}
else
{
        cost = 3 * 100 + (extraHeads - 3) * 2;
        std::cout << "Please go to Deluxe Room."
                  << std::endl;
}
\end{console}
\begin{ex}
Once you see code like the above you immediately ask
yourself this question: How many constants do you see in the above
program? (I can see four). Rewrite it with constants instead of
hard-coding integer values.
\end{ex}
\begin{ex}
Write a program that computes the area of either a
circle or a right-angle triangle. Recall that you can declare character
variables. It turns out that you can compare character values too, using
the \texttt{==} operator. The type for the character values is
\texttt{char}. Here's a code skeleton for you:
\begin{console}
std::cout << "Area Calculator!!! \n";
std::cout << "I can do circles or triangles! \n"
          << "Option (c or t): ";
char >> option;
std::cin >> option;
if (option == 'c')
{
        // Prompt the user for radius and display the
        // area of the circle with the given radius.
}
else    // if it's not 'c', assume it is 't'
{
        // Prompt the user for the base and height of
        // the right-angled triangle and display the
        // area.
}
\end{console}
\end{ex}

