\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Boolean expressions again}

We started with boolean expressions such as

\[x > 0\]

and went on to expressions such as

\[x > y\]

and then to

\[x * x + y * y != 0\]

and finally to

\[\texttt{a * a + b * b  c * c + d * d}\]

i.e.
\begin{center}
\EMPHASIZE{\texttt{[expr1] [boolop] [expr2]}}
\end{center}
where \texttt{[expr1]} and \texttt{[expr2]} are boolean expressions
and \texttt{[boolop]} is a binary boolean operator accepting two
numeric values. Let's recall a few more binary operators.

First of all let's not forget that we also have the unary ! operator. The following prints x if \textbf{it is not true that x is 0}.
\begin{console}[commandchars=\~\@\$]
int x = 0;
std::cin >> x;
if (~textbf@!(x == 0)$)
   std::cout << x << std::endl; 
\end{console}

<<<<<<< HEAD
\input{../exercises/9-08-if-part2/question.tex}\input{../exercises/10-08-if-part2/question.tex}
=======
\begin{ex}
Rewrite the above program so that the ! operator is
not used. Your resulting program therefore has to evaluate one boolean
operator and not two.
\end{ex}
\begin{ex}
Simplify the boolean expression in the program so
that the program behaves the same after your modification:
\begin{console}
int x = 0;
std::cin >> x;
if (!(x <= 42) || !(x >= 42))
   std::cout << x << std::endl;
\end{console}
\end{ex}

>>>>>>> parent of 3043c95 (.)
But note you should not simplify expressions (boolean or numeric) if by
doing so it makes your program harder to read. Furthermore, modern-day
compilers are actually smart enough to simplify it for you
automatically!

Of course we also have the boolean operators \texttt{\&\&} and
\texttt{\textbar\textbar}. So we can have the following boolean
expression:
\begin{center}
$(x < y)$   \&\&  $(y < z)$
\end{center}
Note that \texttt{(x < y)} and \texttt{(y < z)} are
boolean expressions. So in general we can have
\begin{center}
\EMPHASIZE{\texttt{[boolexpr1] [boolop] [boolexpr2]}}
\end{center}
where \texttt{[boolexpr1]} and \texttt{[boolexpr2]} are boolean
expressions and \texttt{[boolop]} is either \&\& or \textbar\textbar.

This should be a DIY review:

<<<<<<< HEAD
\input{../exercises/11-08-if-part2/question.tex}
\input{../exercises/12-08-if-part2/question.tex}You should check with your previous notes.

\input{../exercises/13-08-if-part2/question.tex}You should create some test cases by hand and then test your code
=======
\begin{ex}
Complete the truth table for \&\&:

\begin{align*}
& \text{true \&\& true is \_\_\_\_\_}\\
& \text{true \&\& false is \_\_\_\_\_}\\
& \text{false \&\& true is \_\_\_\_\_}\\
& \text{false \&\& false is \_\_\_\_\_}\\
\end{align*}
\end{ex}

\begin{ex}
Complete the truth table for \textbar\textbar:

\begin{align*}
& \text{true \textbar\textbar true is \_\_\_\_\_}\\
& \text{true \textbar\textbar false is \_\_\_\_\_}\\
& \text{false \textbar\textbar true is \_\_\_\_\_}\\
& \text{false \textbar\textbar false is \_\_\_\_\_}\\
\end{align*}
\end{ex}
You should check with your previous notes.

\begin{ex}
You qualify for free head surgery if you have one
head, less than three arms, more fingers on your left than you right
hand, and finally either you are from Earth or the product of your
highest level reached on Pacman and Galaxian is less than 6 (obviously
you need an extra head in that case ...) Of course if you have not
played Pacman, your high level is 0; likewise for Galaxian. Complete
this program:
\begin{console}
int numHeads = 0;
int numArms = 0;
int numLeftFingers = 0;
int numRightFingers = 0;
bool earthling = true;
int pacManLevel = 0;
int galaxianLevel = 0;

// prompt for values for all variables
// For boolean variable enter 1 for true and 0 for false

if (                             )
std::cout << "You qualify for our free head surgery!"
          << std::endl;
else
std::cout << "Try again at newheads.com" << std::endl;
\end{console}
\end{ex}
You should create some test cases by hand and then test your code
>>>>>>> parent of 3043c95 (.)
against your test cases.

