%-*-latex-*-
\input{myassignmentpreamble}
\input{ciss358}
\input{yliow}
\renewcommand\TITLE{Assignment 3}

\begin{document}
\topmatter

\textsc{Objectives}
\begin{itemize}
  \li Review of proof using induction and WOP. 
  \li Prove the correctness of an algorithm using loop invariants.
\end{itemize}


%------------------------------------------------------------------------------
\newpage
Q1.
Recall from class, I mentioned that there are some proofs
that you must know because these proofs almost form a template for many other
proofs.
The goal of this question is not to design a proof,
but rather to re-study a given proof very carefully and write an absolutely
correct proof in your own words and possible filling gaps if necessary.

Prove the following using WOP.

\textsc{Euclidean property of $\N$.}
If $a, b \in \N = \{0, 1, 2, ...\}$ with $b \leq 0$,
then there exist $q$ and $r$ in $\N$ such that
\[
a = bq + r, \,\,\, 0 \leq r < b
\]
(More is true: $q,r$ are unique. The proof is not difficult
and if probably proven in discrete 1. I definitely talk about it in 451.)

\SOLUTION

\input{q01.tex}

%------------------------------------------------------------------------------
\newpage
Q2. Prove using invariants and induction that the following algorithm
(bottom-up recursion for exponentiation) is
correct.
\begin{console}
ALGORITHM: power
INPUTS: x - base
        e - exponent
OUTPUT: x^e, i.e., x raised to power of e

ret = 1
while e > 0:
    if e % 2 == 1:
        ret = ret * x
    e = e / 2
    x = x * x
    
return ret
\end{console}
(Note: \texttt{/} is integer division.)

\SOLUTION

\input{q02.tex}

%------------------------------------------------------------------------------ 
\newpage
Q3.
Prove that the selection sort algorithm is correct using invariants
and induction.

\SOLUTION

\input{q03.tex}

\end{document}

