\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{An easy simplification}

Here's something you should know: If x is a boolean variable. Then

\tab[3em]{x}

and

\tab[3em]{x == true}

have the same boolean value. That's not too surprising since the
following pairs of statements are the same:

\begin{tabular}{p{5cm} p{7cm}}
“I'm broke.” & “It is true that I'm broke.”\\
“I'm ten feet tall.” & “It is true that I'm ten feet tall.”\\
“I have a pet tiger.” & “It is true that I have a pet tiger.”\\
\end{tabular}

Right?

Let me try to prove that. There are only two possible values for x: it's
either \verb!true! or \verb!false!.

CASE: x is \verb!true!. In that case

\tab[3em]{x == \verb!true!}\\
\tab[3em]{= \verb!true! == \verb!true!}\\
\tab[3em]{= \verb!true!}\\

which is the value of x.
CASE: x is false. In that case

\tab[3em]{x == \texttt{true}}
\tab[3em]{= \texttt{false} == \texttt{true}}
\tab[3em]{= \texttt{false}}

which is, again, the value of x.

So in all cases,

\tab[3em]{x}

and

\tab[3em]{x == \texttt{true}}

have the same boolean value.

Similarly

\tab[3em]{!x}

and

\tab[3em]{x == \texttt{false}}

have the same boolean value

\textbf{Exercise.} Here's a boolean expression for “It is true that the earth is flat, it is true that the speed of light is greater than 180000 miles/second, it is true that I have a pet tiger, and it is false that I cannot tie shoelaces.”:


\tab[3em]{(\texttt{earth\_is\_flat} == \texttt{true})}\\
\tab[3em]{\&\& ((\texttt{LIGHT\_SPEED} $>$ 180000) == \texttt{true})}\\
\tab[3em]{\&\& (\texttt{have\_pet\_tiger} == \texttt{true})}\\
\tab[3em]{\&\& (\texttt{cannot\_tie\_shoelaces} == \texttt{false})}\\


Simplify it (so that == does not appear in your simplified expression.)
Test it with a C++ program.

