\tinysidebar{\debug{exercises/{test/question.tex}}}
\begin{ex}
Here's the header file for weather control devices:

\begin{consolethree}[escapeinside=||]
// WeatherCtrl.h
class WeatherCtrl
{
public:
    WeatherCtrl(double, double);
    double get_temp();
    double get_pressure();
    void set_temp(double);
    void set_pressure(double);
private:
    double temp_;
    double pressure_;
};
\end{consolethree}

Write a cpp file containing the method implementation (i.e., definition)
for all the prototypes that appear in the header file.

Next go over the following \texttt{main.cpp} that uses the weather control
class:

\begin{consolethree}[escapeinside=||]
// main.cpp
int main()
{
// Declare and initialize wc as a WeatherCtrl object with
// initial temperature reading of 50.5 and pressure reading
// of 30.5

// Print the temp and pressure value of wc using std::cout

// Set the temp and pressure value of wc to 60 and 40

// print the temp and pressure value of wc using std::cout

// Uncomment the next statement and compile. What's wrong?
// WeatherCtrl wc2;

// Comment out the above wrong statement and
// move on ...

// Modify the constructor so that it's inlined in
// the header and defaults the temp and pressure
// value to 50 and 60. Use an initializer list.

// Uncomment the next line, compile and run.
// WeatherCtrl wc2;

// Print the temp and pressure value of wc2.

// Declare wc3 as a pointer to WeatherCtrl and
// initialize it to point to a WeatherCtrl object in the
// heap with 110 and 70 for temp and pressure
// values.
// Print the temp and pressure of the object wc3 is
// pointing to.

    return 0;
}
\end{consolethree}
\end{ex}
