\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Object-oriented Thingies}

So far you have been using C++ in a particular way. The programming
style (or paradigm) you have been using is called \EMPHASIZE{structured
programming}.

You can recognize structured programming when you see blocks of code
including but not limited to the following:

\begin{itemize}
\item branching such as if and if-else
\item loops such as for-loop, while-loop
\item functions
\end{itemize}

But this is not the only paradigm. The next paradigm we will focus on is
called \EMPHASIZE{Object-Oriented Programming, OOP}.

(There are many other styles/paradigms of programming. For instance, you
will learn functional programming in CISS445. Then
there's aspect-oriented programming ...)

Both structured programming and OOP came out around 60s. Structured
programming was popularized in the 80s through the introduction of the
Pascal language by N. Wirth. (I learned that when I was in college.)
Although OOP also came out around that time, it took off only around the
90s. So in terms of widespread use, OOP is actually pretty ``new'' or
``recent''. (But in the computer science world, anything older than 5
years is considered ``old'' ...)

While structured programming and OOP refer to using certain language
features to \EMPHASIZE{write} programs, structured analysis and
object-oriented analysis refers to high level \EMPHASIZE{analysis} of a
program (usually using diagrams) with a view toward implementing the
ideas using a structured or an object-oriented language respectively.

Some research shows that once a program goes beyond 100,000 lines,
structured programming methodology breaks down because of the complexity
of the project. (That's not to say that it cannot be
done and that's also not saying that OO is the only way
to handle large projects.)

Instead of going through all the philosophical reasons why OO is
superior to old-style programming (such as structured programming), as
many books do, we'll go through examples. It's pointless
to discuss philosophy first because if you have not seen OO language
features, it would be hard to judge. So ...


