\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Date class version 1}

We want to implement a \texttt{Date} type.

First I use \texttt{struct} to package up year, month, day and have some
useful functions:

\begin{itemize}
\item \texttt{init} to set the month, day, year value of a \texttt{Date}
  variable
\item \texttt{add\_y} to add a year increment value to \texttt{Date} variable
\item \texttt{add\_m} to add a month increment value to \texttt{Date} variable
\item \texttt{add\_d} to add a day increment value to \texttt{Date} variable
\end{itemize}

Here we go (and this is nothing new) ...

\begin{consolethree}[escapeinside=||]
// Date.h
#ifndef DATE_H
#define DATE_H

struct Date
{
    int yyyy_, mm_, dd_;
};

void init(Date &, int, int, int);
void print(const Date &);
void add_y(Date &, int);
void add_m(Date &, int);
void add_d(Date &, int);

#endif
\end{consolethree}

\begin{consolethree}[escapeinside=||]
// Date.cpp
#include "Date.h"

void init(Date & date, int yyyy, int mm, int dd)
{
    date.yyyy_ = yyyy;
    date.mm_ = mm;
    date.dd_ = dd;
}

void print(const Date & date)
{
    std::cout << date.yyyy_ << '/'
              << date.mm_ << '/'
              << date.dd_ << '\n';
}

void add_y(Date & date, int inc)
{
    date.yyyy_ += inc ;
}

// etc.
\end{consolethree}

\begin{python}
from latextool_basic import *
p = Plot()
#p += Grid(0, 0, 10, 10)
strink = r"""
\begin{consolethree}[escapeinside=||]
// main.cpp
#include "Date.h"

int main()
{
    Date today, yesterday;

    init(today, 2014, 12, 25);
    print(today);
    
    init(yesterday, 2014, 12, 24);
    print(yesterday);

    return 0;
}
\end{consolethree}
"""
p += Rect(0, 0, 8, 5, s=strink, linewidth=0)
td = [r'\textbf{today}', r'\textbf{yyyy\_}', r'\textbf{mm\_}', r'\textbf{dd\_}']
p += Rect(7, 4, 9, 5, innersep=0.25, s=td[0], align='c', linewidth=0)
x = 9
for i in range (1, 4):
      p += Rect(x, 4, x + 2, 5, innersep=0.25, s=td[i], align='c', linewidth=0.05, linecolor='red')
      x += 2

p += Rect(9, -1, 13.5, 2, innersep=0.25, s=r'At this point in \texttt{main()\'}s memory, \texttt{today} looks like this... ', align='c', linewidth=0.05)
p += Line(points=[(12, 2), (12, 4)], linewidth=0.05, linecolor='red', endstyle='>')
p += Line(points=[(9, 1), (6.75, 1.75)], linewidth=0.05, linecolor='red', endstyle='>')
print(p)
\end{python}

Now in your \texttt{main.cpp}, go ahead and call \texttt{add\_d(yesterday,
1)} and then \texttt{print(yesterday)}.

Note that the above \texttt{main()} that uses the \texttt{Date} type
(\texttt{Date.h} and \texttt{Date.cpp}) does not need to know about the
details (i.e., the members) of a \texttt{Date} variable: you do not need
to know that a \texttt{Date} variable such as \texttt{today} contains
\texttt{yyyy\_}, \texttt{mm\_}, \texttt{dd\_}.

This is the beginning of the concepts of \EMPHASIZE{information hiding} and
\EMPHASIZE{encapsulation}.

Specifically, in the above, we collect up the concepts of a year, a
month and a day and create the concept of a date. That's
\EMPHASIZE{encapsulation}.

Encapsulation allows you to \EMPHASIZE{think at a higher level} of
abstraction. It's easier to focus on higher level
concepts once lower level details are encapsulated.

Note that in version 1 the functions are not tied to \texttt{Date} struct.

We want to think of the functions as ``belonging'' to the \texttt{Date}
struct.

You can actually put the functions into the \texttt{struct} definition to
get \EMPHASIZE{member functions} .

So we get a \texttt{struct} with \EMPHASIZE{member variables} and
\EMPHASIZE{member functions} in one package.

After moving the functions into the \texttt{struct}, you need to make some
modifications.

\input{../exercises/0-60-classes-and-objects/question.tex}

