\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Date class version 2}

Make the following changes:

\EMPHASIZE{In }\texttt{\EMPHASIZE{Date.h}}\EMPHASIZE{:} move the function
prototypes into the \texttt{struct} and remove all first parameters which
are \texttt{Date} reference parameters.

\EMPHASIZE{In }\texttt{\EMPHASIZE{Date.cpp}}\EMPHASIZE{:} remove all first
parameters which are \texttt{Date} reference parameters and remove
``\texttt{date.}'' in the body of all functions.

\EMPHASIZE{In }\texttt{\EMPHASIZE{main.cpp}}\EMPHASIZE{:} move all \texttt{Date}
variables which are first arguments outside the function call. Add a dot
after these \texttt{Date} variables.
Here are the files with the changes

\begin{consolethree}[escapeinside=||]
// Date.h
#ifndef DATE_H
#define DATE_H
struct Date
{
    int yyyy_, mm_, dd_;
    void init(|\sout{Date \&,}| int, int, int);
    void print(|\sout{const Date \&}|);
    void add_y(|\sout{Date \&, int}|);
    void add_m(|\sout{Date \&, int}|);
    void add_d(|\sout{Date \&, int}|);
};
#endif
\end{consolethree}

\begin{consolethree}[escapeinside=||]
// Date.cpp
#include "Date.h"
void |\EMPHASIZE{Date::}|init(|\sout{Date \& date,}| int yyyy, int mm, int dd)
{
    |\sout{date.}|yyyy_ = yyyy;
    |\sout{date.}|mm_= mm;
    |\sout{date.}|dd_= dd;
}

void |\EMPHASIZE{Date::}|print(|\sout{const Date \& date}|)
{
    std::cout << |\sout{date.}|yyyy_ << '/'
              << |\sout{date.}|mm_ << '/'
              << |\sout{date.}|dd_<< '\n';
}

void |\EMPHASIZE{Date::}|add_y(|\sout{Date \& date,}| int inc)
{
    |\sout{date.}|yyyy_+= inc ;
}

// etc.
\end{consolethree}
\begin{consolethree}[escapeinside=||]
// main.cpp
#include "Date.h"
int main()
{
    Date today, yesterday;

    |\EMPHASIZE{today.}|init(|\sout{today,}| 2014, 12, 25);
    |\EMPHASIZE{today.}|print(|\sout{today}|);

    |\EMPHASIZE{yesterday.}|init(|\sout{yesterday,}| 2014, 12, 24);
    |\EMPHASIZE{yesterday.}|print(|\sout{yesterday}|);

    return 0;
}
\end{consolethree}
Here's what you get:

\begin{consolethree}[escapeinside=||]
// Date.h
#ifndef DATE_H
#define DATE_H

struct Date
{
    int yyyy_, mm_, dd_;

    void init(int, int, int);
    void print();
    void add_y(int);
    void add_m(int);
    void add_d(int);
};

#endif
\end{consolethree}

\begin{consolethree}[escapeinside=||]
// Date.cpp
#include "Date.h"

void Date::init(int yyyy, int mm, int dd)
{
    yyyy_ = yyyy;
    mm_ = mm;
    dd_ = dd;
}

void Date::print()
{
    std::cout << yyyy_ << '/'
              << mm_ << '/'
              << dd_ << '\n';
}

void Date::add_y(int inc)
{
    yyyy_ += inc ;
}

// etc.
\end{consolethree}

\begin{python}
from latextool_basic import *
p = Plot()
#p += Grid(0, 0, 10, 10)
strink = r"""
\begin{consolethree}[escapeinside=||]
// main.cpp
#include "Date.h"

int main()
{
    Date today, yesterday;

    |\textbf{today.init(2014, 12, 25);}|
    today.print();
    
    yesterday.init(2014, 12, 24);
    yesterday.print();

    return 0;
}
\end{consolethree}
"""
p += Rect(0, 0, 8, 5, s=strink, linewidth=0)
td = [r'\textbf{today}', r'\textbf{yyyy\_}', r'\textbf{mm\_}', r'\textbf{dd\_}', r'\textbf{init}', r'\textbf{print}', r'\textbf{add\_y}', r'\textbf{add\_d}', r'\textbf{add\_m}']
p += Rect(7, 5, 9, 6, innersep=0.25, s=td[0], align='c', linewidth=0)
x = 9
for i in range (1, 4):
      p += Rect(x, 6, x + 2, 7, innersep=0.25, s=td[i], align='c', linewidth=0.05, linecolor='red')
      x += 2
x = 9
for i in range (4, 7):
      p += Rect(x, 5, x + 2, 6, innersep=0.25, s=td[i], align='c', linewidth=0.05, linecolor='red')
      x += 2

p += Rect(9, 4, 12, 5, innersep=0.25, s=td[8], align='c', linewidth=0.05, linecolor='red')
p += Rect(12, 4, 15, 5, innersep=0.25, s=td[7], align='c', linewidth=0.05, linecolor='red')

p += Rect(9, -1, 13.5, 2, innersep=0.25, s=r'At this point in \texttt{main()\'}s memory, \texttt{today} looks like this... ', align='c', linewidth=0.05)
p += Line(points=[(12, 2), (12, 4)], linewidth=0.05, linecolor='red', endstyle='>')
p += Line(points=[(9, 1), (6.75, 1.75)], linewidth=0.05, linecolor='red', endstyle='>')
print(p)
\end{python}

As stated earlier, we put the function prototypes into the \texttt{struct} definition.

Note that in \texttt{Date.cpp}, we need to put \EMPHASIZE{Date::} in
front of the implementation of the member functions. This is because the
\texttt{Date} \texttt{struct} creates a scope. Therefore outside the
\texttt{struct}, you need to say ``the \texttt{init} function \EMPHASIZE{of
}\texttt{\EMPHASIZE{Date}}''. If you just call it \texttt{init,} C++ will think
of some \texttt{init} function outside the \texttt{Date} \texttt{struct}. This means that
C++ actually allows you to do this:
\begin{consolethree}
// Date.h

...

struct Date
{
    ...
    void init(int, int, int);
    ...
};

void init(int, int, int);

#endif
\end{consolethree}
In the code, you can see the concept of ``variable calls a function''.
For instance, you see \texttt{today.init(2014, 12, 25)}.

In the implementation of \texttt{Date::init}, you see \texttt{yyyy\_},
\texttt{mm\_}, and \texttt{dd\_}. Which \texttt{yyyy\_} are we talking about?
There's the \texttt{yyyy\_} of \texttt{today} and the
\texttt{yyyy\_} of \texttt{yesterday}!!! Well, in the body of
\texttt{Date::init},
\begin{center}
\texttt{yyyy\_} refers to \texttt{yyyy\_} of the variable calling \texttt{init()}

\texttt{mm\_} refers to \texttt{mm\_} of the variable calling \texttt{init()}

\texttt{dd\_} refers to \texttt{dd\_} of the variable calling \texttt{init()}
\end{center}
\begin{python}
from latextool_basic import *
p = Plot()
#p += Grid(0, 0, 10, 10)
str0 = r"""
\begin{consolethree}[escapeinside=||]
...

int main()
{
    ...

    |\EMPHASIZE{today.init(2014, 12, 25);}|

    ...
\end{consolethree}
"""
str1 = r"""
\begin{consolethree}[escapeinside=||]
void Date::init(int yyyy, int mm, int dd)
{
    |\EMPHASIZE{yyyy\_}| = y0;
    mm_ = m0;
    dd_ = y0;
}
\end{consolethree}
"""
str2 = r"""
\begin{consolethree}[escapeinside=||]
...

int main()
{
    ...

    |\EMPHASIZE{yesterday.init(2014, 12, 25);}|

    ...
\end{consolethree}
"""
p += Rect(-2, 0, 10, 3, linewidth=0, s=str2)

p += Rect(7, 5.75, 16.5, 6.5, linewidth=0, s=str1)

p += Rect(-2, 9, 10, 12, linewidth=0, s=str0)

p += Rect(10.5, 9, 16.5, 10.25, linewidth=0, s=r'While \\executing \textbf{\texttt{today.init(2014, 12, 25)}}, the \textbf{\texttt{yyyy\_}} refers to the \textbf{\texttt{yyyy\_}} of \textbf{\texttt{today}} ...')
p += Rect(10.5, 0, 16.5, 3.25, linewidth=0, s=r'While \\executing \textbf{\texttt{yesterday.init(2014, 12, 24)}}, the \textbf{\texttt{yyyy\_}} refers to the \textbf{\texttt{yyyy\_}} of \textbf{\texttt{yesterday}} ...')
p += Line(points=[(10.25, 9), (8.15, 9.15)], linewidth=0.075, linecolor='red', endstyle='>')
p += Line(points=[(12, 8), (12, 6.55)], linewidth=0.075, linecolor='red', endstyle='>')
p += Line(points=[(10.25, 2), (9, 0.65)], linewidth=0.075, linecolor='red', endstyle='>')
p += Line(points=[(13.5, 2.3), (13.5, 3.65)], linewidth=0.075, linecolor='red', endstyle='>')
print(p)
\end{python}


Note that
\begin{itemize}
\item functions are part of \texttt{Date}
\item \texttt{Date} variables can invoke \texttt{Date} member functions
\item Member functions ``know'' which variables invoked them and which
  \texttt{yyyy\_}, \texttt{mm\_}, \texttt{dd\_} they should work with.
\item For instance, \texttt{today.mm\_} : Refers to the member variable
  \texttt{mm\_} in \texttt{today}
\item For instance, \texttt{today.print()} : Calls/invokes the \texttt{print()}
  member function of \texttt{today} and while executing the function,
  \texttt{yyyy\_} refers to the \texttt{yyyy\_} of \texttt{today.}
\end{itemize}

Make sure you see the differences between version 1 and 2!!!

\input{temp/60-classes-and-objects/exercises/1-60-classes-and-objects/question.tex}
In the above exercise, you will have statements such as

\texttt{c3p0.moveRight(3);}

You can and \EMPHASIZE{should} think of the \texttt{Robot} variable
\texttt{c3p0} as having the \EMPHASIZE{ability} to \texttt{moveRight} by 3 on
its own. In other words, you want to think of the \texttt{Robot} as having
some \EMPHASIZE{autonomous ability} to perform the \texttt{moveRight}
operation. That is one very important philosophy behind packaging
functions into a \texttt{struct} so that they become members (i.e. member
functions) as the \texttt{struct} variable. In
fact, historically, the earliest example of a programming language that
allows this is from the MIT AI lab in the late 50s.

This is very different from say if you do

\texttt{Robot\_moveRight(c3p0, 3);}

Written this way, it reads more like \EMPHASIZE{your} program is
\EMPHASIZE{controlling} the \texttt{Robot c3p0} and making him/her/it move.

For complex software engineering, we want to analyze and develop
software not by focusing on data and functions separately, but rather by
focusing on data \EMPHASIZE{with} their functions. This is achieved in version 2
above. For instance, look at

\tab[3em]{\texttt{today.init(12, 25, 2003);}}

You think of the \texttt{init()} function as being part of \texttt{today}, just like you
have \texttt{today.mm\_}, \texttt{today.dd\_}, \texttt{today.yyyy\_} etc.

In fact, to emphasize again that \texttt{today} has the autonomous ability
to execute \texttt{init()}, I will frequently say:

\EMPHASIZE{today invokes init()}

and not ``your program invokes the \texttt{init()} of \texttt{today}''.

\input{temp/60-classes-and-objects/exercises/2-60-classes-and-objects/question.tex}
\input{temp/60-classes-and-objects/exercises/3-60-classes-and-objects/question.tex}
\input{temp/60-classes-and-objects/exercises/4-60-classes-and-objects/question.tex}
\input{temp/60-classes-and-objects/exercises/5-60-classes-and-objects/question.tex}
\input{temp/60-classes-and-objects/exercises/6-60-classes-and-objects/question.tex}

