%-*-latex-*-
\textsc{\cpp\
\textnormal{\texttt{std::stack}},
\textnormal{\texttt{std::queue}},
\textnormal{\texttt{std::deque}},
\textnormal{\texttt{std::foward\_list}},
and
\textnormal{\texttt{std::list}} classes}

You can find information on the above \cpp\ STL classes in my CISS350 notes.
You can also use the web to read up more on these classes.
Also, you have already seen the \verb!std::vector! class from CISS245.

In the questions of this assignment, you can use these class.
(You can also write your own dynamic array, singly-linked list,
doubly-linked list, stack, queue, and deque classes.)

Here's an example of some basic methods in the \verb!std::stack! class
template:
{\small
\VerbatimInput[frame=single,fontsize=\footnotesize]{stack.cpp}
}

Run the program and look at the output.
Check the web for more methods for this class.

Here's an example of some basic methods in the \verb!std::queue! class
template:
{\small
\VerbatimInput[frame=single,fontsize=\footnotesize]{queue.cpp}
}

Run the program and look at the output.
Check the web for more methods for this class.

If \verb!s! is a stack and \verb!q! is a queue, then:
{\small
\begin{Verbatim}[frame=single,fontsize=\footnotesize]
s.empty() -- true if s is empty; otherwise false
q.empty() -- true if q is empty; otherwise false

s.size()  -- size of s
q.size()  -- size of q

s.push(x) -- push x onto top of s
q.push(x) -- enqueue x into back of q

s.pop()   -- remove top of s
q.pop()   -- remove front of q

s.top()   -- reference to the top of stack s
q.front() -- reference to the front of queue q
q.back()  -- reference to the end of queue q
\end{Verbatim}
}

\cpp\ \verb!std::list! class is a doubly-linked list class.
Here's an example of some basic methods in the \verb!std::list! class
template:
{\small
\VerbatimInput[frame=single,fontsize=\footnotesize]{list.cpp}
}
Run the program and look at the output.
Check on the web for more methods for this class.
A constant iterator cannot modify the value that it points to.
In other words if \verb$p$ is a constant iterator, you cannot do
\begin{console}[fontsize=\footnotesize]
*p = newvalue;
\end{console}

If list is a \verb!std::list! object, then
{\small
\begin{Verbatim}[frame=single,fontsize=\footnotesize]
list.empty()       -- true if list is empty; otherwise false

list.size()        -- size of list

list.push_front(x) -- insert x at head
list.push_back(x)  -- insert x at tail

list.pop_front()   -- remove front of list
list.pop_back()    -- remove back of list

list.front()       -- reference to the front/head of list
list.back()        -- reference to the end/tail of list

list.clear()       -- delete all nodes from list
list.erase(p)      -- delete node that iterator p points to
\end{Verbatim}
Here \verb!p! is an iterator of the list (see example code), i.e.,
it's an object that acts like a pointer to the list.
}
