\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Bubblesort}

Recall that we have the follow earlier. Suppose you have four integer variables a, b, c, d. You want to sort the values in a, b, c, d into ascending order and put them into a, b, c, d. For instance if we initially have this:
\begin{center}
a = 5 b = 3 c = 6 d = 0
\end{center}
we want to end up with this:
\begin{center}
a = 0 b = 3 c = 5 d = 6
\end{center}
Recall that I gave you the \EMPHASIZE{bubblesort algorithm} for doing that:

\tab[3em]{
Pass 1:\\
swap the values in a, b so that $a <= b$\\
swap the values in b, c so that b <= c\\
swap the values in c, d so that c <= d\\
// at this point the largest value of a, b, c, d must be in d\\\\

Pass 2:\\
swap the values in a, b so that a <= b\\
swap the values in b, c so that b <= c\\
// at this point the largest value of a, b, c must be in c\\\\

Pass 3:\\
swap the values in a, b so that a <= b\\
// at this point the largest value of a, b must be in b\\
}

[Don't recall any of the above? Quickly review your previous notes on bubblesort!!! Now!!!]

Of course you know what's going to happen next.

We're going to sort values in an array!!! (cheers ... clapping ... and wild whistling).

Let's do this slowly. Using the above pseudocode but with a, b, c, d replaced by a[0], a[1], a[2], a[3], we have the following:

\tab[3em]{Pass 1:\\
swap the values in a[0], a[1] so that a[0] <= a[1]\\
swap the values in a[1], a[2] so that a[1] <= a[2]\\
swap the values in a[2], a[3] so that a[2] <= a[3]\\
// at this point the largest value of a[0], ..., a[3] must be in a[3]\\\\

Pass 2:\\
swap the values in a[0], a[1] so that a[0] <= a[1]\\
swap the values in a[1], a[2] so that a[1] <= a[2]\\
// at this point the largest value of a[0], ..., a[2] must be in a[2]\\\\

Pass 3:\\
swap the values in a[0], a[1] so that a[0] <= a[1]\\
// at this point the largest value of a[0], a[1] must be in a[1]\\
}

We can rewrite Pass 1 as a for-loop:

\tab[3em]{Pass 1:\\for i = 0, 1, 2:\\\tab[3em]{swap the values in a[i], a[i+1] so that a[i] <= a[i+1]\\}
}

See it? Pass 2 looks like this:

\tab[3em]{
Pass 2:\\
for i = 0, 1:\\
\tab[3em]{swap the values in a[i], a[i+1] so that a[i] <= a[i+1]}\\
}

Hmmm ... Pass 2 looks like Pass 1 ... You can combine them like this:

\begin{tabular}{l l}
Pass 1 and 2:\\
for j = 2, 1:\\
& for i = 0, ..., j:\\
& \tab[3em]{swap the values in a[i], a[i+1] so that a[i] <= a[i+1]}\\
\end{tabular}

See that? What about Pass 3? Pass 3 looks like this (I'm writing this as a for-loop just to see the pattern):

\begin{tabular}{l}
Pass 3:
for i = 0:
swap the values in a[i], a[i+1] so that a[i] <= a[i+1]
\end{tabular}

AHA! This can also be combined with Pass 1 and Pass 2:

\begin{tabular}{l}
Pass 1 and 2 and 3:
for j = 2, 1, 0:
for i = 0, ..., j:
swap the values in a[i], a[i+1] so that a[i] <= a[i+1]
\end{tabular}

Note that this is for an array of size 4. What about an array of size 5?
Do you see that the first value of j in the above (i.e. 2) comes from 4
-- 2. In general if you have an array with size SIZE, the bubblesort to
sort in ascending order is this:

\begin{tabular}{l}
for j = SIZE -- 2, ..., 1, 0:\\
\tab[3em]{for i = 0, ..., j:}\\
\tab[6em]{swap the values in a[i], a[i+1] so that a[i] <= a[i+1]}\\
\end{tabular}

Now we translate this to C++:

\begin{consolethree}[escapeinside=||]
for (int j = SIZE - 2; j >= 0; --j)
{
    for (int i = 0; i <= j; ++i)
    {
        if (a[i] > a[i+1])
        {
            int t = a[i];
            a[i] = a[i+1];
            a[i+1] = t;
        }
    }
}
\end{consolethree}

Make sure you understand and memorize this!!!

<<<<<<< HEAD
\input{../exercises/18-17-one-dimensional-arrays/question.tex}
\input{../exercises/19-17-one-dimensional-arrays/question.tex}
\input{../exercises/20-17-one-dimensional-arrays/question.tex}
=======
\begin{ex}
Using the above code, create an array of 100 random
integers from 0 to 1000. Sort the array in ascending order and print all
the integers. Check visually that the output does give you integers in
ascending order.
\end{ex}

\begin{ex}
Redo the previous exercise, sorting the array in
\texttt{\textbf{descending}} order.
\end{ex}

\begin{ex}
Note that the bubblesort algorithm (ascending or
descending version) or course works for an array of doubles as well.
Create an array of 100 random doubles from 0.0 to 1.0, sort it in
ascending order and print the values of the array so that you can
visually check that the values in the array is in ascending order.
\end{ex}

>>>>>>> parent of 3043c95 (.)
