\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Exercises}

<<<<<<< HEAD
\input{../exercises/21-17-one-dimensional-arrays/question.tex}
\input{../exercises/22-17-one-dimensional-arrays/question.tex}
\input{../exercises/23-17-one-dimensional-arrays/question.tex}
\input{../exercises/24-17-one-dimensional-arrays/question.tex}
\input{../exercises/25-17-one-dimensional-arrays/question.tex}
\input{../exercises/26-17-one-dimensional-arrays/question.tex}
=======
\begin{ex}
Declare an array of 1000 integers. Set a[0] to 0 and a[1] to 1. For the remain values in the array, set each value to the sum of the previous two values. For instance set a[2] to the sum of a[0] and a[1]. Set a[3] to the sum of a[1] and a[2]. Etc. In general set a[i] to the sum of a[i-1] and a[i-2] for appropriate values of a[i]. In other words, the value at an index position is the sum of the values at two previous positions. Print all the values in your array. Visually verify the correctness of the values. The first values are of course
\begin{center}

0, 1, 1, 2, 3, 5, 8, 13, 21, \ldots

\end{center}

This is the famous Fibonacci sequence and appears in many areas of science including math, computer science, physics, etc.
\end{ex}

\begin{ex}
Declare an array of 100 random doubles between 0.0 and 1.0. Now set all the elements of the array with \texttt{\textbf{even}} index values to 0.0. Print all the values in your array. Visually verify your work.
\end{ex}

\begin{ex}
Declare an array of 20 characters. Initialize all the value of the array to the space character. In a while-loop continually print the characters separated by spaces, prompt the user for an index position and a character and set the value in the array to the character specified by the user. If the user enters -1 for the index, the program ends the while-loop and print the characters. Here is an execution:

\begin{consolethree}[escapeinside=||]
index: |\textbf{0}|
character: |\textbf{h}|

h
index: |\textbf{1}|
character: |\textbf{e}|

h e
index: |\textbf{2}|
character: |\textbf{l}|

h e l
index: |\textbf{3}|
character: |\textbf{l}|

h e l l
index: |\textbf{4}|
character: |\textbf{o}|

h e l l o
index: |\textbf{6}|
character: |\textbf{w}|

h e l l o   w
index: |\textbf{7}|
character: |\textbf{o}|

h e l l o   w o
index: |\textbf{8}|
character: |\textbf{r}|

h e l l o   w o r
index: |\textbf{9}|
character: |\textbf{l}|

h e l l o   w o r l
index: |\textbf{10}|
character: |\textbf{d}|

h e l l o   w o r l d
index: |\textbf{11}|
character: |\textbf{!}|

h e l l o   w o r l d !
index: |\textbf{-1}|
final: |\textbf{hello world!}|
\end{consolethree}
\end{ex}

\begin{ex}
Write a program that generates 10 random integers between 0 and 9 (inclusive), puts these values into an array, prints the integers in the array, counts the number of even integers in the array using a for-loop, and prints the number of even integers in the array.
\end{ex}

\begin{ex}
Write a program that generates 10 random integers between 0 and 9 (inclusive), puts these values into an array, prints the integers in the array, counts the number of even integers in the array from the first index to either the last index or to the first 9 in the array, and prints the number of even integers (up to when you stop) in the array.
\end{ex}

\begin{ex}
For a tic-tac-toe game, one could use an array of 9 characters to denote the characters on the tic-tac-toe board. In other words, if we create this variable:

\begin{consolethree}[escapeinside=||]
char board[9];
\end{consolethree}

then \texttt{board[0]} is the character at the top left corner of the tic-tac-toe board, ..., and \texttt{board[8]} is the bottom right corner of the board. Rewrite our tic-tac-toe game using this variable. In fact with arrays you can write your game for any n-by-n board:

\begin{consolethree}[escapeinside=||]
const int SIZE = 5;
char board[SIZE * SIZE];
\end{consolethree}

Here's a hint. The code for drawing the tic-tac-toe board would look something like this:

\begin{consolethree}
for (int i = 0; i < SIZE * SIZE; ++i)
{
    if (i % SIZE < SIZE - 1)
    {
        // print board[i] and '|'
    }
    else
    {
        // print board[i], newline, the row divider line
    }
}
\end{consolethree}
\end{ex}

>>>>>>> parent of 3043c95 (.)
