\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Declaring arrays}

We have seen many types: \texttt{int}, \texttt{double} (and \texttt{float}), \texttt{bool}, and \texttt{char}. These are called basic (or primitive) types. There are others.

Now we want to talk about creating a variable which is made up of a \EMPHASIZE{collection} of \EMPHASIZE{values} of the \EMPHASIZE{same type}. In other words such a variable contains more than one value.

Try this:
\xsidebox[-6cm]{0cm}{}{arr1box}{The \EMPHASIZE{size} of \texttt{a} is 3}
\begin{consolethree}[escapeinside=||]
int |\tikzmarknode{arr0anchor}{{\EMPHASIZE{a}}}\EMPHASIZE{[3]};||\tikzmarknode{arr1anchor}{}|
       
a[0] = 4;

a[1] = 2;

a[2] = 42;

|\xsidebox[-4cm]{3cm}{}{arr0box}{\texttt{a} is an \EMPHASIZE{array} of 3 \EMPHASIZE{integers}}|
std::cout << |\tikzmarknode{arr2anchor}{a}|[0] << std::endl;|\xsidebox[-2cm]{0cm}{}{arr2box}{the \textbf{index} of this value in array \texttt{a} is 0. If the size of \texttt{a} is 3, the possible index values are 0, 1 and 2}
|
int x = a[1];
std::cout << x << std::endl;  
\end{consolethree}

\DrawArrow{arr0box}{arr0anchor}
\DrawArrow{arr1box}{arr1anchor}
\DrawArrow{arr2box}{arr2anchor}
\newpage
This is how you visualize an array. The following is a picture describing the memory of \texttt{main()}. Compare \texttt{a} to \texttt{x} in the picture.
\begin{python}
from latextool_basic import *
p = Plot()
p += Rect(0, 0, 13, 6, linewidth=0.05)
p += Rect(1, 3, 4, 4, linewidth=0.05, label='4')
p += Rect(4, 3, 8, 4, linewidth=0.05, label='2')
p += Rect(8, 3, 12, 4, linewidth=0.05, label='42')
p += Rect(0.5, 3.5, 0.5, 3.5, linewidth=0, label='a')
p += Rect(0.5, 2, 0.5, 2, linewidth=0, label='x')
p += Rect(1, 1.5, 4, 2.5, linewidth=0.05, label='4')

p += Rect(8, 7, 12, 8, linewidth=0.05, linecolor='red', label='This is a[0]')
p += Rect(8, 6, 12, 7, linewidth=0.05, linecolor='red', label='This is a[1]')
p += Rect(8, 5, 12, 6, linewidth=0.05, linecolor='red', label='This is a[2]')

p += Line(points=[(8, 7.5), (3, 4)], linecolor='red', linewidth=0.05, endstyle='>')
p += Line(points=[(8, 6.5), (6, 4)], linecolor='red', linewidth=0.05, endstyle='>')
p += Line(points=[(10.5, 5), (10, 4)], linecolor='red', linewidth=0.05, endstyle='>')
print(p)
\end{python}
So this is how you declare an array:
\begin{center}
\EMPHASIZE{[type] [array name][ [size] ];}
\end{center}
As a programmer, you can think of \texttt{a[0]}, \texttt{a[1]}, \texttt{a[2]} as integer variables inside the array variable \texttt{a}. But really \texttt{a} itself is the variable.

WATCHOUT! There are \EMPHASIZE{two uses of \texttt{[]}}. During the declaration of the array it is used to specify the \EMPHASIZE{size} of the array. When accessing a value of the array, \texttt{[]} is used to specify the \EMPHASIZE{index} of the value. Make sure you see the difference!

Although I'm calling \texttt{a[0]}, \texttt{a[1]}, and \texttt{a[2]} the values into array \texttt{a}, you can \EMPHASIZE{think of them as integer variables}. Therefore you \texttt\textbf{do} know how to use them.

You might say: "Well what's the point? Can't we just use three variables such as \texttt{x}, \texttt{y}, and \texttt{z} instead of \texttt{a[0]}, \texttt{a[1]}, and \texttt{a[2]}???" Don't worry we'll come to that very soon.

\input{../exercises/0-17-one-dimensional-arrays/question.tex}
\input{../exercises/1-17-one-dimensional-arrays/question.tex}
