\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{\texttt{if} statements}

So far you know that C++ can do arithmetic, print integer values, print
string values, accept integer values from the keyboards, create and
update variables.

You should expect more! (Otherwise C++ is pretty dumb.)

For instance you would expect programs to make decisions. Now
wouldn' t it be nice if C++ can do this:
\begin{align*}
&\texttt{if alien is killed:}\\
& \tab[3em]{\texttt{score = score + 10}}\\
\end{align*}
or
\begin{align*}
&\texttt{if my ship' s energy is 0:}\\
& \tab[3em]{\texttt{end the game}}\\
\end{align*}
or
\begin{align*}
&\texttt{if plane' s left engine has failed and }\\
&\texttt{plane' s right engine has failed:}\\
& \tab[3em]{\texttt{activate seat ejection!!!}}\\
\end{align*}
Note that for the first case, if the alien is not killed, you do
\textit{not} want to execute the statement \verb!score = score + 10!.
Otherwise everyone would keeping getting the same score! What kind of
dumb game is that?

In fact C++ \textit{does} understand the \verb!if! statement.

Run this program\\
\begin{console}
int x = 0;
std::cin >> x;
if (x > 0)
       std::cout << "spam!" << std::endl;
std::cout << "eggs" << std::endl;
\end{console}

Run the program several time, entering different values for x. Note that
the statement printing \verb@"spam!"@ is controlled by the \verb!if!.
However the statement that prints \EMPHASIZE{"eggs"} is not.

For readability, the if statement is written like this:\\
\begin{console}[commandchars=\~\@\$]
int x = 0;
std::cin >> x;
~textbf@if (x > 0)$
   ~textbf@std::cout << "spam!" << std::endl;$
std::cout << "eggs" << std::endl;
\end{console}
Big picture: The format of the \texttt{if} statement looks like this
\begin{center}
\EMPHASIZE{\textbf{if (}\textbf{[bool expr]}\textbf{)}}

\EMPHASIZE{[stmt]}
\end{center}
where [\verb!bool expr!] is a \EMPHASIZE{boolean expression} -- this
means that it's an expression that can be evaluated to a
boolean value -- and [\verb!stmt!] is a C++ statement (of course
there's a semi-colon at the end). Since we need boolean
expressions it's a good time now to recall the
following:
\begin{align*}
 & \texttt{!} & \text{"not"}\\
 & \texttt{\&\&} & \text{"and"}\\
 & \texttt{\textbar\textbar{}} & \text{"or"}\\
 & \texttt{==} & \text{"is equal to"}\\
 & \texttt{!=} & \text{"is not equal to"}\\
 & \texttt{\textless{}} & \text{"is less than"}\\
 & \texttt{\textless=} & \text{"is less than or equal to"}\\
 & \texttt{>} & \text{"is greater than"}\\
 & \texttt{>=} & \text{"is greater than or equal to"}\\
\end{align*}
An example of a boolean expression is

\EMPHASIZE{[variable] [bool op] [value]}

where [\verb!variable!] is a variable of int or double or float type, [\verb!bool op!] is one of the above boolean operators and [\verb!value!] is
an int or double or float. Don' t forget that if there is
a type mismatch, C++ will try to type promote one value to match the
type of the other. For instance if \verb!x! is an \verb!int! variable
and the boolean expression is

\tab[3em]{\verb!x < 1.23!}

then C++ will use the \verb!double! of \verb!x! (i.e., \verb!x! is
automatically promoted to a \verb!double!):

\tab[3em]{\verb!double(x) < 1.23!}

<<<<<<< HEAD
\input{../exercises/0-07-if-part1/question.tex}
\input{../exercises/1-07-if-part1/question.tex}
\input{../exercises/2-07-if-part1/question.tex}
\input{../exercises/3-07-if-part1/question.tex}
\input{../exercises/4-07-if-part1/question.tex}
\input{../exercises/5-07-if-part1/question.tex}
\input{../exercises/6-07-if-part1/question.tex}
Now, we're comparing two expressions.

\input{../exercises/7-07-if-part1/question.tex}
\input{../exercises/8-07-if-part1/question.tex}
=======
\begin{ex}
Modify the above program so that \verb~"spam!"~ is printed when the integer entered is 42. Test your program.
\end{ex}

\begin{ex}
Modify the above program so that the program prompts the user for his/her favorite number and print \lq\lq you don' t like 42?!?'' if the integer entered is not 42. Here' s an execution:
\begin{console}[commandchars=\~\!\@]
what' s your favorite number? ~textbf!42@ 
\end{console}
Here' s another:
\begin{console}[commandchars=\~\@\$]
what' s your favorite number? ~textbf@1$
~textbf@you don't like 42?!?$  
\end{console}
\end{ex}

\begin{ex}
Write a program that prompts the user for an integer and prints \lq\lq That' s even. Am I smart?'' if the integer entered is even. Here' s an execution
\begin{console}[commandchars=\~\!\@]
~textbf!100@

That' s even. Am I smart?
\end{console}

Here' s another
\begin{console}
101 
\end{console}
\end{ex}

\begin{ex}
Write a program that prompts the user for his/her age and prints \lq\lq You' re lying! Think you can fool me???'' if the value entered is less than 18. Here' s an execution:
\begin{console}[commandchars=\~\@\$]
Enter your age: ~textbf@16$
You' re lying! Think you can fool me???
\end{console}
Here' s another:

\begin{console}[commandchars=\~\!\@]
Enter your age: textbf!100@
\end{console}
Of course this is another boolean expression:
\begin{center}
\EMPHASIZE{[var1] [bool op] [var2]}
\end{center}
In this case you' re comparing two (integer or double)
variables.
\end{ex}

\begin{ex}
Write a program that prompts the user for two integer values and prints \lq\lq First is bigger'' if the first integer entered is larger than the second. Here' s an execution of the program:
\begin{console}[commandchars=\~\!\@]
~textbf!1 2@
\end{console}
Here' s another

\begin{console}[commandchars=\~\!\@]
~textbf!2 -1@ 
First is bigger
\end{console}

Here's yet another boolean expression:

\begin{center}
\EMPHASIZE{[expr] [bool op] [value]}
\end{center}

where \verb![expr]! is an (integer or double) expression.
\end{ex}

\begin{ex}
Write a program that prompts the user for his/her height (in ft) and weight (in lbs) and prints \lq\lq you will live more than 100 years!!!'' if the product of the height and weight is at least 200.45. (This is not supported by any form of research.) Here' s an execution of the program:
\begin{console}[commandchars=\~\!\@]
~textbf!6.1 180.180@
\end{console}
\begin{console}
you will live more than 100 years!!!!
\end{console}
Here' s another
\begin{console}[commandchars=\~\@\$]
~textbf@3.4 23.4$
\end{console}
\end{ex}

\begin{ex}
Write a program that prompts the user for the year of his/her date of birth and the current year and prints his/her approximate age and then prints \lq\lq you must have watched lots of movies by now'' if the difference of the two values is at least 20.

Here' s an execution:
\begin{console}[commandchars=\~\@\$]
~textbf@1900 2000$
you are about 100 years old
you must have watched lots of movies by now
\end{console}

Here' s another:
\begin{console}[commandchars=\~\!\@]
~textbf!2000 2006@
you are about 6 years old
\end{console}


One last one ... here' s one more boolean expression:
\begin{center}
\EMPHASIZE{[expr1] [bool op] [expr2]}
\end{center}
\end{ex}

Now, we're comparing two expressions.

\begin{ex}
Write a program that prompts the user for the month, day of his/her date of birth and also his/her height (in ft) and weight (in lbs). Print the approximate number of days from January 1 of the year he/she was born and the sum of his/her height and weight. If the first printed number is greater than the second, print \lq\lq According to the theory of relativity, you are going to win the next powerball.''.
\begin{console}[commandchars=\~\!\@]
Enter the month, day of your DOB: ~textbf!1 1@
Enter your height and weight: ~textbf!6 100@
1 106
\end{console}
Here' s another:
\begin{console}[commandchars=\~\!\@]
Enter the month, day of your DOB: ~textbf!12 25@
Enter your height and weight: ~textbf!6 200@
355 206
According to the theory of relativity,
you are going to win the next powerball.
\end{console}
(You may assume the number of days in a month is 30.)
\end{ex}

\begin{ex}
Try this:\\
\begin{console}[commandchars=\~\!\@]
int numHeads = 0;
std::cout << "How many heads do you have? ";
std::cin >> numHeads;
if (numHeads == 2)
   std::cout << "Are you Zaphod?" << std::endl;
\end{console}

And then this:

\begin{console}
int numHeads = 0;
std::cout << "How many heads do you have? ";
std::cin >> numHeads;
if (numHeads != 2)
   std::cout << "You are not Zaphod" << std::endl;
\end{console}

Combine both into one program so that I get the following execution:\\
\begin{console}[commandchars=\~\!\@]
How many heads do you have?~textbf!1@
You are not Zaphod
\end{console}
And here's an execution of the same program:\\
\begin{console}[commandchars=\~\!\@]
How many heads do you have?~textbf!2@
Are you Zaphod?
\end{console}
Next modify the program so that, in addition to the above requirements,
if the number of heads entered is greater than 2, the program prints "No
kidding!". Finally, modify the program so that, in addition to the
above, if the number of heads is 1, the program prints "Let me introduce
you to our surgeons." Test your program by entering 0, 1, 2, 3.

(There' s a better way to do this exercise ...)
\end{ex}

>>>>>>> parent of 3043c95 (.)
