\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Blocks}

What if you want to execute \EMPHASIZE{\textit{several}} statements when a
boolean expression is true? Here's one way:\\
\begin{console}[commandchars=\~\@\#]
int x = 0;
std::cin >> x;
if (x > 0)
   std::cout << "spam!" << std::endl;
if (x > 0)
   std::cout << "ham!" << std::endl;
std::cout << "eggs" << std::endl;
\end{console}

It does work \dots but \dots that's the \EMPHASIZE{BAD}
way of executing two statements when a condition holds.

\EMPHASIZE{\textit{This}} is the right way:
\begin{console}[commandchars=\~\@\#]
int x = 0;
std::cin >> x;
if (x > 0)
~textbf@{#
        std::cout << "spam!" << std::endl;
        std::cout << "ham!" << std::endl;
~textbf@}#
std::cout << "eggs" << std::endl;
\end{console}

The \EMPHASIZE{\{...\}} is called a \EMPHASIZE{block} or \EMPHASIZE{block of
statements. }(You' ve actually seen this already. Look at
your \texttt{main() \dots}) You can have as many statements as you like
in a block, not just two.

\begin{ex}
Here's part of a game that you're writing \dots complete it. If the missile hits the alien, increment the \verb!score! by 100, set \verb!play_explosion! to \verb`true`, increment your \verb!energy! by the alien's energy (i.e. \verb`alien_energy`), and set the alien's energy to 0. Test your code with different inputs.
\begin{console}
int score = 1234;
int energy = 6789;
int alien_energy = 42;
bool play_explosion = false;
bool missile_hit_alien;
std::cin >> missile_hit_alien;


std::cout << "score: " << score << '\n'
          << "energy: " << energy << '\n'
          << "alien_energy: " << alien_energy << '\n'
          << "play_explosion: " << score << '\n';
\end{console}
\end{ex}

