\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Random integers}

Generating a random integer occurs commonly in computer programs. For instance if you want to write a strategy role-playing game, you might have different rooms in a mansion and you want to put different things in different rooms. Suppose you have a magic potion in a bottle. Now suppose the rooms are numbered from 1 to 1000. You don' t want your magic potion bottle to be in room 5 whenever you start the game. You want each game to be different.

At a more complex level you might want to randomly make doors between two rooms. This creates a random maze.

So you' d better learn how to generate random things.

Try this:
\begin{console}[commandchars=\~\@\$]
#include < iostream>

int main()
{
std::cout << "the potion is in room " << ~textbf@rand()$
          << std::endl;
std::cout << "the potion is in room " << ~textbf@rand()$
          << std::endl;
std::cout << "the potion is in room "<< ~textbf@rand()$
          << std::endl;
std::cout << "total number of rooms: "
          << ~textbf@RAND_MAX$ << std::endl;
          
return 0;
} 
\end{console}
(Note: You might need \verb!#include <cstdlib>!\\just after \verb!#include <iostream>!. This depends on your compiler.)

Looks like you' re getting random integers. But wait ...
run the program a second time ... a third time ... a fourth time. Notice
something?

OK. Let' s modify the program:
\begin{console}[commandchars=\~\@\$]
#include <iostream>
~textbf@#include <ctime>$
int main()
{
~textbf@srand((unsigned int) time(NULL));$   
std::cout << "the potion is in room " << rand()
          << std::endl;
std::cout << "the potion is in room " << rand()
          << std::endl;
std::cout << "the potion is in room " << rand()
          << std::endl;
std::cout << "total number of rooms: "
          << RAND_MAX << std::endl;
return 0;
\end{console}

Run this program several times. Notice that the room number is different
each time you run the program.

I will not go into details on the statement
\begin{center}
\texttt{srand((unsigned int) time(NULL));}
\end{center}
The only thing I will say is that it \lq\lq seeds'' the random generator so
as to make it \lq\lq more random''. The important thing to remember is that
you must execute this statement and furthermore you execute it
\textbf{\underline{once}} (usually) and \textbf{\underline{before}} you call the \texttt{rand()} function. The standard practice is to execute the seeding
in \texttt{main()} and at the beginning:
\begin{console}[commandchars=\~\@\$]
#include <iostream>
~textbf@#include <ctime>$
int main()
{
   ~textbf@srand((unsigned int) time(NULL));$
   . . .
   return 0;
}
\end{console}
Since \verb!rand()! gives you a random number from \verb!0! to \verb!RAND_MAX!,
\begin{center}
\texttt{rand() / RAND\_MAX}
\end{center}
will give you a random \verb!double! between \verb!0.0! and \verb!1.0!.
But this is not right because the / is an \textit{\textbf{integer}}
division. Therefore we do this:
\begin{center}
\texttt{double(rand()) / RAND\_MAX}
\end{center}
to get a random double from 0.0 to 1.0.

<<<<<<< HEAD
\input{../exercises/14-07-if-part1/question.tex}
\input{../exercises/15-07-if-part1/question.tex}
\input{../exercises/16-07-if-part1/question.tex}
=======
\begin{ex}
Check that I was not lying by printing 10 random doubles from 1.0 to 0.0. By the way, if the numbers are truly random, you would expect their average to be close to 0.5. Check that too.

There are times when you need to generate a random \textit{\textbf{integer}} between two bounds. For instance if you have a maze game and the rooms are numbered 1 to 100, you want a random room number for a magic potion. How many values are there from 1 to 100? You have 100. Since \texttt{rand()} gives you a random integer from 0 to \texttt{RAND\_MAX}, if you do
\begin{center}
        \texttt{rand() \% 100}
\end{center}
you get 100 numbers, from 0 to 99. You add 1 to it and you get a random
integer from 1 to 100. In other words
\begin{center}
        \texttt{1 + rand() \% 100}
\end{center}

does the trick.
\end{ex}

\begin{ex}
How do you generate random integers from 1,..., 6? (For instance to simulate a dice roll.) Verify with a C++ program.
\end{ex}

\begin{ex}
How do you generate random integers from 1,..., 10? Verify with a C++ program.
\end{ex}

>>>>>>> parent of 3043c95 (.)
What about generating random integers from 5,..., 20? In this case the the lower bound is 5 and not 1. You use the same idea. Note that if you do something like
\begin{center}
\texttt{rand() \% 10}
\end{center}
or
\begin{center}
\texttt{rand() \% 18}
\end{center}
the range of numbers starts from 0. So if you want to generate random
numbers in the range of 5,..., 20, the first thing you do is to think
about the range
\begin{center}
0, \dots, 15
\end{center}
first. In other words move the given range 5, \ldots, 20 so that the
lower bound of the range becomes 0. That means subtracting 5 from the
range. To generate a number in the range of 0, \ldots, 15, you do
\begin{center}
\texttt{rand() \% 16}
\end{center}
This gives a number in the range 0, \ldots, 15. You then add 5 to it:
\begin{center}
\texttt{5 + rand() \% 10}
\end{center}
\textbf{Exercise.} Verify that the expression\\
\begin{center}
\texttt{5 + rand() \% 10}
\end{center}
does give you numbers in the range of 5, \dots, 20. .

<<<<<<< HEAD
\input{../exercises/17-07-if-part1/question.tex}
\input{../exercises/18-07-if-part1/question.tex}
\input{../exercises/19-07-if-part1/question.tex}
\input{../exercises/20-07-if-part1/question.tex}
\input{../exercises/21-07-if-part1/question.tex}
\input{../exercises/22-07-if-part1/question.tex}
=======
\begin{ex}
How do you generate random integers from 9,...142? Verify with a C++ program.
\end{ex}

\begin{ex}
How do you generate random integers from -5,...,5? Verify with a C++ program.
\end{ex}

\begin{ex}
How do you generate random \verb`double` from 0,...,2.0? Verify with a C++ program.
\end{ex}

\begin{ex}
How do you generate random double from -1.0,...,1.0? Verify with a C++ program.
\end{ex}

\begin{ex}
How do you generate random double from 0,...,3.5? Verify with a C++ program.
\end{ex}

\begin{ex}
How do you generate random double from -2.0,...,2.0? Verify with a C++ program.
\end{ex}

>>>>>>> parent of 3043c95 (.)
