\section{Toward Normal Forms}

We want to restrict the form of the productions. The reason is
that we want the top-down and bottom-up parsers  to terminate. If
we have time we'll talk about these but they really belong the to
a course such as compiler construction.

Anyway, the goal is to eliminate bad productions. 
Here are the bad
guys:

\begin{defn}
 \begin{mylist}
  \item[(a)] An \defterm{$\ep$-production} is a production of the form $A
  \rightarrow \ep$ where $A$ is a variable.
  \item[(b)] A \defterm{unit production} is a production of the form $A
  \rightarrow B$ where $A$ and $B$ are variables.
  \item[(c)] A variable $A$ is \defterm{useless} (nice technical term,
  isn't it?) if $A$ is not used in the derivation of any string of
  terminals, i.e., if there is no $x \in T^*$, $\{ \alpha,\beta \}
  \subseteq (V \cup T)^*$ such that
  \[
   S \istar \alpha A \beta \istar x
  \]
 \end{mylist}
\end{defn}

Why are they bad?

\begin{eg}
Consider 
\begin{align*}
S &\rightarrow AB \\
A &\rightarrow \ep \\
B & \rightarrow b
\end{align*}
First of all $A \rightarrow \ep$ (an $\ep$--production) is redundant
since the above CFG generate the same language as
\begin{align*}
S &\rightarrow B \\
B &\rightarrow b
\end{align*}
Correct?
The point is that in the above example,
the $A \rightarrow \ep$ is an $\ep$-production and 
can be removed: $A$ is going to disappear during the derivation process
anyway:
\[
S 
\implies^{\hspace{-0.3cm}*} ... A ... 
\implies^{\hspace{-0.3cm}*} ... \ep ... 
\]
Note however that the rule
\[
S \rightarrow \ep
\]
cannot be be removed: your grammar would lose it's
ability to generate $\ep$!!!
\end{eg}

\begin{eg}
Next consider the grammar
\begin{align*}
S &\rightarrow A \\
A &\rightarrow B \\
B &\rightarrow  Bb \st \ep 
\end{align*}
The unit products $A \rightarrow B$ is redundant.
The above grammar generates the same language as 
\[
S \rightarrow B, \,\,\, B \rightarrow Bb \st \ep
\]
\end{eg}

\begin{eg}
The last example shows you examples of useless symbols:
\begin{align*}
S &\rightarrow \ep \mid aS \mid A \\
B &\rightarrow b
\end{align*}
$A$ is useless because it does not derive terminals; $B$ is useless 
because it cannot be reached from $S$.
\end{eg}

The above are over-simplied examples.
But it will help you in understanding the general situation.







\newpage
\subsection{$\ep$-Productions}

First we will get rid of the $\ep$-productions. First we need to
compute the set of nullable variables:


\begin{defn} A variable $A$ is \defterm{nullable} if $A \istar
\ep$. A more recursive definition: $A$ is nullable if either $A
\rightarrow \ep$ or $A \rightarrow \alpha$, a product of
variables, and each variable in $\alpha$ is nullable.
$\NULLABLES$ is the set of all nullables of the grammar.
\end{defn}

The algorithm to compute the set of nullables is easy:
\begin{mylist}
 \item[(1)] Let $\NULLABLES = \{V \,|\, V\rightarrow \ep\}$
 ($\ep$-productions)
 \item[(2)] Repeat
 \begin{mylist}
  \item[(2.1)] if $V \notin \NULLABLES$, $V \rightarrow
  \alpha$ is a production, $\alpha$
  is a product of variables in $\NULLABLES$,
  then add $V$ to $\NULLABLES$.
 \end{mylist}
 \item[(3)] Until $\NULLABLES$ stops changing
\end{mylist}

Here's the algorithm to get rid of $\ep$-productions.
Let $G$ be the grammar.
\begin{mylist}
 \item[(1)] Compute $\NULLABLES$
 \item[(2)] For each production $A \rightarrow X_1 \ldots X_n$
 where $X_i \in V \cup T$, create all possible productions $A \rightarrow
 \alpha_1 \cdots \alpha_n$ where
 \[
 \alpha_i =
 \begin{cases}
   X_i & \text{ if $X_i \notin \NULLABLES$} \\
   X_i \text{ or } \ep & \text{ if $X_i \in \NULLABLES$}
 \end{cases}
 \]
 \item[(3)] Eliminate all $V \rightarrow \ep$ productions.
\end{mylist}
Let $G'$ be the new grammar.
The resulting grammar generates the same language except that 
if $S \rightarrow \ep$ was removed, then the new grammar
will not generate $\ep$.
In other words if $S \rightarrow \ep$ was removed, then
\[
L(G') = L(G) - \{\ep\}
\]
Otherwise
\[
L(G') = L(G)
\]

What do you do if $G$ does generate $\ep$?
You perform the above operation to get $G'$ and to $G'$ you add the
only exception $S \rightarrow \ep$.
This would be the only $\ep$--production.
In this case, $G'$ is in Chomsky normal form except for
$S \rightarrow \ep$.

In general when 
an author says that $G$ does not have $\ep$ production rules,
he/she meant that the grammar does not have $\ep$ production rules
\textit{except possibly for $S \rightarrow \ep$}.
You just have to look at the context very carefully.


\newpage
\begin{eg}
Let $G$ be the following context-free grammar:
\begin{align*}
  S &\rightarrow ACA \\
  A &\rightarrow aAa \,|\, B \,|\, C \\
  B &\rightarrow bB \,|\, b \\
  C &\rightarrow cC \,|\, \ep 
\end{align*}
\begin{tightlist}
\item What is the $\NULLABLES$ for the grammar
\item Eliminate the $\ep$-productions.
\end{tightlist}
\end{eg}

\textbf{Solution.}
First we compute the $\NULLABLES$.
\begin{tightlist}
\item First of all the only production rule directly of the form 
$X \rightarrow \ep$ is 
$C \rightarrow \ep$.
So we have
\[
\NULLABLES = \{V \mid V\rightarrow \ep \} = \{C\}
\]
\item Next, we look for new production rules of the form $X \rightarrow C$.
The only one is $A \rightarrow C$.
So the $\NULLABLES$ at this point is
\[
\NULLABLES = \{C, A\}
\]
\item Repeating the process again, we look for new production rules of the
form $X \rightarrow w$ where $w$ is a product of
variables in 
$\NULLABLES = \{C, A\}$. The only new rule is $S \rightarrow ACA$.
Therefore at this point we have
\[
\NULLABLES = \{C, A, S\}
\]
\item The process stops.
\end{tightlist}

Second, I will perform substitutions on the nullable variables. 
\begin{enumerate}
\item First I'm going to handle $C$.
  By performing the substitution of $C$ by $\ep$ to get
  new rules, the grammar
  \begin{align*}
    S &\rightarrow ACA \\
    A &\rightarrow aAa \mid B \mid C \\
    B &\rightarrow bB \mid b \\
    C &\rightarrow cC \mid \ep
  \end{align*}
  becomes
  \begin{align*}
    S &\rightarrow ACA | \underline{AA} \\
    A &\rightarrow aAa \mid B \mid C \mid \underline{\ep} \\
    B &\rightarrow bB \mid b \\
    C &\rightarrow cC \mid \underline{c} \mid \ep
  \end{align*}
  For readability, I've underlined the right-hand side of the new rules.
  For instance, the original rule 
  \[
  S \rightarrow ACA
  \] 
  gives
  me the new rule 
  \[
  S \rightarrow A\ep A = AA
  \]
  The reason is 
  that in any derivation, 
  $C$ can be replaced by $\ep$ because of the rule $C \rightarrow \ep$.
  Right?
The whole point of doing this is that later, I want to remove the rule
$C \rightarrow \ep$ from the grammar.
Note that I cannot throw away the rule original 
$S \rightarrow ACA$; it must be 
included. 
Why? 
Because a derivation that uses $S \rightarrow ACA$ need not follow up
with the rule $C \rightarrow \ep$. Right?
Oh by the way, notice that at this stage, we inherit the $\ep$-product rule
$A \rightarrow \ep$. Think about that: it should be pretty obvious why this
is the case.

\item Now I'm going to do the same for $A$.
  The grammar becomes:
  \begin{align*}
    S &\rightarrow ACA 
    \mid \underline{CA}
    \mid \underline{AC} 
    \mid \underline{C} 
    \mid AA 
    \mid \underline{A}
    \mid \underline{\ep}
    \\
    A &\rightarrow aAa \mid \underline{aa} \mid B \mid C \mid \ep \\
    B &\rightarrow bB \mid b \\
    C &\rightarrow cC \mid c \mid \ep
  \end{align*}
For instance the rule $S \rightarrow ACA$ gives rise to the following
new rules:
\[
S \rightarrow \ep CA \mid AC \ep \mid \ep C \ep
\]
i.e.,
\[
S \rightarrow CA \mid AC \mid C
\]

\item Finally, I do the same for $S$. However in this case,
since $S$ does not appear on the right-hand side of any 
rule, there are no new rules.
\end{enumerate}

At this point, the grammar is 
\begin{align*}
  S &\rightarrow ACA 
    \mid CA
    \mid AC 
    \mid C 
    \mid AA 
    \mid A
    \mid \ep
    \\
    A &\rightarrow aAa \mid aa \mid B \mid C \mid \ep \\
    B &\rightarrow bB \mid b \\
    C &\rightarrow cC \mid c \mid \ep
  \end{align*}

The last stage is to remove the $\ep$--production rules.
I get
\begin{align*}
  S &\rightarrow ACA 
    \mid CA
    \mid AC 
    \mid C 
    \mid AA 
    \mid A
    \\
    A &\rightarrow aAa \mid aa \mid B \mid C \\
    B &\rightarrow bB \mid b \\
    C &\rightarrow cC \mid c 
  \end{align*}
Note that I removed $S \rightarrow \ep$.
The resulting grammar does not have any $\ep$-rule but it does generate
$\ep$ ($S$ is in $\NULLABLES$).
This means that the resulting grammar
generates the same language except that it's missing the 
single word $\ep$.
So I have to make the exception for the rule $S \rightarrow \ep$.
Adding $S \rightarrow \ep$ back, I get a grammar that does not have
any $\ep$-production other than the rule $S \rightarrow \ep$:
\begin{align*}
  S &\rightarrow ACA 
    \mid CA
    \mid AC 
    \mid C 
    \mid AA 
    \mid A \mid \ep
    \\
    A &\rightarrow aAa \mid aa \mid B \mid C \\
    B &\rightarrow bB \mid b \\
    C &\rightarrow cC \mid c 
  \end{align*}

\newpage
\begin{eg} Remove $\ep$--production rules from the following grammar:
  \begin{align*}
    S &\rightarrow ABC \\
    A &\rightarrow aA \,|\, \ep \\
    B &\rightarrow bB \,|\, \ep \\
    C &\rightarrow cC \,|\, \ep \\
  \end{align*}

\end{eg}




\newpage
\subsection{Unit Productions}

The next thing to do is to get rid of unit productions. We will
assume that $L(G) \neq \emptyset$. Furthermore we will assume that
$G$ has no $\ep$-productions. The main idea, like the removal of
$\ep$-productions, is very simple. Suppose you do have a unit
production $A \rightarrow B$. Furthermore suppose $B \rightarrow
b$. Then of course $A$ can derive $b$. So if we include $A
\rightarrow b$ as a production, we don't really need the unit
production $A \rightarrow B$. Of course you need to look at all
possible derivations from $A$ to a terminal string that uses $B$.
For instance there might be another production $B \rightarrow b'$.
So here's the algorithm:

\begin{mylist}
 \item[(1)] For every pair of variables $A$ and $B$, if
 $A \istar B$:
 \begin{mylist}
  \item[(1.1)] Add productions $A \rightarrow \beta_1 \,|\, \beta_2 \,|\,
  \ldots \,|\, \beta_n$ where $B \rightarrow \beta_1 \,|\, \beta_2 \,|\,
  \ldots \,|\, \beta_n$ are all the nonunit productions for $B$
  ($\beta_i$ are sentential forms).
 \end{mylist}
 \item[(2)] Eliminate all unit productions.
\end{mylist}

Note that the above algorithm requires verification of $A \istar
B$. I'll show you how to do that later. First let's prove the
following:

\begin{thm}
Suppose $G$ is the original CFG and $G'$ is the CFG obtained after
removal of $\ep$-productions and unit productions. Then $L(G) =
L(G')$.
\end{thm}

\begin{proof}
The fact that $L(G') \subseteq L(G)$ is clear: If $x \in L(G')$
and $x$ is derived without using the new productions, then of
course $x \in L(G)$. If $x$ is derived using an new production,
say $A \rightarrow b$, then by definition of $G'$, we know that $A
\istar_G B$ (derivation in $G$) and $B \rightarrow b$ is a
production in $G$. Hence $A \istar_G b$ (derivation in $G$). In
other words $x$ can be derived using $G$, i.e., $x \in L(G)$.

Now we prove that $L(G) \subseteq L(G')$. Let $x \in L(G)$. The
idea is very simple. Let's look at a leftmost derivation of $x$
which is derived using $G$ and that it involves at least one unit
production:
\[
 S \istar w A \alpha \rightarrow wB \alpha \rightarrow wC \alpha \istar
 wD\alpha \implies \ldots \implies x
 \rightarrow w\gamma\alpha
\]
where $A, B, C, D$ are variables, $w,w'$ are all terminal strings,
and $\alpha, \gamma$ are sentential forms (i.e., strings involving
variables and terminals); note that we see the unit productions $A
\rightarrow B$, $B \rightarrow C$, etc. AHA! This tells us that $A
\istar D \rightarrow \gamma$ which by our definition of $G'$
implies that $A \rightarrow \gamma$ is in fact in $G'$!!! In other
words the following derivation
\[
 S \istar w A \alpha \implies wD\alpha \rightarrow w\gamma
 \alpha \implies \ldots \implies x
\]
has some unit productions removed and it uses a production
\textit{ in
$G'$.} Using this idea we can slowly change the original
derivation using $G$ to one using $G'$: Just apply the above idea
to the first time a unit production appears in the derivation and
repeat this process to the next unit product, etc. Since a
derivation is finite in length, obviously the process terminates
and we have a derivation \textit{ in $G$}. Voila.

\end{proof}

There's still the problem of verifying whether $A \istar B$. The
idea is also simple. Suppose $G$ does not have any
$\ep$-productions. For $A \istar B$ to occur, the derivation must
involve only unit productions:
\[
 A \implies X_1 \implies X_2 \implies \cdots \cdots \implies X_n
 \implies B
\]
(Why?). So all you need to do is to make a (directed) graph where
the nodes are all the variables of the grammar and $X \rightarrow
Y$ is an edge in the graph if $X \rightarrow Y$ is a production in
the grammar. Then $A \istar B$ iff there is a directed path in the
graph from node $A$ and node $B$. To check if there is a path from
node $A$ to node $B$, do a depth-first search. That's it!!!

\newpage
\begin{eg}
Consider the grammar
\begin{align*}
S &\rightarrow ACA \mid CA \mid AA \mid AC \mid A \mid C \mid \ep \\
A &\rightarrow aAa \mid aa \mid B \mid C \\
B &\rightarrow bB \mid b \\
C &\rightarrow cC \mid c
\end{align*}
Other than $S \rightarrow \ep$, this grammar does not have any 
$\ep$--productions.
\end{eg}

First we need to know when a symbol can derive another.
The unit production rules are
\begin{align*}
S &\rightarrow A \\
S &\rightarrow C \\
A &\rightarrow B \\
A &\rightarrow C \\
\end{align*}
and here's the directed graph of variables taking part in unit productions: 
\begin{python}
from latextool_basic import *
print(automata(layout=r'''
   A  B
S  
   C
''',
S = 'label=$S$',
A = 'label=$A$',
B = 'label=$B$',
C = 'label=$C$',
edges='S,,A|S,,C|A,,B|A,,C'))
\end{python}
The relevant derivations are
\begin{align*}
S \istar A \\
S \istar B \\
S \istar C \\
A \istar B \\
A \istar C
\end{align*}
So here are the productions we will add
\begin{center}
\begin{tabular}{|l|l|} \hline
Unit derivation & Production to add \\ \hline
$S \istar A$    & $S \rightarrow aAa \mid aa$ \\ \hline
$S \istar B$    & $S \rightarrow bB \mid b$ \\ \hline
$S \istar C$    & $S \rightarrow cC \mid c$ \\ \hline
$A \istar B$    & $A \rightarrow bB \mid b$ \\ \hline
$A \istar C$    & $A \rightarrow cC \mid c$ \\ \hline
\end{tabular}
\end{center}
Hence we now have the following productions:
\begin{align*}
S &\rightarrow ACA \mid CA \mid AA \mid AC \mid A \mid C \mid \ep \mid aAa \mid aa \mid
  bB \mid b \mid cC \mid c \\
A &\rightarrow aAa \mid aa \mid B \mid C \mid bB \mid b \mid cC \mid c \\
B &\rightarrow bB \mid b \\
C &\rightarrow cC \mid c  
\end{align*}

And now we remove the unit productions to get
\begin{align*}
S &\rightarrow ACA \mid CA \mid AA \mid AC \mid \ep \mid aAa \mid aa \mid bB \mid b 
   \mid cC \mid c \\
A &\rightarrow aAa \mid aa \mid bB \mid b \mid cC \mid c \\
B &\rightarrow bB \mid b \\
C &\rightarrow cC \mid c
\end{align*}
This grammar does not have $\ep$--productions (other than $S \rightarrow \ep$)
or unit productions.

You note that in general, just like the process for $\ep$--productions, 
more productions are added to the grammar.

\newpage
\subsection{Useless Symbols}

Finally we need to get rid of the useless symbols. Recall that a
symbol $A$ is useless if it is not used in any derivation.
Formally, $A$ is useless if there does not exist $\alpha$
(sentential form), $\beta$ (sentential form), $x$ (terminal
string) such that
\[
 S \istar \alpha A \beta \istar x
\]
The main idea is simple: Symbols which are useful are those which
can derive a terminal string and also, beginning with the starting
symbol, you can reach a sentential form containing that symbol.
Formally, $A$ is useful if
\begin{mylist}
 \item[(a)] $A \istar x$ for some terminal string $x$
 \item[(b)] $S \istar \alpha A \beta$ for sentential forms
 $\alpha, \beta$.
\end{mylist}

We will first focus on (a). Again the algorithm is recursive: We
start with symbols $A$ that can derive terminals in one step, then
we consider productions $A \rightarrow \alpha$ where $\alpha$ is a
string of terminals or variables in our set. We continue until the
set of symbols does not change. Here's the algorithm:
\begin{mylist}
 \item[(1)] Let $\TERMINAL$ be the set of variables $A$ such that
 there is a production $A \rightarrow x$ where $x$ is a terminal
 string
 \item[(2)]  Repeat
 \begin{mylist}
  \item[(2.1)] For each product $A \rightarrow \beta$ where
  $\beta$ is a string of terminals or symbols in $\TERMINAL$, put $A$
  into $\TERMINAL$.
 \end{mylist}
 \item[(3)] Until $\TERMINAL$  stops growing
 \item[(4)] Construct the new grammar with productions that use only
 $\TERMINAL \cup \Sigma$. The variable set of the new grammar is
 $\TERMINAL$. The new $\Sigma$ is of course the set of terminal
 symbols that appear in the new grammar.
\end{mylist}

It's clear that at the end of executing this algorithm, $\TERMINAL$
contains variables $A$ such that $A \istar x$ for some terminal
string $x$.

Now we focus on (b) from the above, i.e., we want to find
variables $A$ such that $S \istar \alpha A \beta$. Such variables
are said to be reachable. The idea is again recursive. We
continually build a set of variables and set of terminals.

\begin{mylist}
 \item[(1)] Set $V' = \{S\}$ and $T' = \emptyset$
 \item[(2)] Repeat
 \begin{mylist}
  \item[(2.1)] If $A \in V'$ and $A \rightarrow \alpha_1\,|\,
  \ldots \,|\, \alpha_n$ in $G$, then
   add the variables of
   $\alpha_1$, \ldots, $\alpha_n$ to $V'$ and add terminals of
   $\alpha_1, \ldots, \alpha_n$ to $T'$
 \end{mylist}
 \item[(3)] Until $V'$ and $T'$ stop changing
 \item[(4)] Let $P'$ be the set of productions using only $V'$
 and $T'$.
\end{mylist}

\begin{thm}
If $G$ is a CFG and $L(G) \neq \emptyset$, then we can find a CFG
$G''$ such that $L(G'') = L(G)$ such that $G''$ has no useless
symbols. In particular, $G''$ is obtained from $G$ by applying the
above algorithms (a) and (b).
\end{thm}

Combining everything above we get this:

\begin{thm}
Given a CFG $G$ such that $L(G) \neq \emptyset$ we can find a CFG
$G'$ such that $L(G') = L(G) - \{\ep\}$ and $G'$ has no
$\ep$-productions, no unit productions, and no useless symbols.
\end{thm}

The new $G'$ is of course derived from $G$ by removing
$\ep$-productions, unit productions, and useless symbols using the
above algorithms.

There are two standard forms of CFG:
\begin{mylist}
 \item Chomsky Normal Form: Every production is either $A
 \rightarrow BC$ or $A \rightarrow a$ where $A,B,C$ are variables
 and $a$ is a terminal symbol
 \item Greibach Normal Form: Every production is of the form $A
 \rightarrow b\alpha$ where $A$ is a variable, $b$ is a terminal
 symbol, and $\alpha$ is a string of variables.
\end{mylist}
