\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Copy constructor}

Suppose you already have a \texttt{Date} object \texttt{date1}. You want to
create another \texttt{Date} object, \texttt{date2}, so that \texttt{date2}
has the same values as \texttt{date1}.

\texttt{date2} is not a reference to date1 like this:

\begin{consolethree}[escapeinside=||]
Date & date2 = date1;
\end{consolethree}

Rather, you want \texttt{date2} to be an actual Date object with
its own member variables.

You can do this

\begin{consolethree}[escapeinside=||]
Date date2(date1.get_y(), date1.get_m(), date1.get_d());
\end{consolethree}

using the constructor that accepts three integers. But ...

What a pain!

This is cleaner:

\begin{consolethree}[escapeinside=||]
Date date1;
Date date2;
date2 = date1;
\end{consolethree}

But this means that \texttt{date1} was first initialized to something that
I do not want, and \texttt{then} I assign the values of \texttt{date1} to
\texttt{date2}. That's not efficient.

We can achieve a cleaner initialization if we define a constructor that
accepts a \texttt{Date} object. In other words I would like to do this:

\begin{consolethree}[escapeinside=||]
Date date2(date1);
\end{consolethree}

The prototype looks like this:

\begin{consolethree}[escapeinside=||]
Date(const Date &);
\end{consolethree}

Note that I'm passing by reference to make the parameter passing
efficient and since I don't intend to modify the parameter, I make it a
constant. I'm therefore passing by constant reference.

The implementation looks like this:

\begin{consolethree}[escapeinside=||]
// Date.cpp
Date::Date(const Date & date)
    : yyyy_(date.yyyy_), mm_(date.mm_), dd_(date.dd_)
{}
\end{consolethree}

You can think of \texttt{date2} as a clone of \texttt{date1}. You can (and
should) think of the copy constructor as a cloning operation.

For any class \texttt{C}, the \EMPHASIZE{copy constructor} of that class is a
constructor with the following prototype:

\texttt{C(const C \&);}

Think of the copy constructor as a \EMPHASIZE{cloning operation} and the
intent is to copy values from one object to another.

<<<<<<< HEAD
\input{../exercises/19-61-constructors/question.tex}
=======
\begin{ex}
Inline our \texttt{Date} copy constructor.
\end{ex}

>>>>>>> parent of 3043c95 (.)

