\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Default copy constructor}

If a copy constructor is not specified, your C++ compiler will supply a
default copy constructor that initializes all instance variables of the
object invoking the constructor with the corresponding members of the
object that is passed in.

\begin{consolethree}[escapeinside=||]
// C has instance variables a_, b_, c_, d_
C::C(const C & obj)
    : a_(obj.a_), b_(obj.b_), c_(obj.c_), d_(obj.d_)
{}
\end{consolethree}

So technically we don't need the copy constructor for our \texttt{Date}
class. However, you will see very soon, that there are cases (many
cases!!!) where you want your own copy constructor to do something else.
So here's the warning just so you catch it ...

\EMPHASIZE{WARNING:} There are cases where the default member-by-member
copy is \EMPHASIZE{\texttt{not}} what you want. (See later examples.)

If you define a copy constructor, of course the compiler will not supply
one. You can't possibly have two copy constructors. (Why?)

Comment out our copy constructor in \texttt{Date} and then execute in
\texttt{main()}:

\begin{consolethree}[escapeinside=||]
Date date1(2050, 1, 1);
Date date2(date1);  // clone date2 from date1
Date date3 = date1; // clone date3 from date1
\end{consolethree}

You will find that the compiler will not complain and the program does
run. Why? Because your C++ compiler supplies the default copy
constructor.

\input{../exercises/20-61-constructors/question.tex}
\input{../exercises/21-61-constructors/question.tex}
\input{../exercises/22-61-constructors/question.tex}
\input{../exercises/23-61-constructors/question.tex}
\input{../exercises/24-61-constructors/question.tex}

