\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Review: Function Overloading}

A function is \EMPHASIZE{overloaded} if its name is used more than once.

The signature of the functions with the same name must be different.
That's because in C++, a function is identified not by
its name, but by it name and its signature, i.e., the name and the list
of types of its parameters.

\input{../exercises/8-61-constructors/question.tex}
\input{../exercises/9-61-constructors/question.tex}
\input{../exercises/10-61-constructors/question.tex}
When a function is called in your code, C++ will use the function that
matches both the function name and the type of the values passed in with
the candidate's signature. If none is found, C++ will try to typecast
the arguments. The one used is the one with the least number of
typecasts. If there isn't one, you get an ambiguous invocation error.

\input{../exercises/11-61-constructors/question.tex}
\input{../exercises/12-61-constructors/question.tex}
If you compile the following with g++:

\begin{consolethree}[escapeinside=||]
void f(double x, int y) {}
void f(int x, double y) {}
int main() { f(1, 1); }
\end{consolethree}

it will complain with this message:

\begin{consolethree}[escapeinside=||]
a.cpp: In function int main():
a.cpp:3:20: error: call of overloaded f(int, int) is
    ambiguous
a.cpp:1:6: note: candidates are: void f(double, int)
a.cpp:2:6: note: void f(int, double)
\end{consolethree}

Basically, when looking for a function to execute

\texttt{f(1, 1);}

there is no exact match. So C++ attempts to typecast. If C++ typecasts
the first value to 1.0:

\verb!f(double(1), 1);!

it will be able to match this:

\verb!void f(double x, int y) {}!

and if C++ typecasts the second value to get 1.0:

\verb!f(1, double(1));!

it will match this:

\verb!void f(int x, double y) {}!

So, in this case, since both functions match \texttt{f(1, 1)} with 1
typecast, your C++ compiler does not know what to do and will complain.
Of course if there's no exact match and
there's only one option with one typecast, then C++ will
happily choose that function that matches\texttt{f(1, 1)} with one typecast.


