%-*-latex-*-
[Bubblesort]

Implement the standard bubblesort algorithm in C\texttt{++}.
The following are the prototypes:
\begin{console}[frame=single, fontsize=\footnotesize]
template < typename T >
void bubblesort(T * start, T * end, bool verbose=false);

template < typename T, typename C >
void bubblesort(T * start, T * end, const C & less, bool verbose=false);
\end{console}
The \verb!verbose! parameter, when \verb!true!, will
display the values at address \verb!start! up to but not including \verb!end!,
and also at the \textit{end} of each pass with the values
separated by \verb!", "! and
a newline is printed after printing each pass.
Actually I will only test your code for the first prototype, where
sorting is done in the default ascending order.
So the second prototype is optional.

Here's an example execution:
\begin{Verbatim}[frame=single,commandchars=\~\!\@,fontsize=\footnotesize]
~userinput!./a.out 42 4 10 1 0@
{1, 9, 4, 2}
{1, 4, 2, 9}
{1, 2, 4, 9}
{1, 2, 4, 9}
\end{Verbatim}
The first output \verb!{1, 9, 4, 2}! is the randomize vector with 4
values (unsorted),
the second output is the vector at the end of the first pass,
the third output is the vector at the end of the second pass, and the
last output is the vector at the end of the third pass.

Make sure your output matches the format \textit{exactly}
or your implementations will be considered incorrect.

I recommend the following steps:
\begin{enumerate}[nosep]
\item Write a bubblesort that accepts pointers to two integers.
\item Add verbose printing.
\item Add functor for specifying sort order.
\item Convert to function template.
\end{enumerate}

