\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Modification of values in an array through a function}

There is something different between the way functions work with an
array and a variable of basic type (\texttt{int} or \texttt{double} or
\texttt{bool} or \texttt{char}). You already know that the following will
not change the value of \texttt{x} in \texttt{main()}:

\begin{consolethree}[escapeinside=||]
void inc(int x)
{
    ++x;
}

int main()
{
    int x = 42;
    inc(x);
    std::cout << x << std::endl; // still 42, right?
    return 0;
}
\end{consolethree}

Try this:

\begin{consolethree}[escapeinside=||]
void inc(int x[])
{
    ++x[0];
}

int main()
{
    int x[] = {42, 43, 44, 45};
    inc(x);
    std::cout << x[0] << std::endl;
    return 0;
}
\end{consolethree}

Functions \EMPHASIZE{can modify} the \EMPHASIZE{values} in an \EMPHASIZE{array that' s passed in}. In other words, by default, \EMPHASIZE{arrays are pass-by-reference}.

Remember that!!!

For the reason why arrays are so different from variables of basic type you will have to wait till we talk about pointers.

Here' s another example. This function sets all the values in the array parameter to zero:

\begin{consolethree}[escapeinside=||]
void zero_out(int x[], int size)
{
    for (int i = 0; i < size; i++)
    {
        x[i] = 0;
    }
}

int main()
{
    int y[] = {1, 2, 3};
    zero_out(y, 3);
    for (int i = 0; i < 3; i++)
    {
        std::cout << y[i] << ' ';
    }
    std::cout << '\n';
    return 0;
}
\end{consolethree}

Here' s a picture to keep in mind. Right after all parameters of the \texttt{zero\_out()} function are initialized (but before the for-loop) the memory of the functions look like this:

\begin{python}
from latextool_basic import *
p = Plot()
#p += Grid(-1, -1, 7, 7)

p += Rect(1, 5, 1, 5, label='\\texttt{zero\\_out}', linewidth=0)
p += Rect(0, 2.5, 4, 4.5, linewidth=0.1)

p += Rect(0.3, 3.3, 0.9, 3.7, label='\\texttt{x}', linewidth=0)

p += Rect(1, 3.3, 2, 3.7, label='\\texttt{size}', linewidth=0)
p += Rect(2.5, 3.2, 3.2, 3.7, label='\\texttt{3}', linewidth=0.03)

p += Rect(0, -0.5, 4, 1.5, linewidth=0.1)
p += Rect(0.5, 2, 0.5, 2, label='\\texttt{main}', linewidth=0.02)

p += Rect(0.3, 0.5, 0.7, 0.9, label='\\texttt{y}', linewidth=0)

p += Line(points=[(0.25, 3.5), (-0.5, 3.5), (-0.5, 0.75), (0.25, 0.75)], linewidth=0.05, linestyle='dashed', endstyle='>')
p += Rect(1, 0.25, 1.75, 1, linewidth=0.05)
p += Rect(1.75, 0.25, 2.5, 1, linewidth=0.05)
p += Rect(2.5, 0.25, 3.25, 1, linewidth=0.05)
print(p)
\end{python}

As you can see the parameter \texttt{size} (of \texttt{zero\_out()}) has its own memory. It' s just like any regular variable. The initialization of \texttt{size} involves copying the value of 3 to \texttt{size}. Recall that this type of parameter passing is called pass-by-value.

However the parameter \texttt{x} (of \texttt{zero\_out()}) does \EMPHASIZE{NOT} receive the values of the array \texttt{y}. The point of the picture is to show you that \texttt{x} actually \textbf{refers} directly to the memory of \texttt{y}. You can and should think of \texttt{x} as another name for \texttt{y} -- an alias. Or you can think of the \texttt{x} in \texttt{zero\_out()} as a parasite that lives on the memory of \texttt{y}. Therefore changing \texttt{x[0]} in \texttt{zero\_out()} is the same as changing the \texttt{y[0]} in \texttt{main()}. This form of parameter-passing is called \textbf{pass-by-reference}.

Test the function to make sure it works.

\input{temp/20-array-parameters/exercises/7-20-array-parameters/question.tex}\input{temp/20-array-parameters/exercises/8-20-array-parameters/question.tex}
\input{temp/20-array-parameters/exercises/9-20-array-parameters/question.tex}
\input{temp/20-array-parameters/exercises/10-20-array-parameters/question.tex}
\input{temp/20-array-parameters/exercises/11-20-array-parameters/question.tex}
