\sectionthree{Decision problems for DFAs}
\begin{python0}
from solutions import *; clear()
\end{python0}

Here are some decision problems for DFAs.

\begin{ex}
  Decision problems for DFAs.
  It's not too surprising that DFAs can also be encoded.
  Therefore one can ask if it's possible to construct TM or deciders
  to accept DFA encodings (possibly also with encoded  inputs) to
  solve certain problems.
  Are the following problems decidable?
  This means \lq\lq can you find a decider for the given problems.?''
  \begin{tightlist}
    \item Given a DFA $M$, is $L(M) = \{\}$?
    (This is $\EMPTY_\DFA$).
    \item Given a DFA $M$, is $L(M) = \Sigma^*$? ($\Sigma$ is the
    input alphabet of $M$).
    \item Given a DFA $M$, is $\ep \in L(M)$?
    \item Given a DFA $M$, is $L(M)$ infinite?
    \item Given a DFA $M$, is $L(M)$ finite? (What I you are given $M$ and $n$
      and the question is to decide is $|L(M)| = n$?)
    \item Given a DFA $M$ and input $w$ (for $M$), is $w$ in $L(M)$?
    (This is the acceptance problem for DFAs)
    \item Given a DFA $M$ and input $w$ (for $M$), does $M$ halt when
    it runs on input $w$?
    (This is the halting problem for DFAs.)
    \item Given two DFAs $M, M'$, is $L(M)$ the same as $L(M')$?
  \end{tightlist}
  \qed
\end{ex}

Don't peek ... some solutions on next page.

\newpage
\SOLUTION

1.
If $M$ is a DFA, then $L(M) = \{\}$ exactly when there is no
path (in the DFA diagram) from the initial state to any of its
accept state. Therefore we do the following:

Construct a TM $D$ that accepts $\langle M \rangle$
(where $M$ is a DFA) and does the following
(if the input is not the encoding of a DFA, $D$ rejects right away).
The tape for $A$ is of the form
\[
\$ \langle M \rangle \# x \# y
\]
First $A$ writes the encoding of the initial state $q_0$:
\[
\$ \langle M \rangle \# \# \langle q_0 \rangle
\]
$y$ collects the states which are \lq\lq not processed''
while $x$ collects the states which are already \lq\lq processed''.
$A$ will process each state $q$ by computing what states
$q$ can go to via transitions.
The states that $q$ can go to and which are not stored in $x$ and $y$
and stored in $y$.
Once a state is processed, the state is removed from $y$ and
moved to $x$.
If an accept state is stored in $x$, then this means that
there's a path from the initial state of $M$ to 
an accept state of $M$.
Therefore if $A$ sees an accept state in $x$, $A$ enters its reject state.
If $y$ is empty (and no accept state of $M$ appears in $x$),
then $A$ enters its accept state.

3. Construct a TM $A$ such that
when given $\langle M \rangle$, $A$ checks if the initial state is an accept
state. If it is so, $A$ enters its accept state.
Otherwise $A$ enters its reject state.
Clearly $L(A) = \{ \langle M \rangle \mid M \text{ is a DFA},
\ep \in L(M)\}$.

4. HINT:
If $M$ is a DFA, $L(M)$ is infinite iff in the DFA diagram
there is a path from the start state to an accept that contains a loop.


6. HINT: $w$ is in $L(M)$ is $w$ is a path in the DFA diagram of $M$
from $M$'s initial state to one of its accept states.

7. Design a TM $A$ that checks that the input is valid and then
accept. Otherwise, reject. (Right?)

8. HINT: Let $L_i = L(M_i)$.
Let $L = ((L_1 \cap \overline{L_2}) \cup (L_2 \cap \overline{L_1}))$
(the symmetric difference of $L_1$ and $L_2$).
Then $L_1 = L_2$ iff $L = \{\}$.

