\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Clearing the Error and Continuing}

Once your file f hits an error, the error is stored in object f. If you do want to somehow continue with your processing of the file, you need to clear the error information stored in \texttt{f}. This is easy enough. You just do this:

\tab[3em]{\texttt{f.clear();}}\\

Now why and when would you want to do this?

There are times when you cannot simply stop your program when you hit
some corrupted data in your file. You have to attempt to guess how
damaged is your file, and try to reach a point where you are passed the
garbage so that your program can continue to do something useful.

For instance suppose a payroll file looks like this on Monday:

\begin{console}
John Doe 1234.56
Jane Smith 2345.67
Harry Lee 3456.78
\end{console}

and on Tuesday someone from the HR department accidentally entered two
new employees, forgetting that middle names should not be included in
the file:

\begin{console}
John Doe 1234.56
Jane Smith 2345.67
Harry Lee 3456.78
George E. Danger 4567.89
Susan Doyle 5678.90
\end{console}

The payroll program that prints out company checks might hit an error on line 4, printing out only checks for the first 3 employees. A smarter program will attempt to read past the corrupted line, and hope that the next line is correct. If so, this program will continue to print the check for Susan Doyle.

This is why we want to be able to clear the error information in the file object. We'll come back to this later when I've explained to you how to for instance read passed the current line of data.

