%-*-latex-*-
\input{myassignmentpreamble}
\input{ciss358}
\input{yliow}
\renewcommand\TITLE{Assignment 2}

\begin{document}
\topmatter

\textsc{Objectives}
\begin{itemize}
  \li Prove a statement using mathematical induction.
\end{itemize}

\newpage
\input{induction.tex}


\newpage
Q1.
Using mathematical induction, give a complete proof of the following fact:
\[
  4(1^3 + 2^3 + \cdots + n^3) = n^2(n+1)^2
\]
for $n \geq 1$.

\SOLUTION

\input{q01.tex}

  
\newpage
\input{primes.tex}


\newpage
Q2.
Consider the following fact:
Every positive integer $n \geq 1$ is a product of primes.
For instance for $n = 20$,
the ordered collection of primes involved are $(2, 2, 5)$ (in ascending order).
We define the product of the empty collection of primes, i.e. $()$, to be $1$.
Of course the collection can have one single number:
If the collection is $(11)$, then the product is 11.

Define $P(n)$ to be the above statement, i.e.,
\[
  P(n) =
  \biggl(
  \text{Every positive integer $n \geq 1$ is a product of primes}
  \biggr)
\]
Give a complete proof of $P(n)$ for all $n \geq 1$
using strong or weak induction.
Here, I'm giving you the $P(n)$ for free.

You can assume Euclid's lemma: If $p$ is a prime dividing $ab$,
then $p$ divides $a$ or $p$ divides $b$.

(In fact not only is every positive integer $n$ representable
as a product of primes,
it is represented as a product of primes in a \textit{unique way}.
In other words, the ordered collection of primes for $n$ is unique.
This can also be proven using induction.
But you don't have to prove it.)

\SOLUTION

\input{q02.tex}

  
\newpage
Q3.
Consider the following statement:
Every positive integer is a sum of distinct 2-powers.
In other words, there is a set of distinct 2-powers whose sum is the 
given positive integer.
For instance the number 10 can be expressed as
\[
10 = 2^1 + 2^3
\]

(a)
Prove the above statement using strong or weak induction.

(b)
Prove the above statement using the well-ordering principle.

(Note: This is the missing piece of information in CISS360.
In CISS360, we assume that, for instance, all 32-bit unsigned int
can be written as a binary sequence which is a sum of powers of 2
up to the power of 31.
Also, the set of 2-powers adding up to $n$ is unique.
For instance $2^1 + 2^3 = 10$ and if you look at another
sum of 2--powers such as $2^1 + 2^2$, you will not get $10$.
But you don't have to prove it.)

\SOLUTION

\input{q03.tex}

\end{document}
