\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Review: Automatic variables and automatic reclaim of memory}

Constructors let you initialize objects when they are created. Objects
also need to ``die'' just like ``regular'' variables, i.e., automatic
variables. When I say a variable ``dies'', I mean that the name of the
variable goes out of scope. When a variable goes out of scope, the
memory used by the variable in the local scope is reclaimed
automatically. (Remember that the memory used in the free store is not
reclaimed automatically -- the memory in the free store was requested
manually by a memory allocation command using \texttt{new} and you have to
reclaim it yourself using \texttt{delete} or \texttt{delete[]}.)

Here are some examples and scenarios ... MAKE SURE YOU STUDY THEM
CAREFULLY!!!

\begin{consolethree}[escapeinside=||]
if (x > 0)
{
    int i; // i created
    ...
} // i goes out of scope at the end of the
  // if-block and i's memory is reclaimed
  // automatically.
\end{consolethree}

\begin{consolethree}[escapeinside=||]
for (int i = 0; i < 10; i++)
{
    int x[100]; // x created
} // x goes out of scope after the body of
  // the for-loop. x is destroyed and memory
  // for x (100 integers) is reclaimed and
  // execution returns to the update section
  // of the for-loop
\end{consolethree}

\begin{consolethree}[escapeinside=||]
void f()
{
    int i; // i created
    return; // i goes out of scope and i's memory
            // is reclaimed automatically.
}
\end{consolethree}

\begin{consolethree}[escapeinside=||]
void g()
{
    double * i; // i created
    i = new double[100]; // 100 doubles allocated
                         // in the heap (and i's
                         // value is to set to the
                         // address of the first
                         // double of this array of
                         // doubles)
    return; // i goes out of scope and i's memory is
            // reclaimed automatically. However the
            // 100 doubles allocated from the heap is
            // not deallocated ... MEMORY LEAK!
}

int main()
{
    g();
    g();
    g();
    g();
    return 0;
}
\end{consolethree}

All the above is OLD STUFF. If you don't remember, you
have to comb through the old notes and look for information on scopes on
variables.

Now ...


