\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Default destructor}

Since objects call their destructors just before they die (i.e. just
before they go out of scope) and our previous version of \texttt{Date}
does not have a destructor, how come the previous version can still
compile and run??? Shouldn't your C++ compiler yell at
you and tell you that \texttt{\~{}Date()} is not found???

\EMPHASIZE{No.}

Because ... if you don't specify a destructor for a
class, the C++ compiler actually includes a do-nothing destructor in the
class. This is the \EMPHASIZE{default destructor}. The default destructor
is a do-nothing method because it has an empty body:

\begin{consolethree}[escapeinside=||]
class C
{
public:
    ...
    ~C() {} // default destructor of C
    ...
};
\end{consolethree}

In the case of the earlier \texttt{Date} class, C++ included this
destructor into the \texttt{Date} class when there was no destructor:

\begin{consolethree}[escapeinside=||]
class Date
{
public:
    // constructors
    ~Date() {}
    // other methods
};
\end{consolethree}

If you do write a destructor, then your C++ compiler will not insert a
default destructor for you.

Note that the default destructor supplied by C++ does not do anything.
Since all instance variables are automatic in \texttt{Date}, a destructor
is not necessary. It's still a good (and common)
practice to include destructors
even if it's not necessary. Being explicit is better
than leaving things implicit.

<<<<<<< HEAD
\input{../exercises/2-62-destructors/question.tex}
=======
\begin{ex}
What is the order in which objects call their
destructor? (This is important!)

\begin{consolethree}[escapeinside=||]
int main()
{
    Date date1(1970, 1, 1);
    Date date2(1985, 1, 1);
    Date date3(2010, 10, 10);
    return 0;
}
\end{consolethree}
\end{ex}

>>>>>>> parent of 3043c95 (.)
Why is knowing the order important? Because in complex systems, there
might be some dependencies between objects so that some object \texttt{x}
can exist only if object \texttt{y} exists, i.e., \texttt{y} must be
constructed before \texttt{x} and \texttt{x} must be destroyed before
\texttt{y}.

\begin{consolethree}[escapeinside=||]
int main()
{
    Date date1(1970, 1, 1);
    Date date2(1985, 1, 1);
    Date date3(2010, 10, 10);
    return 0; // date3 calls its destructor,
              // *then* date2 calls its destructor,
              // *then* date1 calls its destructor.
}
\end{consolethree}

So, remember this: \EMPHASIZE{the order of destructor calls is always opposite to the order of constructor calls.}

<<<<<<< HEAD
\input{../exercises/3-62-destructors/question.tex}
\input{../exercises/4-62-destructors/question.tex}
\input{../exercises/5-62-destructors/question.tex}
\input{../exercises/6-62-destructors/question.tex}
\input{../exercises/7-62-destructors/question.tex}
\input{../exercises/8-62-destructors/question.tex}
=======
\begin{ex}
Is it possible to call the destructor ``manually''?

\begin{consolethree}[escapeinside=||]
int main()
{
    Date date(1970, 1, 1);
    date.~Date();
    return 0;
}
\end{consolethree}
\end{ex}

\begin{ex}
What is the output?

\begin{consolethree}[escapeinside=||]
#include <iostream>

class A
{
public:
    ~A() { std::cout << "A::~A\n"; }
};

class B
{
public:
    ~B() { std::cout << "B::~B\n"; }
};

class C
{
public:
    ~C() { std::cout << "C::~C\n"; }
};

int main()
{
    A a;
    B b;
    for (int i = 0; i < 5; i++)
    {
        C c;
        if (i > 3)
        {
            A a;
            if (i == 4) B b;
        }
    }
    return 0;
}
\end{consolethree}
\end{ex}

\begin{ex}
Write a destructor for \texttt{WeatherCtrl} and test it
by inserting a print statement in the destructor to check that it is called. Are all
the instance variables automatic? So, is the default destructor good
enough?
\end{ex}

\begin{ex}
Write a destructor for \texttt{Vehicle} and test it by
inserting a print statement in the destructor to check that it is called. Are all the
instance variables automatic? So, is the default destructor good enough?
\end{ex}

\begin{ex}
Write a destructor for \texttt{Being} and test it by
inserting a print statement in the destructor to check that it is called. Is the default
destructor good enough?
\end{ex}

\begin{ex}
Write a destructor for \texttt{IntArray} and test it by
inserting a print statement in the destructor to check that it is called. Is the default
destructor good enough?
\end{ex}

>>>>>>> parent of 3043c95 (.)

