\tinysidebar{\debug{exercises/{unsorted1/question.tex}}}
Write a template \verb!Array! class with the following 
(public! ... right?) methods:
\begin{console}[fontsize=\footnotesize]
void      insert(int index, T key);
void      remove(int index);
int       find(const T & key); // returns index or -1 (if not found)
const T & operator[](int index) const;
T &  operator[](int index);
\end{console}
Allow duplicate values in an \verb!Array! object.
Also, write a template function to print the contents of your object:
\begin{console}[fontsize=\footnotesize]
template < class T >
std::ostream & operator<<(std::ostream & cout,  const Array &);
\end{console}
If the \verb!Array< int >! object contains 1, 2, 3, then printing it will
give you
\begin{console}[fontsize=\footnotesize]
[1, 2, 3]
\end{console}
Set the maximum capacity of your \verb!Array! objects to 1000000.
Remember my advice in class:
Always write a non-template version first!!!
Collect timing data for the following scenarios for 
\verb!Array< int >! objects:
\begin{itemize}
\item What is the time taken to insert a value when
$n = 100000$, $200000$, $300000$, $400000$, ..., $1000000$?
\item What is the time taken to delete the value at a given index position
when
$n = 100000$, $200000$, $300000$, $400000$, ..., $1000000$?
\item What is the time taken to search for a value when
$n = 100000$, $200000$, $300000$, $400000$, ..., $1000000$?
\item What is the time taken to access a value at an index position when
$n = 100000$, $200000$, $300000$, $400000$, ..., $1000000$?
\end{itemize}
Make a plot.
For your experiments, make sure your array contains distinct
values.
