Q7.
Now complete the
\begin{console}[commandchars=\\\{\}]
void dec_energylevel(Robot & robot, int d);
\end{console}
function.

Test 1
\begin{console}[commandchars=\\\{\}]
\underline{7 c 1 2 100 5}
<Robot: name=c, x=1, y=2, energylevel=95>
\end{console}



\newpage
\textsc{Skeleton code for world}

Although we do have all the positions of the robots in the robots themselves,
it is still convenient to have all the locations marks in some kind of a 
\lq\lq map''.
I'll call this array variable \verb!world!.
For instance it's useful for drawing purposes.
We'll use a 2-dimensional array.

{\small
\begin{Verbatim}[frame=single]
/*********************************************************************
File  : world.h
Author: 

void init(char world[10][10]);
Set all elements in the array to ' ' and erects the walls.

void print(char world[10][10]);
Prints the 2-d array world. This includes a border of 'X' characters
and integers 0-9 marking the x-coordinates going left to right and 
y-coordinates going top to bottom. Here is an example.
   0123456789
  XXXXXXXXXXXX
0 X          X
1 X   c   G  X
2 X     P    X
3 X P        X
4 X   XXXX   X
5 X  r       X
6 X   X      X
7 X d X   P  X
8 X   X      X
9 X   X      X
  XXXXXXXXXXXX

void set(char world[10][10], int x, int y, char c);
Sets world[x][y] to character c.

void clear(char world[10][10], int x, int y);
Sets world[x][y] to ' '.

bool is_unoccupied(char world[10][10], int x, int y);
Returns true exactly when world[x][y] is a space.

bool is_occupied(char world[10][10], int x, int y);
Returns true exactly when world[x][y] is not a space.

**********************************************************************/

#ifndef WORLD_H
#define WORLD_H


#endif
\end{Verbatim}
\begin{Verbatim}[frame=single]
/*********************************************************************

File  : world.cpp
Author:

Implementation of world.h.

*********************************************************************/
#include <iostream>
#include "world.h"

void init(char world[10][10])
{
}

// Other functions ...

\end{Verbatim}
}


