%-*-latex-*-
Write a program that prints all $3$--by--$3$ magic squares. (The program does
not accept any input.) The following is part of the printout containing $2$
magic squares:
\begin{console}[fontsize=\small]
+-+-+-+
|2|7|6|
+-+-+-+
|9|5|1|
+-+-+-+
|4|3|8|
+-+-+-+
+-+-+-+
|2|9|4|
+-+-+-+
|7|5|3|
+-+-+-+
|6|1|8|
+-+-+-+
\end{console}

Make sure the format of the printing of the magic squares matches the above 
partial sample output. Also, the magic squares MUST be printed in the
following order:

In the above printout
\begin{console}[fontsize=\small]
+-+-+-+
|2|7|6|
+-+-+-+
|9|5|1|
+-+-+-+
|4|3|8|
+-+-+-+
\end{console}
appears before
\begin{console}[fontsize=\small]
+-+-+-+
|2|9|4|
+-+-+-+
|7|5|3|
+-+-+-+
|6|1|8|
+-+-+-+
\end{console}
because the first magic square is related in an obvious way to the integer
$276951438$ which is less than $294753618$ and $294753618$ is related in an
obvious way to the second magic square.

Obviously, your program is meant to discover the magic square on its own. If
you simply write a program to print the above magic squares then that's
considered cheating and you will get an immediate $-10$.  Your program must go
through all the possible $9$--digit configuration of the magic square in order
to discover the magic squares.

Your program must be reasonably efficient. In other words, points will be
taken off if you have not thought about ways to optimize your program.

\textsc{Note}.
A $3$--by--$3$ magic square is a $3$--by--$3$ grid filled with digits
from $1$ to $9$ in such a way that all digits appear exactly once,
the sum of each row, column, and diagonal
gives you the same value. For instance, the following:
\begin{console}[fontsize=\small]
+-+-+-+
|2|7|6|
+-+-+-+
|9|5|1|
+-+-+-+
|4|3|8|
+-+-+-+
\end{console}
is a magic square since each row, column and diagonal adds up to $15$. In
general, one can also talk about $4$--by--$4$ magic squares, $5$--by--$5$
magic squares, etc. But this question is solving for only the $3$--by--$3$
case.
