\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Member Functions/Methods}

\begin{console}
assign()        // Erases a vector and copies the
                // specified elements to the empty
                // vector.

at()            // Returns a reference to the element at
                // a specified location in the vector.

back()          // Returns a reference to the last
                // element of the vector.

capacity()      // Returns the number of elements
                // that the vector could contain
                // without allocating more storage.

clear()         // Erases all elements of the vector.

empty()         // Tests if the vector container is
                // empty.

erase()         // Removes an element or a range
                // of elements in a vector from
                // specified positions.

front()         // Returns a reference to the first
                // element in a vector.

insert()        // Inserts an element or a number of
                // elements into the vector at a
                // specified position.

max_size()      // Returns the maximum possible
                // length of the vector.

pop_back()      // Deletes the element at the end of the
                // vector.

push_back()     // Add an element to the end of the
                // vector.

resize()        // Specifies a new size for a vector.

reserve()       // Reserves a minimum length of storage
                // for a vector object.

size()          // Returns the number of elements in the
                // vector.

swap()          // Exchanges the elements of two vectors.

vector()        // Constructor. Constructs a vector of a
                // specific size or with elements of
                // specific value.

operator[]      // Returns a reference to the vector
                // element at a specified position.
\end{console}

Operators \texttt{==}, \texttt{!=}, \texttt{=} do the obvious things.

Now for a quick experiment on some of the methods. Run the following
program and study the output carefully.

\begin{console}
#include <iostream>
#include <vector>

std::ostream & operator<<(std::ostream & cout,
               const std::vector< int > & v)
{     
      std::string sep = "";
      cout << '{';
      for (int i = 0; i < v.size() - 1; ++i)
      {   
          cout << sep << v[i];
          if (i == 0) sep = ", ";
      }
      cout << '}';
      return cout;
}

int main()
{   
    int s = 3;
    std::vector< int > v(s);
    
    v[0] = 1; v[1] = 10; v[2] = 100;

    std::cout << "v: " << v << "\n";
    v[0] = 42;
    std::cout << "v: " << v << "\n";

    std::vector< int > u(v);
    std::cout << "u: " << u << "\n";

    if (u == v)
    {  
       std::cout << "same\n";
    }

    v.resize(10);
    std::cout << "v: " << v << "\n";

    std::vector< int > w(4);
    std::cout << "w: " << w << "\n";

    w = u;
    std::cout << "w: " << w << "\n";

    return 0;
}
\end{console}

Study the output carefully. Try all the methods on your own -- most of
them are very similar to the methods in the \texttt{IntDynArr} class.

