\begin{python0}
from solutions import *; clear()
\end{python0}
%-*-latex-*-
\chapter{Print statements and C-strings}


\textsc{Objectives}
\sidenote{
{\Large\textred{\textbf{Why the big, fat right margin? For your notes. Get busy.}}}
}
\begin{tightlist}
\li Print strings and characters using \verb!std::cout!
\li Use the \verb!\n!, \verb!\t!, \verb!\"!, \verb!\'!, 
    \verb!\\! characters
\li Use \verb!std::endl! to force newline
\li Write multiple print statements
\end{tightlist}

In this set of notes, we learn to print strings and characters.

A quick advice to ease the pain of learning your first programming language: 
Type the program \EMPHASIZE{exactly} as given. 
Even your spaces and blank lines must match the spaces  and blank lines in my programs.

Let's begin ...


\newpage\section{Hello World}

Go ahead and run your first program:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello, world!\n";

    return 0;
}
\end{console}

{\it ... commercial break ... go to notes on software tool(s) for writing and  running a program ...}
\sidenote{
\begin{tikzpicture}
  \node[draw=red,text width=4cm,minimum height=2cm,minimum width=2cm, line width=0.1cm, inner sep=0.3cm] at (2,-2)
       {\normalsize{Practice doing a hello
         world program until you
         can do it in $<$ 2 minutes
         and without your notes.}};
\end{tikzpicture}
}

Now let's go back to the C++ code. Right now, you should think of the stuff in bold as the program:
\begin{console}[commandchars=\~\!\@]
#include <iostream>

int main()
{
    ~textbf!*** YOUR "PROGRAM" GOES HERE ***@

    return 0;
}
\end{console}


Therefore you can treat this as a template:


\sidenote{
  \begin{tikzpicture}
  \node[draw=red,text width=4cm,minimum height=2cm,minimum width=2cm, line width=0.1cm, inner sep=0.3cm] (a) at (2,-2) 
       {\tikzmark{start} \normalsize{ Enter a single program  here.}};
\end{tikzpicture}
}
\begin{console}[commandchars=\~\!\@]
#include <iostream>

int main()
{

    *** PROGRAM *** ~tikzmark!end@

    return 0;
}
\end{console}
\DrawArrow[red, thick]{start}{end}
\begin{console}[commandchars=\~\!\@]
#include <iostream>

int main()
{

    *** PROGRAM *** ~tikzmark!endArrow@
    
    return 0;
}
\end{console}
\sidenote{TODO ... Put your program here }

I will not explain things like \lq\lq \verb!#include <iostream>!'' or \lq\lq\verb!int main()!'' until later. (Technically, they are also part of the program.)

Try this
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello ... world! ... mom!\n";

    return 0;
}
\end{console}

%\input{exercises/240-0001/main.tex}

\input{temp/01-print-strings/exercises/0-01-print-strings/question.tex}

\input{temp/01-print-strings/exercises/1-01-print-strings/question.tex}


\newpage\textsc{Solutions}

\begin{soln}\label{sec:hellocolumbia_soln}
(Go to Exercise \ref{sec:hellocolumbia} on page \pageref{sec:hellocolumbia})
\begin{console}
#include <iostream>

int main()
{
    std::cout << "hello columbia\n";

    return 0;
}
\end{console}
\end{soln}


\begin{soln}
\label{sec:hellodrliow_soln} 
(Go to Exercise \ref{sec:hellodrliow} 
on page \pageref{sec:hellodrliow})
\begin{console}
#include <iostream>

int main()
{
    std::cout << "hello dr. liow, my name is c++.\n";  
 
    return 0;   
}  
\end{console}   
\end{soln}  


\begin{soln} 
\label{sec:debughelloworld_soln} 
(Go to Exercise \ref{sec:debughelloworld} 
on page \pageref{sec:debughelloworld})
\begin{console}[commandchars=\~\@\$]
#include ~underline@<$iostream~underline@>$

int main()
~underline@{$
    std~underline@::$cout <~underline@<$ "Hello, world!\n";

    return 0;
~underline@}$
\end{console}
\end{soln}








\newpage\section{Statement}

Here's the first jargon. Look at our program again:
\begin{console}[commandchars=\~\%\@]
#include <iostream>

int main()
{
    ~textbf%std::cout << "Hello, world!\n";@

    return 0;
}
\end{console}

The line in bold is a 
\EMPHASIZE{statement}. 
In C++, statements must terminate with a 
\EMPHASIZE{semi-colon}. 
Note that the semi-colon is part of the statement. 

At this point you should think of a statement as something that will cause 
your computer to perform some operation(s) when you run the program.

If you like, you can think of a statement as a sentence and the semicolon as a
period. When you're told to write a C++ statement, don't forget the 
semicolon!!! 
That would be like writing a sentence without a period (or question mark or 
exclamation mark) in an english essay.


\input{temp/01-print-strings/exercises/2-01-print-strings/question.tex}

\input{temp/01-print-strings/exercises/3-01-print-strings/question.tex}
See what I mean by this advice:

\begin{itemize}
\item[]
\textit{A quick advice to ease the pain of learning your first programming 
language: Type the program 
\EMPHASIZE{exactly} as given. 
Even the spaces and blank lines must match my programs.}
\end{itemize}

Computers are dumb (and picky about details.) 
We are the smart ones. So … when you communicate with your computer, 
you have to be exact and explicit in your programs.


\input{temp/01-print-strings/exercises/4-01-print-strings/question.tex}


\newpage\textsc{Solution}

\begin{soln}
\label{sec:nosemicolon_soln} 
(Go to Exercise \ref{sec:nosemicolon} 
on page \pageref{sec:nosemicolon})
\\
No it won't work. 
You must have the semicolon.
\qed
\end{soln}


\begin{soln}
\label{sec:period_soln} 
(Go to Exercise \ref{sec:period} 
on page \pageref{sec:period})
\\
Nope. It won't work either. 
You \textit{must} have the semicolon.
\qed
\end{soln}


\begin{soln}
\label{sec:print_soln} 
(Go to Exercise \ref{sec:print} 
on page \pageref{sec:print})
\begin{console}
#include <iostream>

int main()
{
    std::cout << "do you want green eggs and ham?\n";

    return 0;
}
\end{console}
\qed
\end{soln}

\newpage\section{C-Strings}

The stuff in quotes is called a C-string or just a string:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello, world!\n" ;

    return 0;
}
\end{console}

You can think of a string as textual data. (Soon I'll talk about numeric data for numeric computations. Got to have that for computer games, right?)

Double quotes are used to mark the beginning and ending of the string. So technically they are not part of the string. That's why when you run the above program, you do not see double-quotes.


\input{temp/01-print-strings/exercises/5-01-print-strings/question.tex}
In the string 
\verb@"Hello, world!\n"@, \verb!H! is a character. The next character is \verb!e!. Etc. When we talk about characters we enclose them with single-quotes. So I should say character \verb!'H'! rather than \verb!H! or \verb!"H"!. 
You can think of the character as the smallest unit of data in a string. 

The string \verb@"Hello, world!\n"@ contains characters \verb!'H'! and \verb!'e'! and \verb!'l'! and \verb!'l'! and ... However a character such as \verb!'H'! can contain one and only one unit of textual data. So you cannot say that \verb!'Hello'! is a character. 

Note that a string can contain as many as characters as you like: 10, 20, 50, 100, etc. 
There's actually a limit, but we won't be playing around with a string with 1,000,000 characters anyway for now --that would be a pain to type!!! 

Note that a string can contain no characters at all: \verb!""!. 

Of course a string can contain exactly one character. 
For instance, here's a string with one character: \verb!"H"!. 

Note that \verb!"H"! is a string with one character whereas \verb!'H'! is a character. 

You just have to look at what quotes are used to tell if the thingy is a string or a character. That's all.

\input{temp/01-print-strings/exercises/6-01-print-strings/question.tex}
\input{temp/01-print-strings/exercises/7-01-print-strings/question.tex}

\input{temp/01-print-strings/exercises/8-01-print-strings/question.tex}

