\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Whitespaces}

A whitespace is ... well ... a white space.

Spaces, tabs, and newlines are whitespaces.


\input{temp/01-print-strings/exercises/11-01-print-strings/question.tex}
\input{temp/01-print-strings/exercises/12-01-print-strings/question.tex}


In general you can insert whitespaces between 
\lq\lq basic words'' understood by C++. 
These \lq\lq basic words'' are called \textbf{tokens}. 
(Oooooo another big word.)

If you think of statements as sentences, 
semi-colons as periods, then you can think of tokens as words.


For instance the following are some tokens from the above program: 
\texttt{int}, \texttt{return} and even \texttt{\{}.

We say that C++ \textbf{ignores whitespace}. 


\input{temp/01-print-strings/exercises/13-01-print-strings/question.tex}


The above 
shows you that \textbf{::} is a token – you cannot break it down.


\input{temp/01-print-strings/exercises/14-01-print-strings/question.tex}

Try this:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello,          world!\n" ;

    return 0;
}
\end{console}


The whitespaces between tokens are removed when your computer runs your C++ program. So, to the computer, it doesn't matter how much whitespace you insert between tokens -- it still works.

However spaces in a \textit{\textbf{string}} 
are not removed before the program runs. 
The space characters \texttt{' '}
(there are 10 in the above string) are actually characters within the string.

Note that although you can insert whitespaces, 
in general, good spacing of a program makes it easier to read. 
Therefore you must follow the style of spacing shown in my notes. 
In particular, I 
\EMPHASIZE{don't} want to see monstrous programs like this 
(although the program does work):
\begin{console}
#include <iostream>

int                        main(){std
::
                    cout<<

"Hello, world!\n"; return 0; }
\end{console}
or this: 
\begin{console}
#include <iostream>
int main(){std::cout<<"Hello, world!\n";return 0;}
\end{console}
And don't try to be cute this like ...
\begin{console}
#include <iostream>
int
 main
  ()
   {
    std
     ::
      cout
       <<
        "Hello, world!\n"; 
         return 0; 
          }
\end{console}
The above examples are excellent candidates for the F grade. 
The correct coding style produces this:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello, world!\n";

    return 0;
}
\end{console}

The left trailing spaces of a statement is called the 
\EMPHASIZE{indentation} 
of the statement:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello, world!\n";
    TODO: left-right red arrow for indentation
 
    return 0;
}
\end{console}
\sidenote{
TODO ...
Indentation. 
}
It's a common programming style to use 4 spaces for indentation. 








