\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Special characters}


Hmmmm ... you don't see the \verb!\n! printed out. In fact what's it for???

Try this:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello, \nworld!\n";

    return 0;
}
\end{console}
And this
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hel\nlo, \nworld!\n";

    return 0;
}
\end{console}
And finally this:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "Hello, world!";

    return 0;
}
\end{console} 
So you can think of the print cursor as jumping to the next line whenever a 
\verb!\n! is encountered. 

As a matter of fact you cannot separate the 
\verb!\! from the \verb!n!. 
\verb!\n! is consider \textit{\textbf{one}} single character. 
So technically I should write \verb!'\n'!. 
\verb!'\n'! is called the 
\EMPHASIZE{newline character}.




\input{temp/01-print-strings/exercises/15-01-print-strings/question.tex}


Try this:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "1\t2\tbuckle my shoe\n3\t4\t...\n";

    return 0;
}
\end{console}

The \verb!'\t'! move the print cursor to the next tab position. 
\verb!'\t'! is also a character. 
It's called the \textbf{tab character}.

Special characters like the above (the newline and the tab characters) are not printable and so we have to use some special notation to say 
\lq\lq the newline character'' or the 
\lq\lq the tab character''. 
Computer scientists decided (long time ago) to use 
\verb!\! to indicate a special character. 
These are called \textbf{escape characters}.

Now recall that \verb!"! is used to mark the beginning and end of a string. What if you want to print something like this:
\[
	\texttt{.....".....}
\]
Try this:
\begin{console}
#include <iostream>

int main()
{
    std::cout << ".....".....";

    return 0;
}
\end{console}
Doesn't work right? Do you see why?

Now try this:
\begin{console}
#include <iostream>

int main()
{
    std::cout << "He shouted, \"42!\"\n";

    return 0;
}
\end{console}

So in a string, \verb!\"! will let you print \verb!"!. 

\verb!'\"'! is actually a character. 
There are two \verb!'\"'! characters in the above string  
\verb@"He shouted, \"42!\"\n"@.


\input{temp/01-print-strings/exercises/16-01-print-strings/question.tex}
\input{temp/01-print-strings/exercises/17-01-print-strings/question.tex}

\input{temp/01-print-strings/exercises/18-01-print-strings/question.tex}

\input{temp/01-print-strings/exercises/19-01-print-strings/question.tex}


\input{temp/01-print-strings/exercises/20-01-print-strings/question.tex}
\input{temp/01-print-strings/exercises/21-01-print-strings/question.tex}


\input{temp/01-print-strings/exercises/22-01-print-strings/question.tex}

\input{temp/01-print-strings/exercises/23-01-print-strings/question.tex}

The number of characters is a string is called 
\EMPHASIZE{length}
of the string. For instance the string 
\begin{verbatim}
                    "Hello, world!\n"
\end{verbatim}
has a length of 14. 


\input{temp/01-print-strings/exercises/24-01-print-strings/question.tex}

(You'll see much later that in every string, there's actually an extra secret character. This is however not included in the count of the length of the string. Don't worry about this for now. I will be coming back to this later.)









