\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{27. Pointers and Arrays}

\textsc{Objectives}
\begin{itemize}
\item Pointer arithmetic
\item Understand the relationship between pointers and arrays
\item Allocate and deallocate memory for dynamic arrays in the heap
\item Understand array parameters in functions
\item Implement stacks and queues using dynamic arrays
\item Understand and implement directed graph using adjacency matrix
\item Model social media graphs
\end{itemize}

In this set of notes, we see how pointers can be used to create dynamic arrays, i.e., pointers that point to arrays of with variable sizes.

Pointers are not arrays. But rather they have the address of the first value in an array. However in terms of syntax and usage, pointers do seem like arrays. I'll talk about the some of the similarities and differences between arrays and pointers. But the point is, once p points to an array, then the i-th index value of the array that p points to is p[i], which is exactly the array notation that you're already used to.

An array is like a container of values. We store information in these containers. (At this point you have only seen array of integers, of doubles, etc. Later on you will learn about complex values, values such as products in a company where each product contains product name,
price, UPC code, etc.) The problem is that the arrays you have seen up to this point have fixed sizes. But no one can predict how large a container should be. You might think an array of size 10000 is enough. But what if after some time, you actually need 100000? Think for instance of a game. At the beginning of the game, maybe there are only 10 enemy spaceships and at most 100 lasers are flying around in the screen. But at a later level, maybe there are 1000 enemy spaceships and 1000000 lasers. Is that enough? Maybe not for an even later level.

Dynamic arrays allows you to handle the situations like the above and they are therefore extremely important. There are actually many, many, many types of containers: fixed size arrays, dynamic arrays, and then in CISS350, I'll show you linked lists, trees, hashtables, etc.

Pointers also allow you to create complex 2D arrays which are not rectangles. For instance if you want to create a 2D array where the first row has 1000000 values, the second has 5000, the third has 20000, and the fourth has 20, it can be done. Without pointers, you would have to create a 2D array with 4 rows and each with 1000000 values.


