\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Pointer arithmetic}

Try this:

\begin{consolethree}[escapeinside=||]
int x = 42;
int * p = &x;
int * q = p + 1;
std::cout << (unsigned int)(p) << ' '
          << (unsigned int)(q) << '\n';
\end{consolethree}

Aha! You can add integers to pointers. Well this is not exactly shocking
since addresses are integers.

Note the address value stored in \texttt{q} ... numerically speaking the value of \texttt{q} is actully the numerical value of \texttt{p} plus 4. when you add 1 to the address of an \texttt{int} address -- IT'S NOT ONE!!! Why 4? Because \ldots{} \EMPHASIZE{4 comes from \texttt{sizeof(int)}}.

But why does it work this way? Because this allows you to think like this:

\begin{itemize}
\item ``\textit{p + 1} points to the \textit{int} value that is immediately
  after the \textit{int} value that \textit{p} points to.''
\end{itemize}

Of course p + 2 has address value that is 8 away from 8. So if p has address value 124500, then you should have this picture in your mind:

\begin{python}
from latextool_basic import *
p = Plot()

p += Rect(0, 1, 3, 5, label='\\EMPHASIZE{42}', linewidth=0.0, background='gray!20!')
p += Rect(0, 5, 3, 9, label='\\EMPHASIZE{3467}', linewidth=0.0, background='gray!65!')
p += Rect(0, 9, 3, 13, label='\\EMPHASIZE{829}', linewidth=0.0, background='gray!20!')

p += Rect(0, 14, 3, 15, label='...', linewidth=0.0)
p += Rect(0, 13, 3, 14, linewidth=0.05)
p += Rect(0, 12, 3, 13, linewidth=0.05)
p += Rect(0, 11, 3, 12, linewidth=0.05)
p += Rect(0, 13, 3, 11, linewidth=0.05)
p += Rect(0, 9, 3, 13, linewidth=0.05)
p += Rect(0, 8, 3, 9, linewidth=0.05)
p += Rect(0, 7, 3, 8, linewidth=0.05)
p += Rect(0, 6, 3, 7, linewidth=0.05)
p += Rect(0, 5, 3, 6, linewidth=0.05)
p += Rect(0, 4, 3, 5, linewidth=0.05)
p += Rect(0, 3, 3, 4, linewidth=0.05)
p += Rect(0, 2, 3, 3, linewidth=0.05)
p += Rect(0, 1, 3, 2, linewidth=0.05)
p += Rect(0, 0, 3, 1, linewidth=0.05)
p += Rect(0, -1, 3, 0, linewidth=0.05)
p += Rect(0, -2, 3, -1, label='...', linewidth=0.0)

p += Rect(-3, 13, 0, 14, label='124512', align='r', linewidth=0.0)
p += Rect(-3, 12, 0, 13, label='124511', linewidth=0.0)
p += Rect(-3, 11, 0, 12, label='124510', linewidth=0.0)
p += Rect(-3, 10, 0, 11, label='124509', linewidth=0.0)
p += Rect(-3, 9, 0, 10, label='124508', linewidth=0.0)
p += Rect(-3, 8, 0, 9, label='124507', linewidth=0.0)
p += Rect(-3, 7, 0, 8, label='124506', linewidth=0.0)
p += Rect(-3, 6, 0, 7, label='124505', linewidth=0.0)
p += Rect(-3, 5, 0, 6, label='124504', linewidth=0.0)
p += Rect(-3, 4, 0, 5, label='124503', linewidth=0.0)
p += Rect(-3, 3, 0, 4, label='124502', linewidth=0.0)
p += Rect(-3, 2, 0, 3, label='124501', linewidth=0.0)
p += Rect(-3, 1, 0, 2, label='124500', linewidth=0.0)

p += Rect(-4.5, 1.5, -4.5, 1.5, label='p', align='l', linewidth=0.0)
p += Rect(-4.5, 2.5, -4.5, 2.5, label='p + 1', align='l', linewidth=0.0)
p += Rect(-4.5, 3.5, -4.5, 3.5, label='p + 2', align='l', linewidth=0.0)

p += Line(points=[(-3.7, 1.5), (-2.5, 1.5)], linewidth=0.1, endstyle='>')
p += Line(points=[(-3.7, 2.5), (-2.2, 5)], linewidth=0.1, endstyle='>')
p += Line(points=[(-3.7, 3.5), (-2.2, 9)], linewidth=0.1, endstyle='>')

print(p)
\end{python}

Now you might be wondering: ``Wait \ldots{} \texttt{p} points to \texttt{x} and I do see that the value of \texttt{x} is 42. But what is the 3467 that \texttt{p + 1} points to?''.

\texttt{p + 1} simply points to an integer (because \texttt{p} is a pointer to \texttt{int}), i.e., the four bytes starting at \texttt{p + 1} as an integer just happens to be 3467. It does not belong to any variable. It's just part of your computer's memory. And of course it could be garbage and the value would be different next time you run the same program.

The above applies to any type and not just \texttt{int}. In general suppose \texttt{T} is a type (\texttt{int} or a \texttt{char} or a \texttt{double} or a \texttt{bool} or what-have-you), and you have the following:

\verb!T * p = &x;!

and suppose a value of type \texttt{T} uses k bytes (which can be obtained using \texttt{sizeof(T)}), then numerically the address value of \texttt{p + 1} is always the address value of \texttt{p} plus \texttt{k}. Also, if we have

\begin{consolethree}[escapeinside=||]
T * p = &x;
T * q = p - 1;
\end{consolethree}

then \texttt{q} is k smaller than \texttt{p}.

\EMPHASIZE{Remember ...}

\begin{itemize}
\item if \texttt{p} is an \texttt{int} pointer (and if you're
  using a 32-bit machine), then \texttt{p + 1} does \EMPHASIZE{not} mean
  ``next address value after \texttt{p}''. Rather, it means address of the
  next \EMPHASIZE{integer} after the \EMPHASIZE{integer} \texttt{p} points to.
\item If \texttt{q} is a pointer that points to a \texttt{double}, then \texttt{q
  + 1} is the address of the next \EMPHASIZE{double} after the
  \EMPHASIZE{double} \texttt{q} points to.
\item If \texttt{r} is a pointer that points to a character, then \texttt{r + 1}
  is address of the next \EMPHASIZE{character} after the \EMPHASIZE{character}
  \texttt{r} points to.
\end{itemize}

Get it?

By the way:

\texttt{sizeof(double)}      is 8 (a double value takes up 8 bytes)

\texttt{sizeof(char)}        is 1 (a char value takes up 1 byte)

However the \texttt{sizeof(int)} can vary depending on your machine.

If you want to measure the difference between two pointers, you have to type cast the address values not to unsigned int, but to int so that you can get negative values. Here's an example:

\begin{consolethree}[escapeinside=||]
int x = 42;
int * p = &x;
int * q = p + 1;
int * r = p - 1;
std::cout << (int) q - (int) p << ' '
          << (int) r - (int) p << '\n';
\end{consolethree}

\input{exercises/0-27-pointers-and-arrays/main.tex}
\input{exercises/1-27-pointers-and-arrays/main.tex}
\input{exercises/2-27-pointers-and-arrays/main.tex}
\input{exercises/3-27-pointers-and-arrays/main.tex}
\input{exercises/4-27-pointers-and-arrays/main.tex}
\input{exercises/5-27-pointers-and-arrays/main.tex}
\input{exercises/6-27-pointers-and-arrays/main.tex}
\input{exercises/7-27-pointers-and-arrays/main.tex}

