\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Stack using a dynamic array}

A \EMPHASIZE{stack} is like an array except that it supports the following
operations:

\begin{itemize}
\item
  You can put a value into the stack. This is called \EMPHASIZE{push}. The
  value you just put into the stack is called the \EMPHASIZE{top} of the
  stack.
\item
  You can remove a value from the stack. The value that is removed is
  always the last one that was put into the stack. This is called
  \EMPHASIZE{pop}.
\item
  You have access to the number of values in the stack. I'm going to
  call this the \EMPHASIZE{size} of the stack.
\item
  You can also look at the value that is to be popped. I'll call this
  \EMPHASIZE{peek}. Peeking at a stack returns a copy of the value, it does
  not remove the top.
\end{itemize}

Think of a stack of plates at a buffet restaurant: When you add a plate to the stack, that new plate goes on top. When you take a plate, you take the one on top (well \ldots{} usually).

You can implement a stack using a dynamic array. The top of the stack is the value at the last index of the array. If you push a value onto the stack, you will need to enlarge the array by 1. If you pop the stack, you will need to replace the array with another one that is 1 size smaller; of course you need to copy the original values to the new stack (except for the top.)

Pushing the stack is not new, you have actually seen this in an earlier exercise on the \texttt{array\_append()} function.

Stacks are very important and are used in many areas of CS including AI, language processing, etc. A stack should be thought of as a memory device. You can ask the stack ``Hey stack \ldots{} what was the \EMPHASIZE{last thing} I put into you?''

\begin{ex}
Write a program that implements a stack of integers using dynamic arrays. Test it thoroughly! You can test your stack implementation with the following:

\begin{consolethree}[escapeinside=||]
int main()
{
    int * stack = NULL;
    int stack_size = 0;
    println(stack, stack_size); // Get {}
    push(&stack, &stack_size, 5);
    println(stack, stack_size); // Get {5}
    push(&stack, &stack_size, 3);
    println(stack, stack_size); // Get {5, 3}
    push(&stack, &stack_size, 9);
    println(stack, stack_size); // Get {5, 3, 9}
    int x;
    x = peek(stack, stack_size);
    std::cout << x << std::endl; // Get 9
    println(stack, stack_size); // Get {5, 3, 9}
    pop(&stack, &stack_size);
    println(stack, stack_size); // Get {5, 3}
    return 0;
}
\end{consolethree}
\end{ex}

There are many ways to implement a stack. Later you'll see that a more efficient way of implementing a stack is to use a \EMPHASIZE{singly linked list}, something that uses pointers.

