\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Queue using a dynamic array}

A \EMPHASIZE{queue} is like an array except that it supports the following operations:

\begin{itemize}
\item
  You can put a value into a queue. This is called \EMPHASIZE{enqueue}. The
  value you just put into the queue is called the \EMPHASIZE{back} of the
  queue.
\item
  You can remove a value from the queue. The value that is removed is
  always the value in the queue that has been in the queue the longest.
  This is called the \EMPHASIZE{front} of the queue. This operation is
  called \EMPHASIZE{dequeue}.
\item
  You have access to the number of values in the queue. I'm going to
  call this the \EMPHASIZE{size} of the queue.
\item
  You can also look at the value of the front and back of the queue
  without removing these values. These operations are sometimes called
  \EMPHASIZE{front} and \EMPHASIZE{back}.
\end{itemize}

You think of a queue as a line of people in front of a ticket booth: When a person joins a queue, he/she joins it at the back. When the booth is ready to serve the next customer, the person to leave the queue to go to the booth is the one in front of the queue.

You can implement a queue using a dynamic array. The front of the queue is the value at index 0 of the array. The back of the queue is the value at the last index of the array. If you enqueue a value into the queue, you will need to enlarge the array by 1. If you dequeue the stack, you will need to replace the array with another one that is 1 size smaller.

Like stacks, queues are very important and are used in many areas of CS including AI, language processing, etc. A queue should be thought of as a memory device. You can ask the queue ``Hey queue \ldots{} what was the \EMPHASIZE{earliest thing} that you can think of?''

<<<<<<< HEAD
\input{../exercises/50-27-pointers-and-arrays/question.tex}
=======
\begin{ex}
Write a program that implements a queue of integers using dynamic arrays. Test it thoroughly! You can test your queue implementation with the following:

\begin{consolethree}[escapeinside=||]
int main()
{
    int * queue = NULL;
    int queue_size = 0;
    println(queue, queue_size); // Get {}
    enqueue(&queue, &queue_size, 5);
    println(queue, queue_size); // Get {5}
    enqueue(&queue, &queue_size, 3);
    println(queue, queue_size); // Get {5, 3}
    enqueue(&queue, &queue_size, 9);
    println(queue, queue_size); // Get {5, 3, 9}
    int x;
    x = front(queue);
    std::cout << x << std::endl; // Get 5
    x = back(queue, queue_size);
    std::cout << x << std::endl; // Get 9
    println(queue, queue_size); // Get {5, 3, 9}
    dequeue(&queue, &queue_size);
    println(queue, queue_size); // Get {3, 9}
    return 0;
}
\end{consolethree}
\end{ex}

>>>>>>> parent of 3043c95 (.)
There are many ways to implement a queue. Later you'll see that a more efficient way of implementing a queue is to use a \EMPHASIZE{doubly linked list}, something that uses pointers (and is very similar to singly linked list.)

