\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Set using a dynamic array}

A \EMPHASIZE{set} is a container of values where you don't care about the number of times a value occurs and you don't really care about the index position of a value. It supports the following operations:

\begin{itemize}
\item
  You can \EMPHASIZE{add} a value to a set. You can put it anywhere you
  like in your dynamic array. For instance you can put is at the last
  index position, like the \texttt{array\_append()} function. Since you
  don't really care how many times a value occurs, if you try to add a
  value into the set and the set already contains that value, you simply
  do NOT add the value into the set, i.e., a set does not contain
  duplicates.
\item
  You can \EMPHASIZE{remove} a value from a set. If you attempt to remove a
  value not in the set, then nothing happens -- the set is unchanged.
\item
  You can check if a value is \EMPHASIZE{in} a set. I'll call this
  \texttt{has\_member()} function.
\item
  You have access to the number of values in the set. I'll call this
  \EMPHASIZE{size}.
\end{itemize}

You can think of a set also as a memory device except that you don't care about when you started remembering a value.

<<<<<<< HEAD
\input{../exercises/51-27-pointers-and-arrays/question.tex}
=======
\begin{ex}
Write a program that implements a set of integers using dynamic arrays. Test it thoroughly! You can test your set implementation with the following:

\begin{consolethree}[escapeinside=||]
int main()
{
    int * set = NULL;
    int set_size = 0;
    println(set, set_size); // Get {}
    add(&set, &set_size, 5);
    println(set, set_size); // Get {5}
    add(&stack, &set_size, 3);
    println(set, set_size); // Get {5, 3}
    add(&set, &set_size, 5);
    println(set, set_size); // Get {5, 3}
    add(&set, &set_size, 3);
    println(set, set_size); // Get {5, 3}
    bool b;
    b = has_member(set, set_size, 5);
    std::cout << b << std::endl; // Get 1
    println(set, set_size); // Get {5, 3}
    b = has_member(set, set_size, 3);
    std::cout << b << std::endl; // Get 1
    println(set, set_size); // Get {5, 3}
    b = has_member(set, set_size, 2);
    std::cout << b << std::endl; // Get 0
    println(set, set_size); // Get {5, 3}
    remove(&set, &set_size, 3);
    println(set, set_size); // Get {5}
    remove(&set, &set_size, 0);
    println(set, set_size); // Get {5}
    return 0;
}
\end{consolethree}
\end{ex}

>>>>>>> parent of 3043c95 (.)
There are many ways to implement a set. The above is actually not
efficient. Later you will see that better ways to implement a set
includes using a balanced tree and a hashtable.

