\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{More Examples}

Here's a typical example that uses the \texttt{while}-loop.
\begin{console}
int i = 0;
std::cout << "gimme a number ...";
std::cin >> i;
while (i != 42)
{     
      std::cout << "so 42 is not your fav number?";
                << std::endl;
      std::cout << "gimme a number ...";
      std::cin >> i;
}
\end{console}

Notice something about this \texttt{while}-loop?

Hmmmm ....there is \EMPHASIZE{no control variable} unlike our \texttt{for}-loop examples. Now try this:
\begin{console}[commandchars=\~\@\$]
~textbf@int sum = 0;$
int i = 0;
std::cout << "gimme a number ...";
std::cin >> i;
while (i != 42)
{     
      ~textbf@sum += i;$
      std::cout << "sum: " << sum
                << std::endl;
      std::cout << "gimme a number ...";
      std::cin >> i;
}
std::cout << "final sum: " << sum << std::endl;
\end{console}

In general, if you want to process a bunch of numbers, you can run over
these numbers in two different ways: Either the program knows how to
compute them (example: the numbers are 1, 2, 3, 4, 5, 6, 7, 8, 9, 10) or
the program does not (example: it's based on user
input).

For the first case, you should use the \EMPHASIZE{for-loop} with a
\EMPHASIZE{control or index or counter variable} running over the values.

For the second case, you should use the \EMPHASIZE{while-loop}. To let the
user stop the data processing, you have to pick a special value for the
user to use to indicate end-of-data. This special value is called the
\EMPHASIZE{sentinel value}.

Now I want you to step back and look at the ``structure''of the above
program:

\tab[3em]{initialization}\\
\tab[3em]{prompt user for data}\\
\tab[3em]{while value of data is not sentinel value:}\\
\tab[5em]{process data}\\
\tab[5em]{prompt user for data}\\

Note that there is code duplication:

\tab[3em]{initialization}\\
\tab[3em]{\textbf{prompt user for data}}\\
\tab[3em]{while value of data is not sentinel value:}\\
\tab[5em]{process data}\\
\tab[5em]{\textbf{prompt user for data}}\\


<<<<<<< HEAD
\input{../exercises/0-12-while-loops/question.tex}\input{../exercises/1-12-while-loops/question.tex}
\input{../exercises/2-12-while-loops/question.tex}
=======
\begin{ex}
Write a program that continually ``rolls a die''and prints
the face value of the die until 6 is reached. (Use rand() to generate a
random integer between 1 and 6.)
\end{ex}
\begin{ex} Write a program that computes the product of integers entered
by the user. Use a sentinel value of 0 to indicate the end of data.
Here's the skeleton code:
\begin{console}[commandchars=\~\@\$]
~textbf@int product = _____;$
int i = 0;

std::cout << "gimme a number ...";
std::cin >> i;
~textbf@while (__________)$
{    
     _____________________;
     std::cout << "product: " << product
               << std::endl;
     std::cout << "gimme a number ...";
     std::cin >> i;
}
std::cout << "final product: " << product << std::endl;
\end{console}
\end{ex}

\begin{ex} Write a program that computes the cost of head grafting
surgery. Each head costs \$100.42. Display the total number of heads
requested. The program exits the loop when the number of heads requested
is zero. Here is the skeleton code:
\begin{console}[commandchars=\~\@\$]
int heads = 0;
double cost = 0.0;
int totalHeads = _____;
double totalCost = _____;

std::cout << "how many heads? ";
std::cin >> heads;
~textbf@while (__________)$
{     
      cost = ________________________________;
      totalHeads = ______________________________;
      totalCost = ______________________________;
      std::cout << "total heads requested: "
                << totalHeads << std::endl
                << "total cost: "
                << totalHeads <<
                '\n' << std::endl;
      // prompt for heads
}

std::cout << "your final bill:" << totalCost << '\n'
          << "i hope you enjoy your extra "
          << totalHeads << " heads" << std::endl;
\end{console}
\end{ex}

\begin{ex}
Due to popular demand and lack of staff and surgeons
(and donors), we have to limit the number of extra heads per visit.
Modify the above program so that the program exits the while-loop when
the number of heads requested is zero or when the total number of heads
requested exceeds 42.
\end{ex}
\begin{ex} Write a simple number guessing game. The program
generates a random integer from 1 to 10 (inclusive), and continually
prompts the user to guess the number until the correct answer is given.
When you're done with that, get the program to print the
number of tries.
\end{ex}
\begin{ex} Modify our multiplication game so that it keeps
posing questions until the user answers 5 questions correctly. While
prompting the user, print the number of correct answers. (Once
you're done with that, make a slight modification to
prompt the user for the number of correct answers before terminating the
program.)
\end{ex}
\begin{ex} Modify the above program so that the program
terminates when 5 \texttt{\textbf{consecutive}} correct answers are given.
\end{ex}
>>>>>>> parent of 3043c95 (.)
