\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Associativity and precedence}



College Algebra is a prerequisite. So you know what I mean by
associativity and operator precedence right? Yes? No?


All the operators you saw (+, -, *, /, \%) are binary operators in the sense
that for instance + is applied to two values. For example we have 2 + 3.


\textbf{Associativity.}

Now think about this integer expression:
\[2+3+1\]
What we really mean is
\[(2 + 3) + 1\]
As you can see with the parentheses, each + is applied to two numbers.
\begin{console}
\li The left + is applied to 2 and 3.
\li The right + is applied to (2+3) and 1, i.e., 5 and 1.
\end{console}
Correct?


But wait. There's another way to do it:
\[2 + (3 + 1)\]

So is 2 + 3 + 1
\[(2 + 3) + 1 \text{     or     } 2 + (3 + 1)\text{    ???}\]
Of course it doesn't matter!!! Ultimately they both evaluate to the same value:

\begin{align*}
&(2 + 3) + 1 & & 2 + (3 + 1)\\
&=5+1 & & =2+4\\
&=6 & & =6
\end{align*}

But this is not the case for all operators. For instance, consider the minus
operator:
\begin{align*}
&(2 - 3) - 1 & & 2 - (3 - 1)\\
&=-1 - 1 & & =2-2\\
&=-2 & & =0
\end{align*}

Long time ago, mathematicians decided that they are too efficient to write
\[(2 + 3) + 1\]
So they got together and come to an agreement that if parentheses are
left out for a string of pluses, you always do the pluses from left to right.
We say that \textbf{{+ associates left-to-right or associates to the right.}} So
\[1+2+3+4\]
is really a shorthand for
\[((1 + 2) + 3) + 4\]
The important thing to remember is that whether + is left or right
associated is a matter of convention.

+, –, *, /, \% associates from left to right. So for instance if I write
\[3–4–5–2\]
I mean
\[((3 – 4) – 5) – 2\]
i.e. do the leftmost subtraction first.


\textbf{{Precedence Rules}}

What about the case where you have not just pluses? What about
something like
\[1+3–2*6/3*4+6\]
In C++, you determine which operator goes first using this table:

\noindent\fbox{\begin{minipage}{\textwidth}
    \begin{align*}
      &() & &\text{first priority}\\
      &*, /, \% & &\text{second priority}\\
      &+, - & &\text{third priority}
    \end{align*}
\end{minipage}}

Remember PEMDAS from your math classes? Note the difference is that
* and / are at the same level, and + and – are at the same level.

%\input{exercises/240-0001/main.tex}

It will be clearer when I do an example. So let's evaluate the above
expression:
\[1+3–2*6/3*4+6\]
When we check the priority table, we don't see parentheses, i.e., there's
nothing for the first priority. What about the second priority operators? I
do see * and /. So let's do them. Don't forget that these associate left-to-
right which is just a fancy way of saying “do the leftmost first and keep
moving right”. You see that for * and / (i.e., ignore + and – for now), the
first to occur is the * here:
\[1+3–\underline{2*6}/ 3*4+6\]
So let's do that:
\begin{align*}
  &1+3–\underline{2*6}/ 3*4+6\\
  &= 1 + 3 – 12 / 3 * 4 + 6 & &i.e., 2 * 6 = 12
\end{align*}
The next in the second priority operators is here:
\[1 + 3 – \underline{12 / 3} * 4 + 6\]
Let's do that:
\begin{align*}
  &1 + 3 – \underline{12 / 3} * 4 + 6\\
  &=1+3–4*4+6 & &i.e., 12 * 3 = 4
\end{align*}
Got it?

And the last evaluation for the second priority operators give us this:
\begin{align*}
&1+3–\underline{4*4}+6\\
&= 1 + 3 – 16 + 6 & &i.e., 4 * 4 = 16
  \end{align*}
    
There are now no more second priority operators. Let's move on to the
third priority operators, i.e., the + and the –. Here we go:
\begin{align*}
  &\underline{1 + 3} – 16 + 6 \\
  &= \underline{4 – 16} + 6 & &i.e., 1 + 3 = 4 \\
  &= \underline{–12 + 6} & &i.e., 4 – 16 = -12 \\
  &= –6 & &i.e., -12 + 6 = -12
\end{align*}
Don't forget that + and – associates left-to-right. So we have to always
pick + and – going left-to-right.

That's all there is to it. If I put all the computations together this is what I
get:
\begin{align*}
  &1+3–\underline{2*6}/ 3*4+6 \\
  &= 1 + 3 – \underline{12 / 3} * 4 + 6 & &i.e., 2 * 6 = 12\\
  &=1+3–\underline{4*4}+6 & &i.e., 12/ 3 = 4\\
  &= \underline{1 + 3} – 16 + 6 & &i.e., 4 * 4 = 16\\
  &= \underline{4 – 16} + 6 & &i.e., 1 + 3 = 4\\
  &= \underline{–12 + 6} & &i.e., 4 – 16 = -12\\
  &= –6 & &i.e., -12 + 6 = -12\\
\end{align*}

%\input{exercises/240-0001/main.tex}

Associative and precedence are extremely important because performing
computations in different orders can lead to different results. If you do it
in the wrong order, then ... that's BAD! Just imagine the collapse of
financial institutions or random explosions of missiles/shuttles because of
a misunderstanding of the way C++ computes value!!!


Fortunately C++ is designed so that computations done in the computer
follows the standard associative and precedence rules.

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}


Let's try another example, this time an expression with a pair of
parentheses:
\[3 + 1 * 4 – 4 * 3 * 2 / (2 + 6 * 2 – 1) + 6 * 7\]
In this case, there's one pair of parentheses. So we have to evaluate the
expression within the parentheses first. In the parentheses, there are no
parentheses, i.e., there are no priority one operators. So we look for
priority two operators: * and /. There is one. So we evaluate that
operator:
\begin{align*}
  &3 + 1 * 4 – 4 * 3 * 2 / (2 + \underline{6 * 2} – 1) + 6 * 7\\
  &= 3 + 1 * 4 – 4 * 3 * 2 / (2 + 12 – 1) + 6 * 7 & &i.e., 6 * 2 = 12
\end{align*}
To make it easier to follow, I've underlined the expression I'm evaluating.

There are no priority two operators. So we look for priority 3 operators,
i.e., + and –. There are two so we do this going left-to-right:
\begin{align*}
&3 + 1 * 4 – 4 * 3 * 2 / (\underline{2 + 12} – 1) + 6 * 7\\
&= 3 + 1 * 4 – 4 * 3 * 2 / (\underline{14 – 1}) + 6 * 7 & &i.e., 2 + 12 = 14\\
&= 3 + 1 * 4 – 4 * 3 * 2 / (13) + 6 * 7 & &i.e., 14 – 1 = 13\\
&= 3 + 1 * 4 – 4 * 3 * 2 / 13 + 6 * 7 & &i.e., (13) = 13
\end{align*}
(In the last step we just remove the useless parentheses.) Now there are
no parentheses (priority one). So look for priority two operators and
evaluate them left-to-right:
\begin{align*}
&3 + 1 * 4 – 4 * 3 * 2 / 13 + 6 * 7\\
&= 3 + 4 – 4 * 3 * 2 / 13 + 6 * 7 & &i.e., 1 * 4 = 4\\
&= 3 + 4 – 12 * 2 / 13 + 6 * 7 & &i.e., 4 * 3 = 12\\
&= 3 + 4 – 24 / 13 + 6 * 7 & &i.e., 12 * 2 = 24\\
&=3+4–1+6*7 & &i.e., 24 / 13 = 1\\
&= 3 + 4 – 1 + 42 & &i.e., 6 * 7 = 42
\end{align*}

Don't forget that in the fourth line, 24/13 = 1 because we're doing C++
integer quotient, not real number division in college algebra.
Now there are no more priority two operators. We evaluate the priority
three operators, i.e., the + and –.
\begin{align*}
&3 + 4 – 1 + 42\\
&= 7 – 1 + 42 & &i.e., 3 + 4 = 7\\
= 6 + 42 & &i.e., 7 – 1 = 6\\
= 48 & &i.e., 6 + 42 = 48
\end{align*}
and we're done.
Study every single step of the above computation.

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

f you have an expression with \underline{two} separate pairs of parentheses such
as:
\[2 * (3 + 5) – 3 * (5 + 3 * 2)\]
you should present your answer by evaluating left-to-right, i.e., evaluate
the leftmost parentheses first.
\begin{align*}
&2 * (3 + 5) – 3 * (5 + 3 * 2)
&= 2 * (8) – 3 * (5 + 3 * 2) & &i.e., 3 + 5 = 8\\
&= 2 * 8 – 3 * (5 + 3 * 2) & &i.e., (8) = 8\\
&= 2 * 8 – 3 * (5 + 6) & &i.e., 3 * 2 = 6\\
&= 2 * 8 – 3 * (11) & &i.e., 5 + 6 = 11 \\
&= 2 * 8 – 3 * 11 & &i.e., (11) = 11 \\
&= 16 – 3 * 11 & &i.e., 2 * 8 = 16\\
= 16 – 33 & &i.e., 3 * 11\\
= -17 & &i.e., 16 – 33 = -17
\end{align*}

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

You can also have situations when parentheses are nested like this:
\[1 + (\underline{2 * 1 + (3 + 4 * 5) / 2}) + 5\]
i.e., within a pair of parentheses, you see another pair. As you're
evaluating the first pair of parentheses, you have to apply the priority
rules, which means you have to evaluate the inner parentheses first.
That's all.
\begin{align*}
  &1 + (2 * 1 + (3 + \underline{4 * 5}) / 2) + 5\\
  &= 1 + (2 * 1 + (\underline{3 + 20}) / 2) + 5 & &i.e., 4 * 5 = 20\\
  &= 1 + (2 * 1 + \underline{(23)} / 2) + 5 & &i.e., 3 + 20 = 23\\
  &= 1 + (\underline{2 * 1} + 23 / 2) + 5 & &i.e., (23) = 23\\
  &= 1 + (2 + \underline{23 / 2}) + 5 & &i.e., 2 * 1 = 2\\
  &= 1 + (\underline{2 + 11}) + 5 & &i.e., 23/ 2 = 11\\
  &= 1 + \underline{(13)} + 5 & &i.e., 2 + 11 = 13\\
  &= \underline{1 + 13} + 5 & &i.e., (13) = 13\\
  &= \underline{14 + 5} & &i.e., 14 + 5 = 19\\
  &= 19
\end{align*}

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}


