\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Unary and binary operators}
An operator is \EMPHASIZE{binary} if it is applied to two values. For instance the +
in the following expression:
\[2+3\]
is binary. You have already seen the binary +, –,*,/,\%. They all require
two values to make sense.

There are actually two operators that require only one value. These are \EMPHASIZE{unary} operators. The unary + operator (the “positive” operator)
appears in this expression:

\sidenote{
\begin{tikzpicture}
  \node[draw,text width=4cm,minimum height=1cm,minimum width=1cm, line width=0.1cm, inner sep=0.3cm] at (2,-2)
       {\normalsize{ +5 is 5}};
       
  \node[draw,text width=4cm,minimum height=1cm,minimum width=1cm, line width=0.1cm, inner sep=0.3cm] at (2,0)
       {\normalsize{ -(-3) is 5}};
\end{tikzpicture}
}


\[+2\]
and of course there's the unary – operator (the “negative” operator):
\[–2\]
So the symbol + and – have two different meanings.

Now we need to modify the chart of precedence rules:

  
\noindent\fbox{\begin{minipage}{\textwidth}
    \begin{align*}
      &() & &\text{first priority}\\
      &*, /, \% & &\text{second priority}\\
      &+, - & &\text{third priority}
    \end{align*}
\end{minipage}}

%\input{exercises/240-0001/main.tex}

Try this example involving three unary – operators (the “negative”
operator):
\begin{console}
std::cout << - - - 2;
\end{console}
You should get -2. Right? If you try to put parentheses in the expression,
of course the only reasonable way to do it would be like this:
-\[(-(-2))\]
I mean ... what else can it be??? You can't do this: (((-)-)2)!!! It has
no meaning at all!!!
In other words you apply the rightmost – first. The last – is the leftmost.
The unary – operator associates right to left. It's the same for the unary +
operator (“positive” operator).


