\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Integer operators: +, -, and *}

Now you might say, "Isn't this ..."
\begin{console}
std::cout << 42;
\end{console}
Now you might say, "Isn't this ..."
\begin{console}
std::cout << "42";
\end{console}

Yes. The \underline{output} is the same. But of course the \EMPHASIZE{value}
printed out in
the first program is an integer while the second is a string.

This will REALLY show you the difference. Try this:
\begin{console}
std::cout << 42 + 1 << std::endl;
std::cout << "42 + 1" << std::endl;
\end{console}

Get it?

Soooo ... C++ can do math!!!

Try this:
\begin{console}
std::cout << "42 + 1" << 42 + 1 << std::endl;
std::cout << "42 - 1" << 42 - 1 << std::endl;
std::cout << "2 * 3 = " << 2 * 3 << std::endl;
\end{console}

So far we see that C++ understands three \EMPHASIZE{operators}: +, –, and *.

No surprises there ... I hope!!!

You can (and should) think of the integer operator + as some kind of
machine that accepts two integer inputs and returns to you an integer
value:

\begin{center}
\begin{tikzpicture}
\node[draw,shape=rectangle,minimum size=2cm, inner sep=0](A) at (0, 0){+};
\node[draw=none,shape=rectangle,minimum size=2cm, inner sep=0](A1) at (0, 0.5){};
\node[draw=none,shape=rectangle,minimum size=2cm, inner sep=0](A2) at (0, -0.5){};
\node[draw=none,shape=circle,minimum size=1cm,inner sep=0](B) at (-3,0.5){};
\node[draw=none,shape=circle,minimum size=1cm,inner sep=0](C) at (-3, -0.5){};
\node[draw=none,shape=circle,minimum size=1cm,inner sep=0](D) at (3, 0){};

\draw[line width=0.04cm,black,->] (B) to  (A1);
\draw[line width=0.04cm,black,->] (C) to  (A2);
\draw[line width=0.04cm,black,->] (A) to  (D);


\end{tikzpicture}

\end{center}

And here's what happens when you put 42 and 1 into this machine:
\begin{center}
\begin{tikzpicture}
\node[draw,shape=rectangle,minimum size=2cm, inner sep=0](A) at (0, 0){+};
\node[draw=none,shape=rectangle,minimum size=2cm, inner sep=0](A1) at (0, 0.5){};
\node[draw=none,shape=rectangle,minimum size=2cm, inner sep=0](A2) at (0, -0.5){};
\node[draw=none,shape=circle,minimum size=1cm,inner sep=0](B) at (-3,0.5){};
\node[draw=none,shape=circle,minimum size=1cm,inner sep=0](C) at (-3, -0.5){};
\node[draw=none,shape=circle,minimum size=1cm,inner sep=0](D) at (3, 0){};

\node[anchor=east] at (1.7,0.5)   {43};

\node[anchor=east] at (-1.5,1)   {42};

\node[anchor=east] at (-1.5,-1)   {1};


\draw[line width=0.04cm,black,->] (B) to  (A1);
\draw[line width=0.04cm,black,->] (C) to  (A2);
\draw[line width=0.04cm,black,->] (A) to  (D);


\end{tikzpicture}

\end{center}

n C++, this operator + machine accepts integer inputs. There's another
operator + machine that accepts for instance numbers with decimal
places. The two are totally different machines. This is extremely
important. (I'll talk about numbers with decimal places in another set of
notes. These are called floating point numbers.) If I want to emphasize, I
will call the + in this set of notes, I will call it the
5 of 34 \EMPHASIZE{integer + operator}.

The story is very similar for operator - and *

%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

