\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Integer operators: / and \%}

Division is ... well try this
\begin{console}
std::cout << "4 / 2 = " << 4 / 2 << std::endl;
\end{console}

And then this:
\begin{console}
std::cout << "100 / 20 = " << 100 / 20 << std::endl;
\end{console}

So far so good. No surprises.

BUT ... what about this program:
\begin{console}
std::cout << "9 / 3 = " << 9 / 3 << '\n'
          << "8 / 3 = " << 8 / 3 << '\n'
          << "7 / 3 = " << 7 / 3 << std::endl;
\end{console}
So you see what's happening?

If a and b are integer, then a / b gives the \EMPHASIZE{quotient} of a and b. This
is also called the \EMPHASIZE{integer division} of a by b. This is the same
as losing the fractional part of the usual mathematical division.

Let me repeat that again ...

In C++, 13 / 3 gives the \EMPHASIZE{integer part} of 13 divided by 3. In math, 13 divided by 3 is 4.333333... In C++, when / is used as an
operator for two \EMPHASIZE{integer} values in a C++ program, the result is an
\EMPHASIZE{integer}. So you lose the fractional part of the result from
mathematical division.

%\input{exercises/240-0001/main.tex}

\EMPHASIZE{REMEMBER THAT!}

When / is applied to two integers, we say that this division is an
\EMPHASIZE{integer division}. (Of course since we call this an integer
division operator you would expect the division operator in 1.123 / 3.2343
to be something totally different. We'll talk about this later.)
An integer division will produce an integer except for this case:

\begin{console}
std::cout << 1 / 0 << std::endl;
\end{console}

Run it. You will get an error. I hope that's not surprising! Division by zero
will cause an error. You already know that from your math classes: you
cannot divide by zero.

%\input{exercises/240-0001/main.tex}

Something like 10 + 55 is called an \EMPHASIZE{expression} (or integer
expression since it contains only integers). Here's another integer
expression:
\[ 10 + 55 + 412 – 12 * 43 – 67\]

Computing the resulting value of the expression is called
\EMPHASIZE{evaluating} the expression.

But what about the fractional part that's lost when doing an integer
division? What if you really wanted it?

Consider 13 / 3. C++ will give you 4. You're losing one-third. (Right?) In
math, the correct answer is 4 and one-third:

\[\frac{13}{3} = 4\frac{1}{3}\]

The 1 above is called the \EMPHASIZE{remainder}. Here's another example:

\[\frac{43}{8} = 5\frac{3}{8}\]

i.e., when 43 is divided by 8, you get 5 with a remainder of 3. What this
means is very simple. If you have 43 dollar bills and you want to give
them equally to your 8 good friends, then each will get 5 dollars and
you're left with 3 dollars. Let me repeat that – when you're given

\[\frac{43}{8}\]

then in order to write

\[\frac{43}{8} = 5\frac{3}{8}\]

the 5 is the integer division of 43 by 8 and the 3 is the remainder when
43 is divided by 8.

%\input{exercises/240-0001/main.tex}

irst of all you see that \% is an operator. What does the operator \% do?
Answer: 
\% is called the \EMPHASIZE{mod} operator.

%\input{exercises/240-0001/main.tex}

The integer quotient and remainder operators occur frequently in real life.
They are not just some fancy academic fluff. For instance suppose you
are told: “It took John 135 minutes to paint this wall,” but you prefer to
think in terms of hours and minutes. What would you do? You would do
this (mathematically):

\[\frac{135}{60} = 2\frac{15}{60}\]

i.e:

\begin{center}
  134 minutes = 2 hours and 15 minutes
\end{center}

To \underline{check} that this is correct:
\begin{center}
2 hours + 15 minutes\\
= 2 x 60 minutes + 15 minutes\\
= 120 minutes + 15 minutes\\
= 135 minutes
\end{center}

Correct? Of course the 2 is the integer quotient 135 / 60 (in C++) and the
15 is integer mod 135 \% 60 (in C++). And why do we quotient and mod
\underline{by 60}? Because 1 hour equals 60 minutes. Correct?

%\input{exercises/240-0001/main.tex}

if

\[\frac{192}{60} = 3\frac{12}{60}\]

But that's not the whole story. The remainder operator (i.e., the mod or
the \% operator) is in fact crucial to cryptography. Without it, e-commerce
would have been impossible. That's how important it really is. I bet your
high school teacher didn't tell you that.

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

Remember this:
\begin{console}
std::cout << 5 \% 0 << std::endl;
\end{console}
Question: What are you to remember? And this too:
\begin{console}
std::cout << 5 / 0 << std::endl;
\end{console}

Using integer mod \% (and the integer quotient above) to convert from
minutes to hours and minutes might give you the impression that they
are use only for very trivial things. WRONG!!! Without the concept of
integer mod operator, we would not have the explosion of internet e-
commerce. Modern encryption (and many other things!!!) depends on the
concept of the integer mod operation.
Altogether we see that C++ understands five integer operators:
\begin{center}
  +     addition
  –     subtraction
  *     multiplication
  /     division (or quotient)
  \%     remainder
\end{center}
These operators are \EMPHASIZE{binary}
in the sense that you need to feed two
integers to the operator in order to get a result.

FOCUS! Read the above paragraphs again.

%\input{exercises/240-0001/main.tex}

You should thank the designers of C++ that they use the operator signs
+, – , *, / which you are used to, except for a slight change in meaning
to /. Whenever possible, programming language designers try to make
programming easier by using common mathematical notation and
convention.

