\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{A few simple facts: divisors and primes}

\textbf{{Divisors and Remainders}}

Divisors and primes and prime factorizations are already covered in
elementary algebra. Here's a quick review of some facts on divisors tied
in to C++.

An integer d is a \EMPHASIZE{divisor} of another n if d divides n. In other words an integer d is a divisor of n if you can find an integer x such that mathematically:
\[dx = n\]
For instance 3 is a divisor of 60 since mathematically
\[3x = 60\]
for x = 20. On the other hand 3 is not a divisor of 61 since you cannot find an \underline{integer} x satisfying
\[3x = 61\]
Another way to think of it is that

\EMPHASIZE{d is a divisor of n if the quotient of n
by d gives a 0 remainder}

i.e.,

\EMPHASIZE{d is a divisor of n if n \% d is 0}

For the case of 61, when you divide 61 by 3 you get
\[\frac{61}{3} = 20\frac{1}{3}\]
i.e, the remainder is 1 i.e.,
\[61 \% 3 is 1 which is not 0\]
So 3 is not a divisor of 61, or 61 is not divisible by 3.

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

Of course you know that if you have something like this for positive
integer n:
\[n = dx\]
where d and x are positive integers greater than 1, you can think of that
as breaking up n into two pieces (or factors) d and x. You can continue to
do that for both d and x until you get the prime factorization of n. For
instance for n = 100,
\[100 = 4 x 25\]
This breaks up 100 into two factors 4 and 25. We now factorize 4 as
\[4 = 2x2\]
and so
\[100 = (2x2)x25\]
We are not done since we can factorize 25 as
\[25 = 5x5\]
This gives us the following factorization of 100:
\[100 = (2x2)x(5x5)\]
Since 2 and 5 are primes you can factorize them into smaller positive
integers greater than 1. So we get the prime factorization of 100:
\[100 = 2x2x5x5.\]
And you should know that the prime factorization of any positive integer
greater than 1 into primes is unique. For instance in the above example
we started with 100 and factorize 100 into 4x25. We then factorize 4 and
25 separately until we get
\[100 = 2x2x5x5.\]
If we started the factorization with 50x2 and then factorize 50 and 2
separately we will still arrive
\[50 = 2x5x5\]
and therefore
\[100 = 50x2 = (2x5x5)x2 = 2x5x5x2.\]
After you've organized the above prime factorizations in so that the terms
are ascending order we get
\[100 = 2x2x5x5\]
\[100 = 2x5x5x2 = 2x2x5x5\]
If I use exponentiation notation, I can write this:
\[100 = 2^2 x 5^2\]
Therefore no matter how you perform the prime factorization, after
rearranging the prime factors in ascending order, the prime factorization
is always the same. This fact is called the \EMPHASIZE{fundamental theorem of arithmetic} which states that ...

\EMPHASIZE{All positive integers can be expressed as a product of primes and, up to rearranging the primes, there is only one such prime factorization.}

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

Of course we just used the concept of primes. But what's a prime?

Here's a quick review. A \EMPHASIZE{prime} is a positive integer greater than 1 that is divisible only by 1 and itself. (Note that by definition 1 is not a prime.) For instance 13 is a prime. 42 on the other hand is not a prime since for instance 2 divides 42:
\[42 = 2 x 21\]

Of course 3 divides 42 as well since
\[42 = 3 x 14\]
But in order to say that 42 is not a prime, you only need to find one
positive divisor of 42 that is not 1 and not 42.

To check that 7 is a prime, you need to check
\begin{tightlist}
\li 2 is not a divisor of 7
\li is not a divisor of 7
\li is not a divisor of 7
\li is not a divisor of 7
\li is not a divisor of 7
\end{tightlist}
which is the same as checking:
\begin{tightlist}
\li The remainder when 7 divided by 2 is not 0
\li The remainder when 7 divided by 3 is not 0
\li The remainder when 7 divided by 4 is not 0
\li The remainder when 7 divided by 5 is not 0
\li The remainder when 7 divided by 6 is not 0
\end{tightlist}
which is the same as checking
\begin{tightlist}
\li 7 \% 2 is not 0
\li 7 \% 3 is not 0
\li 7 \% 4 is not 0
\li 7 \% 5 is not 0
\li 7 \% 6 is not 0
\end{tightlist}
Make sure you follow the above sequence of translations of prime
checking down to a sequence of % operator checks!!!

%\input{exercises/240-0001/main.tex}
To check if 11 is a prime, you will need to
\begin{tightlist}
\li 2 is not a divisor of 11
\li 3 is not a divisor of 11
\li 4 is not a divisor of 11
\li 5 is not a divisor of 11
\li 6 is not a divisor of 11
\li 7 is not a divisor of 11
\li 8 is not a divisor of 11
\li 9 is not a divisor of 11
\li 10 is not a divisor of 11
\end{tightlist}
Again, you're trying to hunt down a potential divisor of 11 between 2 and
10 (1 less than 11).

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

Now suppose we attempt to check if 35 is a prime, we would do the
same:
\begin{tightlist}
\li 35 \% 2 is $\textunderscore$ therefore 2 $\textunderscore$ (is or is not) a divisor of 35
\li 35 \% 3 is $\textunderscore$ therefore 3 $\textunderscore$ (is or is not) a divisor of 35
\li 35 \% 4 is $\textunderscore$ therefore 4 $\textunderscore$ (is or is not) a divisor of 35
\li 35 \% 5 is $\textunderscore$ therefore 5 $\textunderscore$ (is or is not) a divisor of 35
\li 35 \% 6 is $\textunderscore$ therefore 6 $\textunderscore$ (is or is not) a divisor of 35
\li 35 \% 7 is $\textunderscore$ therefore 7 $\textunderscore$ (is or is not) a divisor of 35
\li etc.
\end{tightlist}
WAIT!!!! HOLD YOUR HORSES!!!

When we were testing integer division by 5, you see that 35 % 5 is 0.
This means that 5 is a divisor of 35. That means that 35 is (already) not
a prime. There's no point in continuing the checks!!! I should have
stopped then!!! Why waste time?!?


%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

