\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{A few simple facts: 10}

There's something special about integer division by 10 and mod 10.


Try this:
\begin{console}
std::cout << "1358 / 10 = " << 1358 / 10 <<std::endl;
\end{console}


You notice that 1358 / 10 is just 1358 with its rightmost digit chopped off,
i.e., it's 135.


That makes sense right? Mathematically 1358 divided by 10 gives 135.8.
So in C++, since you only get the integer part of the division, the integer
division 1358 / 10 is 135.


%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

The above exercises are important: Extracting data from data is very
common in Computer Science (well ... it's important in every area of
Science). For instance google extracts keywords from web pages for
indexing so that when someone does search for that word, google can
return the web page's URL quickly. But that's slightly different from our
case since google extracts data from strings and not integers.


The above involves \underline{cutting out} digits of chunks of digits from an integer
value. You can \underline{build} integer values using chunks of integers. For
instance you know from your math classes that
\[1358\]
is the same as (using mathematical notation):
\[1\times1000 + 3\times100 + 5\times10 + 8\times1\]
In this case, you are constructing 1358 using the digits 1, 3, 5 and 8
together with powers of 10. (Don't forget that 1 is a power of 10 too ...
10$^0$ = 1. Correct?)

%\input{exercises/240-0001/main.tex}

Note that multiplying a number with a power of 10 is the same as adding
a certain number of zeroes to the left of the given number. For instance
in the above,
\[3\times100 = 300\]
i.e., multiplication by 100 just adds 2 zeroes to the given number. Here's
another example:
\[8 \times 10000 = 80000\]
This works not just for digits (of course). For instance
\[867 \times 100000 = 86700000\]

%\input{e\timesercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}
%\input{exercises/240-0001/main.tex}

%\input{exercises/240-0001/main.tex}

