\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{The problem}

Instead of working with classes (and their objects) alone, we can work
with a hierarchy of classes. Why do we want to do this? To improve
``re-use''.

For instance: Suppose you have an \texttt{Employee} class and you want to create a \texttt{Manager} class. A manager \EMPHASIZE{is an} employee -- everything in the \texttt{Employee} class should be in the \texttt{Manager} class. For instance an \texttt{Employee} might have a last name, \texttt{Manager} should have a last name too.

You basically want the \texttt{Manager} class to have all the member variables and methods of the \texttt{Employee} class. Of course you can copy-and-paste the code from the \texttt{Employee} class into the \texttt{Manager} class:

\begin{consolethree}[escapeinside=||]
class Employee
{
public:
        ...
        |std::string get\_firstname();                \tikzmarknode{emplinestart}{}|                |\tikzmarknode{emplinebendST}{}|
        void set_firstname(const std::string &);
private:
        std::string firstname;
        ...
};
|\sidebox{empcopybox}{Copy-and-paste??}|
class Manager
{
public:
        ...
        |\tikzmarknode{emplineend}{std::string get\_firstname();}|                 |\tikzmarknode{emplinebendED}||
        
        void set_firstname(const std::string &);
        // other stuff not from Employee class
        ...
private:
        std::string firstname;
        // other stuff not from Employee class
        ...
}; 
\end{consolethree}
\DrawArrowPoints[red,thick]{emplinestart}{emplineend}{-- ++(2,0) -- ++(2,0) --++ (0,-4.6)}
Of course you would expect the \texttt{Manager} class to contain more
things than the \texttt{Employee} class. For instance in the company where
the above software is written, a manager might have a secretary while
non-managers do not.

We don't like to duplicate code. If you need to modify the code from the
\texttt{Employee} class, then you also need to modify the same code in the
\texttt{Manager} class. And if code in the \texttt{Employee} class is used
in 10 other classes, the maintenance work on code would be horrific.

What should we do?

