\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Inheritance}

The solution is to use \EMPHASIZE{inheritance}!!!

The general idea is very simple: Basically C++ allows you to define a
class so that it will have all the ``features'' of another.

That's all. BUT ... there is \EMPHASIZE{a lot} of details beyond the
general idea of inheritance. So \EMPHASIZE{pay attention}.

Here is the basic syntax to get a class to inherit ``features'' from
another class:
\begin{console}[commandchars=\~\@\$]
class Employee
{ ... };
class Manager: ~EMPHASIZE@public Employee$
{ ... }; 
\end{console}

This is called \EMPHASIZE{public inheritance}, i.e., the \texttt{Manager}
class \EMPHASIZE{inherits} everything (i.e. member variables and methods)
of the \texttt{Employee} class in the sense that the \texttt{Manager} class
will have all the features (member variables and methods) from the
\texttt{Employee} class.

Here is some basic terminology that you must know:

\begin{itemize}
\item
  \texttt{Employee} is the \EMPHASIZE{parent class} or the
  \EMPHASIZE{superclass} and \texttt{Manager} is the \EMPHASIZE{child class} or
  the \EMPHASIZE{subclass}.
\item
  \texttt{Manager} is \EMPHASIZE{derived} from \texttt{Employee} or
  \texttt{Manager} is a \EMPHASIZE{derived subclass} of \texttt{Employee}.
\end{itemize}

I'm going to use the following very simple example to illustrate
important principles when working with inheritance.

\begin{itemize}
\item
  The parent class \texttt{P} has the integer member \verb!x_! together
  with \verb!get_x! and \verb!set_x! methods.
\item
  The child class \texttt{C} has an integer member \verb!y_! together
  with \verb!get_y! and \verb!set_y! methods.
\end{itemize}

I don't have to tell you ... run it.

\begin{consolethree}[escapeinside=||]
#include <iostream>

class P
{
|\sidebox{methodbox}{\textbf{I'm putting the methods in one line so that we don't have to page up and down.}}|
public:
        P() : x_(1) {}
        int get_x() const { return x_; }
        void set_x(int x) { x_ = x; }
private:
        int x_;
};

class C : public P
{
public:
        C() : y_(2) {}
        int get_y() const { return y_; }
        void set_y(int y) { y_ = y; }
private:
        int y_;
};

int main()
{   
    C c;
    std::cout << "c.x_ =" << c.get_x() << ", "
              << "c.y_ =" << c.get_y() << '\n';
return 0;

} 
\end{consolethree}

Notice that the \texttt{c} object of class \texttt{C} inherits \verb!x_!,
\verb!get_x!, and \verb!set_x! from \texttt{P}. That's why you can
execute \verb!c.get_x()! even though class \texttt{C} does not have a
\verb!get_x()! method.

Make sure you keep the above code handy since I'll be using this again
and again in this set of notes. I'm going to call the above program our
``Basic Example''.

\input{../exercises/0-80-inheritance/question.tex}
\input{../exercises/1-80-inheritance/question.tex}
\input{../exercises/2-80-inheritance/question.tex}
So \verb!C! seems to have a ``copy'' of \verb!P!'s
declaration in it. Is that so? There are lots of other questions:

\begin{itemize}
\item
  What about the private section of \verb!P!? Can \verb!C! define
  methods to access the private members of \verb!P!? If inheritance is a
  message to the compiler to ``copy-and-paste'' code from \verb!P! to
  \verb!C!, then surely methods in \verb!C! can access the private
  members of \verb!P!. Is that so?
\item
  Also, what if \verb!C! uses some features of \verb!P! but not all --
  and in fact some features need to be modified. What can we do?
\item
  What about constructor? Clearly \verb!P::P()! constructor was called
  since in the above experiment \verb!x_! is clearly 1. Can we call a
  parent constructor explicitly so that we can initialize the
  \verb!x_!?
\item
  What about object destruction?
\item
  Etc., etc., etc.!!!
\end{itemize}

\input{../exercises/3-80-inheritance/question.tex}
