\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Constructor}

Now add some print statements in the parent and child constructor:
\begin{console}[commandchars=\~\@\$]
#include <iostream>

class P
{
public:
        P() : x_(1) { ~textbf@std::cout << "P::P()\n"; $}
        int get_x() const { return x_; }
        void set_x(int x) { x_ = x; }
private:
        int x_;
};

class C: public P
{
public:
        C() : x_(3), y_(2) { ~textbf@std::cout << "C::C()";$ }
        int get_y() const { return y_; }       
        void set_y(int y) { y_ = y; }
private:
        int x_;
        int y_;
};

int main()
{   
    C c;
    std::cout << "c.x_ =" << c.get_x() << ", "
              << "c.y_ =" << c.get_y() << '\n';
    std::cout << "c.x_ =" << c.P::get_x() << ", "
              << "c.y_ =" << c.C::get_y() << '\n';
return 0;

}
\end{console}

Run it. You'll see that the parent constructor is indeed executed. This means that the child constructor actually called the parent's constructor.

Note that the \EMPHASIZE{parent constructor} was \EMPHASIZE{called before} the \EMPHASIZE{child constructor.} Why is that? Because during child construction, you may use data already constructed in the parent. You cannot do the opposite.

Now what if you want to choose how to initialize the parent part of the
\verb!c! object? How do you explicitly call the parent's constructor?
Let's add a new constructor in \verb!P!, one that accepts an integer
value and get the \verb!C!'s constructor to call \verb!P::P(int):!
\begin{consolethree}[escapeinside=||]
#include <iostream>
|\sidebox{parentctor}{\textbf{Call parent's constructor explicitly in the initializer list}}|
class P
{
public:
        P() : x_(1) { std::cout << "P::P()\n"; }
        |\textbf{P(int x) : x\_(x) \{ std::cout << "P::P(int)"}||\verb!\n"; }!|
        int get_x() const { return x_; }
        void set_x(int x) { x_ = x; }
private:
        int x_;
};

class C: public P
{
public:
        C() : |\tikzmarknode{end42}{\EMPHASIZE{P(42)}}|, x_(3), y_(2)
                     { std::cout << "C::C()"; }
        int get_y() const { return y_; }
        void set_y(int y) { y_ = y; }
private:
        int x_;
        int y_;
};

int main()
{   
    C c;
    std::cout << "c.x_ =" << c.get_x() << ", "
              << "c.y_ =" << c.get_y() << '\n';
    std::cout << "c.x_ =" << c.P::get_x() << ", "
              << "c.y_ =" << c.C::get_y() << '\n';
    return 0;
}
\end{consolethree}
\DrawArrow[red, thick]{parentctor.south west}{end42}

<<<<<<< HEAD
\input{../exercises/6-80-inheritance/question.tex}\input{../exercises/7-80-inheritance/question.tex}Add code so that during the initialization of a \verb!Laser! object, the
=======
\begin{ex} You are given this incomplete code. Add to the
constructor of \verb!Car! and \verb!Motorcycle! so that a \verb!Car!
object is initialized to have 4 wheels and a \verb!Motorcycle! object is
initialized to have 2 wheels.

\begin{console}
class Vehicle
{
public:
        Vehicle(int numWheels)
           : numWheels_(numWheels)
        {}
        int numWheels() const { return numWheels_; }
private:
        int numWheels_;
};

class Car: public Vehicle
{
public:
        Car(bool keyless_entry)
           : keyless_entry_(keyless_entry)
        {}
private:
        bool keyless_entry_;
};

class Motorcycle: public Vehicle
{
public:
        Motorcycle()
            :
        {}
};

int main()
{   
    Car aCar;               // aCar has 4 wheels
    Motorcycle aMotorcycle; // aMotorcycle has 2
                            // wheels
    std::cout << "aCar has "
              << aCar.numWheels() << " wheels\n";
    std::cout << "aMotorcycle has "
              << aMotorcycle.numWheels()
              << " wheels\n";
    return 0;
}
\end{console}
\sidenote{
  \begin{tikzpicture}
  \node[draw,text width=2cm,minimum height=2cm, line width=0.05cm, inner sep=0.3cm, anchor=center] (car) at (1,-2) 
       {\tikzmark{c} \normalsize\texttt{Car}};
  \node[draw,text width=2cm,minimum height=2cm, line width=0.05cm, inner sep=0.3cm, anchor=center] (mcycle) at (4,-2) 
       {\tikzmark{m} \normalsize\texttt{Motor-\\ cycle}};
  \node[draw,text width=2cm,minimum height=2cm,minimum width=2cm, line width=0.05cm, inner sep=0.3cm, anchor=base] (vehicle) at (2.5,1) 
       {\tikzmark{v} \normalsize\texttt{Vehicle}};
  \node[above=4mm of car.north] (Lstart){};
  \node[above=4mm of mcycle.north] (Lend){};
  \node[below=6mm of vehicle.south] (Lmid){};
  \draw[ultra thick]        (car.north)   -- (Lstart) -- (Lend) -- (mcycle.north);
  \draw[ultra thick, ->]        (Lmid)   -- (vehicle.south);
\end{tikzpicture}
}

\end{ex}
\begin{ex}
This is an incomplete piece of code:
\begin{console}
class GameObject
{
private:
        bool is_alive_;
        int x_;
        int y_;
};

class Laser: public GameObject
{
};

class Alien: public GameObject
{
};
\end{console}
\end{ex}
Add code so that during the initialization of a \verb!Laser! object, the
>>>>>>> parent of 3043c95 (.)
\verb!is_alive_! is set to \verb!false! and during the initialization
of an \verb!Alien! object, the \verb!is_alive_! is set to \verb!true!.

