\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Class Reuse: ``has-a'' and ``is-a''}

Do not confuse inheritance and composition. They re-use old classes in
different ways. Suppose \verb!P! is a class.

There are two ways to reuse \verb!P!:

\begin{itemize}
\item
  Using inheritance: \verb!C! inherits from \verb!P!
\item
  Using composition: \verb!C! contains a member that is of type \verb!P!
\end{itemize}

Look at the following example carefully. Make sure you understand what
is happening.
%this line goes off the page for some reason? cant figure out a quick fix atm
\begin{console}
#include <iostream>

class P
{
public:
        P() : x_(1) { std::cout << "P::P()\n"; }
        P(int x) : x_(x) { std::cout << "P::P(int)\n"; }
        ~P() { std::cout << "P::~P()\n"; }
        int get_x() const { return x_; }
        void set_x(int x) { x_ = x; }
private:
        int x_;
};

class C: public P
{
public:
        C() : P(42), x_(3), y_(2)
        { std::cout << "C::C()"; }
        ~C() { std::cout << "C::~C()\n"; }
        int get_y() const { return y_; }
        void set_y(int y) { y_ = y; }
private:
        int x_;
        int y_;
};
\end{console}

\newpage\begin{consolethree}[escapeinside=||]
...
|\textbf{class D}|
|\textbf{\{}|
|\textbf{public:}|
        |\textbf{D() : p\_(42), x\_(3), y\_(2)}|
        |\textbf{\{ std::cout << "D::D()"; \}}|
        |\textbf{~D() \{ std::cout << "D::~D()\textbackslash n"; \}} \sidebox{pobject}{D has a P object}|
        |\textbf{int get\_y() const \{ return y\_; \}}|
        |\textbf{void set\_y(int y) \{ y\_ = y; \}}|
|\textbf{private:}|
        |\tikzmarknode{enddtor}{\textbf{P p\_;}}| 
        |\textbf{int x\_;}|
        |\textbf{int y\_;}|
|\textbf{\};}|
\end{consolethree}
\DrawArrowVH[red, thick]{pobject}{enddtor}
\begin{console}[commandchars=\~\@\$]    
int main()
{   
    C c;
    std::cout << "c.x_ =" << c.get_x() << ", "
              << "c.y_ =" << c.get_y() << '\n';
    std::cout << "c.x_ =" << c.P::get_x() << ", "
              << "c.y_ =" << c.C::get_y() << '\n';
    P p;
    p = c;
    
    !textbf@D d;$

    return 0;
} 
\end{console}

Study the above carefully. The \verb!c! object has its own \verb!x_!
and \verb!y_! and the \verb!get_y! and \verb!set_y! methods. It also
has (through inheritance) its \verb!P::x_! and method \verb!P::get_x!
and \verb!P::set_x!. The object \verb!d! also has \verb!x_! and
\verb!y_! and the \verb!get_x! and \verb!set_x! and it also has
\verb!p_.x_! and \verb!p_.get_x! and \verb!p_.set_x!. In some
sense both \verb!c! and \verb!d! have the same ``features''.

\verb!c! has \verb!P!'s features by inheritance while \verb!d! has
\verb!P!'s features through composition.

If what you want in a class can be found in another class, should you
use composition or inheritance?

Look at ``is-a'' and ``has-a'' relationships between concepts.

Suppose you have a \verb!Vehicle! class. You want to build a \verb!Car!
class. Think about the relationship between the concept of a car and a
vehicle. Which one sounds right:

\begin{itemize}
\item
  A car ``is a'' vehicle or a vehicle ``is a'' car
\item
  A car ``has a'' vehicle or a vehicle ''has a'' car
\end{itemize}

Clearly this one sounds better.

\begin{itemize}
\item
  A car ``is a'' vehicle
\end{itemize}

This tells you that you should probably use inheritance:
\begin{console}
class Vehicle {...};
class Car: public Vehicle {...};
\end{console}

Suppose you have an \verb!Engine! class. You want to build a \verb!Car!
class. What is the relationship between an engine and a car? Which one
sounds right:

\begin{itemize}
\item
  An engine ``is a'' car or an engine ``is a'' car
\item
  An engine car ``has a'' vehicle or a vehicle ``has an'' engine
\end{itemize}

Clearly this is the best:
\begin{itemize}
\item
  A vehicle ``has an'' engine
\end{itemize}

This tells you that you should probably use composition:
\begin{console}
class Engine {...};
class Car {
...
private:
        Engine anEngine;
        ...
};
\end{console}

For composition, you want to create methods for methods in the instance
members (delegation):
\begin{console}
class Engine {
public:
        void start();
};

class Car {
public:
        void start() { anEngine.start(); }
private:
        Engine anEngine;
};
\end{console}

Users can then do this:
\begin{console}
Car honda_civic;
honda_civic.start();
// i.e., honda_civic.start() actually
// executes honda_civic.anEngine.start()
// i.e., honda_civic delegates the work of
// "start" to it's engine
\end{console}

Otherwise you would need this:
\begin{console}
class Engine {
public:
        void start();
};

class Car {
public:
        Engine & get_engine() { return anEngine; }
private:
        Engine anEngine;
};
\end{console}

And users would have to do this:
\begin{console}
Car honda_civic;
honda_civic.get_engine().start();
\end{console}

This is \EMPHASIZE{not} so good. Why? Because you want to \EMPHASIZE{hide lower level details (engines)} and let users of your \verb!Car! class focus on higher level details (cars) which in this case is to ``start a car'' and not ``start the engine of a car''.

In terms of re-use of code:
\begin{itemize}
\item
  Derivation/inheritance is a form of white-box re-use
\item
  Composition is a form of black-box re-use
\end{itemize}

So in summary make sure you remember this:
\begin{itemize}
\item
  ``is a'' -- inheritance
\item
  ``has a'' -- composition
\end{itemize}

