\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Why?}

OK. The above looks bizarre. But what good is it? Why not just use
\textit{avg}?

Because now even functions can be viewed as values. The following will
make things clear.

For instance suppose you have two functions \textit{f1}, \textit{f2}, both
accepting two integers and returning a \textit{double} and you have a
program that looks like where \textit{f1}, \textit{f2} are used a lot but in
two different scenarios:

\begin{consolethree}[escapeinside=||]
if (x < 0)
{
    double a = f1(x + a, b);
    double b = f1(a, b + c / 2);
    double c = a + b + f1(5.0, a + b);
    z = a + b + c + f1(1, 1);
}
else
{
    double a = f2(x + a, b);
    double b = f2(a, b + c / 2);
    double c = a + b + f2(5.0, a + b);
    z = a + b + c + f2(1, 1);
}
std::cout << z << '\n';
\end{consolethree}

Note that the computation in both case are essentially the same other
than a difference in which of \textit{f1}, \textit{f2} should be used. Now
if your programming language has the concept of function pointers like
C++, you can do this:

\begin{consolethree}[escapeinside=||]
double (*g)(int, int) = (x < 0 ? &f1 : &f2);
double a = g(x + a, b);
double b = g(a, b + c / 2);
double c = a + b + g(5.0, a + b);
z = a + b + c + g(1, 1);
std::cout << z << '\n';
\end{consolethree}

Note in particular, you can create a function that accepts functions as
parameters! If you think of functions as computing with inputs values
such as integer and doubles, then you can also think of functions as
performing computations on functions ... functions operating on
functions!!!

Are there such things in real life? Of course. Suppose you want to find
the maximum value of a function like the function x*x + x - 5 in the
interval [a, b]. You might do this:

\begin{consolethree}[escapeinside=||]
double max(double a, double b)
{
    double m = a * a + a - 5;
    for (double x = a + 0.001; x <= b; x += 0.001)
    {
        double y = x * x + x - 5;
        if (y > m)
        {
            m = y;
        }
    }
    return m;
}
\end{consolethree}

Unfortunately, the function x * x + x - 5 is ``hard-coded'' into the
above \textit{max()} function. It cannot be used to compute for instance
the maximum value of 3 * x * x * x - 5 * x * x + 12. You would have to
write another function, say \textit{max2()}, that does more or less the
same thing.

But ... ahhh ... since C/C++ allows function pointers, you can
do this:

\begin{consolethree}[escapeinside=||]
double max(double a, double b, double (*f) (double))
{
    double m = f(a);
    for (double x = a + 0.001; x < b; x += 0.001)
    {
        double y = f(x);
        if (y > m)
        {
            m = y;
        }
    }
    double y = f(b);
    if (y > m)
    {
        m = y;
    }
    return m;
}
\end{consolethree}

Now if you want to compute the maximum of 3 * x * x * x - 5 * x * x +
12 for 1<= x <= 100, all you need to do is to do this:

\begin{consolethree}[escapeinside=||]
double f(double x)
{
    return 3 * x * x * x - 5 * x * x + 12;
}
\end{consolethree}

and then call the above \textit{max} function with this \textit{f}:

\begin{consolethree}[escapeinside=||]
std::cout << max(1, 100, &f) << std::endl;
\end{consolethree}

\begin{ex}
The above \textit{max} function tests x points spaced out in
such a way that consecutive x values differ by 0.001. Write a max
function that lets you specify this gap, called \textit{dx}, where the
default value is \textit{0.001}.
\end{ex}

\begin{ex}
Write a \textit{min} function that does the obvious thing
similar to the above.
\end{ex}

\begin{ex}
Write an area function that computes the area of a function
\textit{f}. For instance \textit{area(1, 5, \&f, 0.0001)} computes the area
under the curve y = f(x) from x = 1 to x = 5 using the sum of tiny
rectangular strips of width 0.0001. Let the default width be 0.001.
\end{ex}

\begin{ex}
Look at your bubblesort function again:

\begin{consolethree}[escapeinside=||]
void bubblesort(int x[], int size);
\end{consolethree}

The problem is that it sorts either in ascending or descending order.
Note that the difference is only in comparison of values in array x. You
need to know if x[i] > x[i+1] is true. We can have
two boolean comparison functions:

\begin{consolethree}[escapeinside=||]
bool lessthan(int x, int y);
bool greaterthan(int x, int y);
\end{consolethree}

and write a \textit{bubblesort} function so that you can call

\begin{consolethree}[escapeinside=||]
bubblesort(x, size, &lessthan);
\end{consolethree}

to sort \textit{x} in ascending order and call

\begin{consolethree}[escapeinside=||]
bubblesort(x, size, &greaterthan);
\end{consolethree}

to sort \textit{x} in descending order.
\end{ex}


