%-*-latex-*-
\sectionthree{Definition of big-O}
\begin{python0}
from solutions import *; clear()
\end{python0}

You are now ready for the definition of big-O, at least graphically.
A more mathematical methods will come later.

Suppose $f(n)$ and $g(n)$ are functions (of $n$ of course).
We say that $f(n)$ is the big-O of $g(n)$ and we
write 
\[
f(n) = O(g(n))
\]
if a \textit{multiple} of $|g(n)|$ beats (i.e., is $\geq$) $|f(n)|$,
not necessarily for all $n$
but for \textit{all large values} of $n$.
Of course,
the absolutely value $| \cdot |$ is not necessary if the functions are
positive. 
For this section, our $g(n)$ will be $n^0, n^1, n^2, ...$.

This means that given $f(n)$ and $g(n)$ 
in order to say $f(n) = O(g(n))$, I need to find a $C$ and an $N$ such that
$C|g(n)|$ beats $|f(n)|$ for $n$ beyond $N$.

Let me give you some examples.

Suppose we're looking at this function
\[
f(n) = 2n + 3
\]
\input{definition_of_big_O_1.tex}
Let's compare that to
\[
g(n) = n^2
\]
Here they are in a plot:
\input{definition_of_big_O_2.tex}
You see that 
\[
f(n) \leq g(n) \text{ for $n \geq 3$}
\]
If I choose $C = 1$ and $N = 3$, then for $n \geq N = 3$, 
we have (from the graph):
\[
f(n) \leq Cg(n)
\]
So we say that 
\[
f(n) = O(g(n))
\]
Note that the choice of $C$ and $N$ is not unique.
You can also choose $C = 2, N = 10$.

Here's another example.

Suppose
\[
f(n) = 3n^2 + 5 + 10 n \sin (n)
\]
Here's the plot:
\input{definition_of_big_O_3.tex}
For this example $f(n)$ is positive so $|f(n)| = f(n)$.
Let's see if we can \textit{cap} it with 
\[
g(n) = n
\]
\input{definition_of_big_O_4.tex}
Not good.
But don't forget that if we do want to say $f(n) = O(g(n))$,
then we are allowed to use multiples of $g(n)$.
So let's try
\[
10g(n) = 10 n
\]
\input{definition_of_big_O_5.tex}

Better!!!
But just by looking at the graph,
my multiple of $g(n)$ must at least 
overcome the bump of $f(n)$ at around $n=8.5$.
Let's try 
\[
30g(n) = 30 n
\]
\input{definition_of_big_O_37.tex}

Finally let's try
\[
50g(n) = 50n
\]
\input{definition_of_big_O_6.tex}

The picture is not that clear for small $n$ values.
So let's zoom in near $n = 0$ and see how $50n$ performs
against $f(n)$:
\input{definition_of_big_O_7.tex}
Clearly from the plot for $0 \leq n \leq 1$, we see that
$50g(n)$ beats $f(n)$ after 0.1.

So from the previous two graphs
can we say that for $n \geq 1$, $50g(n)$ beats $f(n)$? ... i.e.,
can we say
\[
f(n) \leq 50 g(n) \text{ for $n \geq 1$}
\]
and conclude that $f(n) = O(g(n))$?

NO!!!

The problem is that our graphs cannot show \textit{all}
large values of $n$.
In fact when we plot for $n$ up to 20, we see that trend changes:
\input{definition_of_big_O_8.tex}

It's even more revealing when I plot up to $n = 100$:
\input{definition_of_big_O_9.tex}

Clearly \textit{no} straight line is going to dominate $f(n)$ because
it seems that the graph of $f(n)$ \textit{bends} up (with wiggles along the way).

Here's an important advice:
\[
\text{\textit{GRAPHS ARE USEFUL TOOLS BUT THEY CAN DECEIVE!!!}}
\]

Let's see more of the graph to see if the pattern of bending
upward with wiggles persists.
Let's plot up to $n = 200$:
\input{definition_of_big_O_10.tex}

So let's abandon our $g(n) = n$ altogether.

What should we use to chase $f(n)$?
Of course you know that $g(n) = n^2$ is a parabola and bends up.
But does it increase (or bends) fast enough?
If you look at 
\[
f(n) = 3n^2 + 5 + 10 n \sin (n)
\]
You see that it is made up of three functions:
$3n^2$, $5$, and $10n \sin n$.
Of course $3n^2$ is going to beat $5$.
So the growth of $3n^2 + 5$ is primarily determined by $3n^2$.
Let's look at them together in a graph:
\input{definition_of_big_O_11.tex}
Of course since we can control $f(n)$ with multiples $g(n) = n^2$,
later we just need to choose a huge multiple of $n^2$ to 
beat $3n^2 + 5$, for instance $1000000 g(n) = 1000000 n^2$.

What about $10n \sin n$?
If we plot that with $3g(n) = 3n^2$ we get:
\input{definition_of_big_O_12.tex}
To make sure that the pattern persists, I'm going to plot up
to $n = 100$:
\input{definition_of_big_O_13.tex}

At this point, we suddenly recall that the sine function
wobbles between the value of $-1$ and $1$.
Therefore $10n \sin n$ wobbles between $10n (-1)$ and $10n (+1)$,
i.e., $10n \sin n$ can be at most $10n$.
Let's check that with a plot for $n = 1$ to $n = 20$:
\input{definition_of_big_O_14.tex}

AHA!!!

This means that $3g(n) = 3n^2$ will beat $10n \sin n$ for large values of $n$.

Altogether, this means that the growth of 
\[
f(n) = 3n^2 + 5 + 10 n \sin (n)
\]
is roughly $3g(n) = 3n^2$
and therefore a higher multiple of $g(n) = n^2$ will beat
$f(n) = 3n^2 + 5 + 10 n \sin (n)$ for large values of $n$.

Here's the plot of $3g(n) = 3n^2$ and $f(n)$ for $0 \leq n \leq 10$:
\input{definition_of_big_O_15.tex}
It does seem like $f(n)$ winds itself around $3g(n) = 3n^2$.
To get a better feel for the pattern of things, I'm going 
to plot up to $n = 100$:
\input{definition_of_big_O_16.tex}
and then up to $n = 500$:
\input{definition_of_big_O_17.tex}

GRRREAT!!!

I see that 
$f(n) = 3n^2 + 5 + 10 n \sin (n)$
follows $3g(n)$ very tightly.
So to beat $f(n)$, let me try $4g(n) = 4n^2$
(... well ... I'm sure $(3.5)g(n)$ works too ... but I'll stick to $4g(n)$):
\input{definition_of_big_O_18.tex}

To see roughly when $4g(n) = 4n^2$ beats 
$f(n) = 3n^2 + 5 + 10n  \sin n$, I zoom in and plot 
the graphs for $0 \leq n \leq 30$:
\input{definition_of_big_O_19.tex}
It looks like $4g(n)$ definitely beats $f(n)$ for \textit{all} $n$
such that $n \geq 10$.
(Actually I can zoom in further and see if \lq\lq $n \geq 9$''
works as well ... but I'll just use \lq\lq $n \geq 10$''.)

That's it!!!

We can now say that 
\[
3n^2 + 5 + 10 n \sin (n) \leq 4n^2 \text{ for $n \geq 10$}
\]
Therefore, if I choose $C = 4$ and $N = 10$, I can say that
\[
|f(n)| \leq C |g(n)| \text{ for $n \geq N$}
\]
(Don't forget that $f(n)$ is positive for $n \geq 0$.)
This means that I can now say
\[
f(n) = O(g(n))
\]

Now we are ready for the formal definition of big-O:

\begin{defn}
Let $f(n)$ and $g(n)$ be functions.
We write 
\[
f(n) = O(g(n)) \tinysidebar{$O$}
\]
and say that 
\lq\lq $f(n)$ is \defone{big-O} of $g(n)$"
if we can find a $C$ and an $N$ such that 
for all $n \geq N$, we have
\[
|f(n)| \leq C|g(n)|
\]
\end{defn}

So now I'm ready to give a proper presentation of my solution to
the previous problem:

\newpage
\begin{eg}
Show graphically that if 
$f(n) = 3n^2 + 5 + 10 n \sin (n)$
and $g(n) = n^2$, then
\[
f(n) = O(g(n))
\]
\end{eg}

\textit{Solution.}
Let $N = 10$ and $C = 4$.
From the following graphs:
\input{definition_of_big_O_21.tex}
\input{definition_of_big_O_22.tex}
we see that, for $n \geq N$, we have:
\[
|3n^2 + 5 + 10n \sin(n)| \leq C |n^2|
\]
i.e.,
\[
|f(n)| \leq C|g(n)|
\]
Hence $f(n) = O(g(n)$.
\qed

Remember that technically speaking you cannot prove a mathematical
fact like $f(n) = O(g(n))$ using graphs
because a graph cannot possibly show you that a multiple of $|g(n)|$
beats $|f(n)|$ for \textit{all} sufficiently large $n$.
A graph can only show you the graphs up to some \textit{finite} $n$.
How would you know that the graph 
of $|f(n)|$ won't suddenly shoot up and overtake $4|g(n)|$ at 
$n = 1000000000000$?
Later, I'll show you how to prove big-O statements without a shadow of doubt.

Now, let me give you another example.
Suppose I change my $f(n)$ to \textit{this}:
\[
f(n) = -3n^2 + 5 - 10 n \sin (n)
\]
then when I plot
$|-3n^2 + 5 - 10 n \sin (n)|$ 
and $4|n^2|$ for $0 \leq n \leq 5$,
I get
\input{definition_of_big_O_23.tex}
and then for $0 \leq n \leq 100$:
\input{definition_of_big_O_24.tex}
In this case $n^2$ still works, i.e., 
$-3n^2 + 5 - 10 n \sin (n) = O(n^2)$.
Also, it seems that $4n^2$ breaks away from 
$|-3n^2+5-10n \sin(n)|$ around $n = 20$.
So let's plot the graphs for $0 \leq n \leq 20$:
\input{definition_of_big_O_25.tex}
Clearly for $n \geq 10$, $4|g(n)|$ beats $|f(n)|$.
So I'm going to pick $N = 10$.

Now I'm ready to present this ...

\newpage

\begin{eg}
Let $f(n) = -3n^2+5-10n \sin(n)$ and $g(n) = n^2$.
Show graphically that 
\[
f(n) = O(g(n))
\]
\end{eg}

\textit{Solution.}
From the following graphs:
\input{definition_of_big_O_26.tex}
\input{definition_of_big_O_27.tex}
we see that if we choose $C = 4$ and $N  = 20$, then for $n \geq N$,
we have
\[
|-3n^2+5-10n \sin(n)| \leq 4|n^2|
\]
i.e.,
\[
|f(n)| \leq C|g(n)|
\]
Hence $f(n) = O(g(n))$.
\qed

\newpage

It's not surprising that if 
\[
3n^2 + 5 + 10 n \sin (n) = O(n^2)
\]
then it is also true that 
\[
3n^2 + 5 + 10 n \sin (n) = O(n^3)
\]
since $n^3$ grows faster than $n^2$.
So there are many possible functions to \lq\lq control''
$3n^2 + 5 + 10n \sin(n)$ from above.
Also, if
\[
|3n^2 + 5 + 10 n \sin (n)| \leq C|n^2|
\]
then of course if you replace $C$ by a large value, say $C + 1423$,
then of course
\[
|3n^2 + 5 + 10 n \sin (n)| \leq (C + 1423)|n^2|
\]
is still true for $n \geq N$.
Furthermore, this is also true if you replace $N$ by a larger value
such as $N + 15313$.
Therefore the choice of $C$ and $N$ is not unique.


Here's another example.

Let
\[
f(n) = \frac{2n^5}{1 + n} \sin \frac{1000}{n} - 100
\]
Let's find some function $g(n)$ of the form $n^0, n^1, n^2, n^3$ or ...
such that $f(n) = O(g(n))$.
Let's have a few plots to get a feel for the function.

Here's the plot of $f(n)$ for $0 \leq n \leq 10$:
\input{definition_of_big_O_28.tex}
Aha ... $f(n)$ can be negative.
I'd better plot $|f(n)|$ instead of $f(n)$.
Also, clearly $g(n)$ can't be $n^0 = 1$ (i.e., no multiple of
$g(n) = 1$ is going to beat $|f(n)|$ ... right?) 
so I'm going to skip that.
I'll try higher powers.
With $g(n) = n$:
\input{definition_of_big_O_29.tex}

With $g(n) = n^2$:
\input{definition_of_big_O_30.tex}
Wow. $f(n)$ is exploding so fast that I can't even see $n^2$ bending up at
all!
With $g(n) = n^3$:
\input{definition_of_big_O_31.tex}
OK ... at least I can see $n^3$ bending up a little. 
But it's still not high enough to bound $f(n)$.

% 32 not used

With $g(n) = n^4$:
\input{definition_of_big_O_33.tex}
Aha! $n^4$ beats $f(n)$!!!

%Here's the plot for $|f(n)|$ and $g(n) = n^4$ for $0 \leq n \leq 10$:
%\input{definition_of_big_O_33.tex}

You notice that $g(n)$ beats $|f(n)|$ around 
$5 \leq n \leq 6.5$.
Let me zoom in.
Here's the plot for $3 \leq n \leq 7$:
\input{definition_of_big_O_34.tex}

From the graphs we see that for $n \geq 6$
\[
\left| 
\frac{n^5}{1 + n} \sin \frac{1000}{n} - 100
\right| 
\leq 
\left|
n^4
\right|
\]
So if I choose $C = 1$ and $N = 6$, then for $n \geq N$,
\[
\left|
\frac{n^5}{1 + n} \sin \frac{1000}{n} - 100
\right| 
\leq C
\left|
n^4
\right|
\]
i.e.,
\[
|f(n)| \leq C|n^4|
\]
Hence
\[
f(n) = O(n^4)
\]

So here's the solution ...

\newpage

\begin{eg}
Let 
\[ 
f(n) = \frac{n^5}{1 + n} \sin \frac{1000}{n} - 100
\]
Find the smallest integer $k$ such that for $g(n) = n^k$, we have
\[
f(n) = O(g(n))
\]
\end{eg}

\textit{Solution.}
The following are plots of $|f(n)|$ and $g(n) = n^4$:
\input{definition_of_big_O_35.tex}
\input{definition_of_big_O_36.tex}
If we choose $N = 6$ and $C = 1$,
we see that for $n \geq N$,
\[
\left|
\frac{n^5}{1 + n} \sin \frac{1000}{n} - 100
\right| 
\leq C
\left|
n^4
\right|
\]
i.e.,
\[
|f(n)| \leq C|g(n)|
\]
Hence
\[
f(n) = O(n^4)
\]
\qed

\newpage\input{exercises/7n4-20/main.tex}
\newpage\input{exercises/0-31415n4-1000n3/main.tex}
\newpage\input{exercises/n4-1234n3/main.tex}
\newpage\input{exercises/3n3-5-42-n5-n2-1//main.tex}
\newpage\input{exercises/20-n5-n3-1-5n3-cos-n-7-8/main.tex}
\newpage\input{exercises/123456789--1n-n3/main.tex}
\newpage\input{exercises/7n2-7-lg-n-20-n5-n-1/main.tex}
\newpage\input{exercises/big-O-0/main.tex}
\newpage\input{exercises/big-O-1/main.tex}
