\begin{python0}
from solutions import *; clear()
\end{python0}
\input{exercises/12-11-for-loops-part2/main.tex}\input{exercises/13-11-for-loops-part2/main.tex}\sectionthree{break}

Remember \texttt{break} from the \texttt{switch} statement? If you execute
\texttt{break}, you immediately go out of the \texttt{switch} block.
\texttt{break} can also be used in a \texttt{for}-loop.
\begin{console}
First run this:
for (int i = 0; i < 1000; i += 2)
{
std::cout << "entered body of for-loop \n";
std::cout << i << std::endl;
std::cout << "exiting body of for-loop \n";
} 
\end{console}
Now run this:
\begin{console}[commandchars=\~\@\$]
for (int i = 0; i < 1000; i += 2)
{   
    std::cout << "entered body of for-loop\n";
    std::cout << i << std::endl;
    ~textbf@if (i == 100) break;$
    std::cout << "exiting body of for-loop\n";
}
\end{console}
See the point of \texttt{break}?

The \texttt{break} when executed in a \texttt{for}-loop will get out of that \texttt{for}-loop. That's all there is to it.

Of course some times you can avoid the break. For the above program you
can rewrite it as
\begin{console}
for (int i = 0; i < 101; i += 2)
{   
    std::cout << i << std::endl;
}
\end{console}
There's a general principle (or good practice) that you
should \EMPHASIZE{avoid having too many exit points out of a loop}. Too
many exit points make the code difficult to trace.
\begin{python}
from latextool_basic import *
from latexcircuit import *
p = Plot()
#p += Grid(x0=-4,y0=-4,x1=5,y1=5)
p += Rect(-3,-2,5,2, linewidth=0.2)
p += Rect(-1,-1,3,1, linewidth=0.1)
p += Rect(-0.5,-0.5,2.5,0.5, linewidth=0.05, label=r'\texttt{\textbf{break;}}')


p += Line(points=[(6,3),(3,0.8)], linewidth=0.1, endstyle='>', linestyle='dashed')
p += Line(points=[(2.5,0),(3.5, 0)], linewidth=0.1, linecolor='red')
p += Line(points=[(3.5,0),(3.5,-1.5)], linewidth=0.1, linecolor='red')
p += Line(points=[(3.5,-1.5),(5,-1.5)], linewidth=0.1, linecolor='red', endstyle='>')
p += Line(points=[(5,-1.5),(6,-1.5)], linewidth=0.1, linecolor='red', endstyle='>')
p += Line(points=[(-4.5,1.5),(-3,1.5)], linewidth=0.1, linecolor='red', endstyle='>')

X = POINT(x=6, y=3, r=0, label='body of the looping structure', anchor='south')
p += str(X)

print(p)
\end{python}
It's very important to note that the execution of break
only exits the current for-loop. Make sure you trace and run try this:
\begin{console}
for (int i = 0; i < 5; i++)
{   
    std::cout << "A: " << i << std::endl;
    for (int j = 0; j < 5; j++)
    {   
        std::cout << "B: " << j << std::endl;
        if (j == 1) break;
        std::cout << "C: " << j << std::endl;
    }
    std::cout << "D: " << i << std::endl;
}
\end{console}
