\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Unrolling the \texttt{for}-loop: How to see the \texttt{for}-loop}

Sometimes it's hard to come up with the appropriate for-loop. In that case it's always easier to \EMPHASIZE{``unroll''} the for-loop for several special cases by hand. Unrolling a for-loop is just writing down what the program executes without using the for-loop. For instance, unrolling the following:
\begin{console}
for (int i = 0; i < 3; i++)
{   
    std::cout << i << std::endl;
}

gives us
\begin{console}
std::cout << 0 << std::endl;
std::cout << 1 << std::endl;
std::cout << 2 << std::endl;
\end{console}
For instance, suppose you want the program to draw this:
\begin{console}
*
**
***
****
\end{console}
if the user enters 4 (this is a triangle). And if he enters 3, the
program draws:
\begin{console}
*
**
***
\end{console}
So just take a special case, say when the user enters 3, and write the
pseudocode \EMPHASIZE{without the for-loop}:
\begin{console}
draw a line of length 1
draw a line of length 2
draw a line of length 3
\end{console}
The pseudocode for height 4 is
\begin{console}
draw a line of length 1
draw a line of length 2
draw a line of length 3
draw a line of length 4
\end{console}
Of course you want to write one program and not two! The program must
work for both cases. We must find a way to ``combine''them.

The first pseudocode ... \EMPHASIZE{NOT C++!!!} ... can be written:
\begin{console}
for i = 1, 2, 3:
           draw a line of length i
\end{console}
and the second can be written:
\begin{console}
for i = 1, 2, 3, 4:
           draw a line of length i
\end{console}

Now they are almost the same! If the height entered by the user is kept
in variable h, then the pseudocode is
\begin{console}
for i = 1, 2, ..., h:
           draw a line of length i
\end{console}
AHA! \EMPHASIZE{This pseudocode works for both!!!}

Of course you already know how to draw a line (of stars) of length i. So
altogether the pseudocode becomes:
\begin{console}
for i = 1, 2, ..., h:
         for j = 1, 2, ..., i:
                 draw a star
         go to the next line
\end{console}

And finally a straightforward translation to C++ gives:
\begin{console}
for (int i = 1; i <= h; i++)
{   
    for (int j = 1; j <= i; j++)
    {
        std::cout << '*';
    }
    std::cout << std::endl;
}
\end{console}
Now we add the icing by prompting the user for h:
\begin{console}
int h = 0;
std::cin >> h;
for (int i = 1; i <= h; i++)
{   
    for (int j = 1; j <= i; j++)
    {   
        std::cout << '*';
    }
std::cout << std::endl;
}
\end{console}

