%-*-latex-*-
\sectionthree{Separating polynomial functions}
\begin{python0}
from solutions import *; clear()
\end{python0}

I'll be giving you the formal definition of big-O soon.
But before that, I'm going to motivate the (formal) definition of big-O
by talking about the way the graphs of polynomial climb.
The rate at which they climb essentially tells you the story of big-O among 
polynomials.
This will give you the intuitive idea behind big-O before I hit you
with the formal definition.

In this section, I will show you that when you plot polynomial functions,
they bunch up into groups.
These groups are very well-defined and simple:
they are determined by the \textit{degree} of polynomials.

Look at this mess of 15 polynomial functions 
(I won't give you the polynomials just yet):
\input{asymptotics/separating_polynomial_functions_1.tex}

This looks like wires from a behind a rack of servers.
Now if I increase the domain up to $n = 10$, we see:
\input{asymptotics/separating_polynomial_functions_2.tex}

Notice that the growth behavior of these functions are now clearer.
Now if I increase the domain to $0 \leq n \leq 100$, we see:
\input{asymptotics/separating_polynomial_functions_3.tex}

Five functions have separated away from the others.
Now I reveal to you that these five functions are:
\begin{align*}
&n^4 \\
&n^4 + n^2 - 20n - 9 \\
&n^4 +  n^3 - 25n - 1 \\
&n^4 - 15n^2 - 5n + 1 \\
&n^4 - 2n^2 - 7n - 100
\end{align*}

Now I'm going to remove these 5 functions and plot the remaining 10:
\input{asymptotics/separating_polynomial_functions_4.tex}

The five functions which are higher up are:
\begin{align*}
&n^3 \\
&n^3 + 3n^2 - 20n - 5 \\
&n^3 +   n^2 - 25n - 15 \\
&n^3 - 15n^2 - 5n + 1 \\
&n^3 - 7n^2 + 5n - 100
\end{align*}
and the last group of five functions are:
\begin{align*}
&n^2 \\
&n^2 + 2 \\
&n^2 + 1 \\
&n^2 + 5n - 5 \\
&n^2 - 3n - 8
\end{align*}

As you can see, in terms of growth, for large values of $n$, the 15 functions
\begin{align*}
&n^4 \\
&n^4 + n^2 - 20n - 9 \\
&n^4 +  n^3 - 25n - 1 \\
&n^4 - 15x^2 - 5n + 1 \\
&n^4 - 2x^2 - 7m - 100 \\
&n^3 \\
&n^3 + 3n^2 - 20n - 5 \\
&n^3 +   n^2 - 25n - 15 \\
&n^3 - 15n^2 - 5n + 1 \\
&n^3 - 7n^2 + 5n - 100 \\
&n^2 \\
&n^2 + 2 \\
&n^2 + 1 \\
&n^2 + 5n - 5 \\
&n^2 - 3n - 8
\end{align*}
bunches themselves up into 3 groups determined by their degrees.
You can think of the following as leaders in the three groups:
\begin{align*}
&n^4 \\
&n^3 \\
&n^2
\end{align*}
(because they are the simplest.)

In general \textit{all} polynomial functions with 1 for the leading coefficient -- such polynomials
are said to be \defterm{monic} polynomials -- 
group themselves up into
bunches led by the following leaders:
\[
1, \,\,\,\,\,
n, \,\,\,\,\,
n^2, \,\,\,\,\,
n^3, \,\,\,\,\,
n^4, \,\,\,\,\,
n^5, \,\,\,\,\,
n^6, \,\,\,\,\,
\ldots
\]
The bunching up for large $n$ is due to the fact that they grow (or climb) at the same rate for large $n$.
This means that the function
\[
n^2 - 42n + 691
\]
has the same growth rate as $n^2$ for large $n$.
Graphically, this means that when you zoom out (i.e., when you draw their graph for a
large domain), the graphs collapse into one.
Intuitively, you can think of it this way:
\[
\text{$n^2 - 42n + 691$ \lq\lq roughly $=$" $n^2$ \,\,\,\,\ for large $n$}
\]

Now let's get back to big-O.
Whereas the above examples talked about 
\[
\text{... \lq\lq roughly $=$'' ... \,\,\,\,\, for large $n$}
\]
big-O is more like
\[
\text{... \lq\lq roughly $\leq$'' ... \,\,\,\,\, for large $n$}
\]
Graphically, if the graph of $f(n)$ is \textit{below} the graph to $g(n)$
\textit{for large $n$}, then we can say
\[
f(n) = O(g(n))
\]
Now, one of the above 15 functions is this:
\[
n^3 + 3n^2 - 20n - 5
\]
We have already seen that the graph of 
$n^3 + 3n^2 - 20n - 5$
is the same as the graph of $n^3$ for large $n$. 
I can say
\[
n^3 + 3n^2 - 20n - 5 = O(n^3)
\]
\textit{But, there's more.}
The graph of $n^3 + 3n^2 - 20n - 5$ is roughly the graph of $n^3$
(for large $n$) and 
is of course the graph of $n^3$ is below the graph of $n^4$.
Therefore the graph of 
$n^3 + 3n^2 - 20n - 5$
is roughly below the graph of $n^4$ (for large $n$).
Therefore I can also say
\[
n^3 + 3n^2 - 20n - 5 = O(n^4)
\]
Altogether I have
\begin{align*}
n^3 + 3n^2 - 20n - 5 &= O(n^3) \\
n^3 + 3n^2 - 20n - 5 &= O(n^4)
\end{align*}
It is also true that
\begin{align*}
n^3 + 3n^2 - 20n - 5 &= O(n^3 + 1) \\
n^3 + 3n^2 - 20n - 5 &= O(n^4 + 1)
\end{align*}

OK, let's try another function.
Here's another function from the 15:
\[
n^2 + 5n - 5
\]
Using the same reasoning we have all the following:
\begin{align*}
n^2 + 5n - 5 &= O(n^2) \\
n^2 + 5n - 5 &= O(n^3) \\
n^2 + 5n - 5 &= O(n^4)
\end{align*}


But there's a little bit more to big-O.
What about multiples of the above functions?
Recall that in the previous section,
I said that you should ignore multiples by replacing constants with 1.
It seems to mean that constants don't determine function growth rate.
Is that true?

Here are the original 15 functions again:
\begin{align*}
&n^4 \\
&n^4 + n^2 - 20n - 9 \\
&n^4 +  n^3 - 25n - 1 \\
&n^4 - 15x^2 - 5n + 1 \\
&n^4 - 2x^2 - 7m - 100 \\
&n^3 \\
&n^3 + 3n^2 - 20n - 5 \\
&n^3 +   n^2 - 25n - 15 \\
&n^3 - 15n^2 - 5n + 1 \\
&n^3 - 7n^2 + 5n - 100 \\
&n^2 \\
&n^2 + 2 \\
&n^2 + 1 \\
&n^2 + 5n - 5 \\
&n^2 - 3n - 8
\end{align*}
and now I'm going change the leading coefficients
\begin{align*}
&n^4 \\
&2n^4 + n^2 - 20n - 9 \\
&3n^4 +  n^3 - 25n - 1 \\
&4n^4 - 15x^2 - 5n + 1 \\
&5n^4 - 2x^2 - 7m - 100 \\
&6n^3 \\
&7n^3 + 3n^2 - 20n - 5 \\
&8n^3 +   n^2 - 25n - 15 \\
&9n^3 - 15n^2 - 5n + 1 \\
&10n^3 - 7n^2 + 5n - 100 \\
&11n^2 \\
&12n^2 + 2 \\
&13n^2 + 1 \\
&14n^2 + 5n - 5 \\
&15n^2 - 3n - 8
\end{align*}
and then plot the new functions on the domain $0 \leq n \leq 100$:
\input{asymptotics/separating_polynomial_functions_4.tex}

The graphs have shifted vertically but the grouping is still somewhat visible.
(Of course since the polynomials in each group differ by 
multiples you would expect their graphs to separate a little.)
Regardless of the shifts, you would notice one crucial thing:
If you plot a large enough domain, regardless of the multiple,
a degree 3 polynomial will not beat a degree 4 polynomial.

Graphically, if you plot monic polynomials for a large enough domain,
the polynomials bunches up and each bunch ultimately becomes a thin line.
If you plot polynomials in general (not necessarily monic), then
each group of polynomials occupy sort of a band.
This means that a degree 3 polynomial cannot enter the band for the degree 4 
polynomials for large $n$.

So here's the definition of big-O if I use graphs.
In order to say
\[
f(n) = O(g(n))
\]
I have to show that the graph of $f(n)$ is below a multiple of $g(n)$
for large $n$.
I'll give you more examples in the next section together with
the formal definition of big-O that does not depends on graphs.

The following summarizes what I have just said.
I will prove the statement later.

\begin{thm}
Let $f(n)$ be a polynomial of degree $d$ and $g(n)$ be a polynomial
of degree $e$.
If $d \leq e$, then
\[
f(n) = O(g(n))
\]
In particular
\[
f(n) = O(n^d)
\]
If $d > e$, then
\[
f(n) \neq O(g(n))
\]
\qed
\end{thm}

\begin{eg}
  Here are some examples that uses the above theorem.
  \begin{enumerate}[nosep,label=(\alph*)]
  \item $3n^3 + n + 1 = O(n^3)$
  \item $3n^3 + n + 1 = O(n^4)$  
  \item $3n^3 + n + 1 = O(n^{100})$
  \item $3n^3 + n + 1 = O(2n^3 + n^2 + n - 1)$
  \item $3n^3 + 1 \neq O(n^2)$
  \item $3n^3 + 1 \neq O(n + 10000)$
  \end{enumerate}
\end{eg}

Usually we will pick the simplest and \lq\lq smallest" $g(n)$
for our big-O statement in
\[
f(n) = O(g(n))
\]
For instance if it is true that
\[
f(n) = O(n^3), \,\,\,
f(n) = O(3n^3 - 10n + 1), \,\,\,
f(n) = O(n^{1000})
\]
then we prefer
\[
f(n) = O(n^3)
\]

\newpage\input{asymptotics/exercises/big-O-of-functions/main.tex}
