\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Returning nothing}

You can write a function that does not return a value:

\begin{consolethree}[escapeinside=||]
#include <iostream>

void printHelloWorld(int x)
{
    std::cout << "Hello, World!" << std::endl
              << "Function received: "
              << x
              << std::endl;
    return;
}

int main()
{
    printHelloWorld(3);

    return 0;
}
\end{consolethree}

Question: What is the return type of a function that does not return a value? (Look at the code above.)

\begin{ex}
Write a function \texttt{printYesNo()} that accepts a boolean value and prints yes if the value is \texttt{true} and no otherwise. Test your code. [Hint: You can use the \texttt{printHelloWorld()} function above. The parameter \texttt{x} is an \texttt{int}. You need to change it to a \texttt{bool}.]
\end{ex}

\begin{ex}
What's wrong here?

\begin{consolethree}[escapeinside=||]
void f(int x)
{
    return x * x;
}

int main()
{
    int y = f(5);
    return 0;
}
\end{consolethree}
\end{ex}

\begin{ex}
What is the output (without running the program)?

\begin{consolethree}[escapeinside=||]
void increment(int x)
{
    x++;
    std::cout << x << std:::endl;
    return;
}

int main()
{
    int x = 1;
    increment(x);
    std::cout << x << std::endl;

    return 0;
}
\end{consolethree}

Now verify by running the program. Correct the program using this:

\begin{consolethree}[escapeinside=||]
____________ increment(int x)
{
    x++;
    return ____________;
}

int main()
{
    int x = 1;
    x = increment(x);
    std::cout << x << std::endl;
    return 0;
}
\end{consolethree}
\end{ex}

\begin{ex}
Invent your own ASCII art problem (or take the code from earlier notes or assignments and write a function for it) and write a function for it. Here's one: Write the function \texttt{drawSquare()} that accepts an int value and draw a box with height and width of the specified \texttt{int} length. For instance calling \texttt{drawSquare(3)} prints this to the console window:

\begin{console}
***
* *
***
\end{console}

and calling \texttt{drawSquare(8)} prints this:

\begin{console}
********
*      *
*      *
*      *
*      *
*      *
*      *
********
\end{console}
\end{ex}


