%-*-latex-*-
\sectionthree{Roots of Asymptotics (DIY)}
\begin{python0}
from solutions import *; clear()
\end{python0}

Historically, the use of asymptotics to analyze the
growth of functions started around 1800s and was used extensively
by mathematicians studying analysis (think Calculus) and
especially in the study of analytic number theory. In fact it was
introduced by
\href{https://en.wikipedia.org/wiki/Paul_Gustav_Heinrich_Bachmann}{Paul Bachmann}
who was an analytic number theorist.

Number theory is the study of integers.
One big goal of number
theorists is to know everything about prime numbers 2, 3, 5, 7,
11,13, $\ldots$.
An example of a question a number theorist might be interested in
is this:
What is the $n$--th prime $p_n$ for positive $n$?
So $p_1 = 2$, $p_5 = 11$.
They want to know if there is a formula for $p_n$ so that for
instance they can work out very quickly what is $p_{1000000}$.

Although number theory involved the study of whole
numbers, later on techniques involving numbers
which are not integers are used to analyze integers.
Analytic number theory uses methods from analysis.
Here \lq\lq analysis" means more or less calculus,
which means analytic number theory uses real numbers.

Number theorist, especially analytic number theorists,
are interested also in the number of primes less then a fixed
number say $x$.
This is called $\pi(x)$, the prime counting function.
For instance
$\pi(1) = 0$,
$\pi(2) = 1$,
$\pi(3) = 2$,
$\pi(4) = 2$, and
$\pi(6.2) = 3$.
Here's $\pi(x)$ for $0 \leq x \leq 20$:
\input{pi_x_20.tex}

By the way I hope you see that the two
concepts $p_n$ and $\pi(x)$ are related.
Do you see that $n = \pi(p_n)$? So if for
instance I have a formula for $\pi(x)$:
\[
 \pi(x) = \ldots \text{ formula in $x$ } \ldots
\]
I just replace the $x$ in the formula by $p_n$ to get:
\[
 n = \pi(p_n) = \ldots \text{ formula in $p_n$ } \ldots
\]
This is an equation involving $n$ and $p_n$. I then try to solve
for $p_n$ to get a formula for $p_n$ in terms of $n$.

From the above graph of $\pi(x)$, you see that the graph is
not smooth.
Here's $\pi(x)$ for $0 \leq x \leq 100$:
\input{pi_x_100.tex}
Here's $\pi(x)$ for $0 \leq x \leq 100000$ compared against the function $x/\ln x$:
\input{pi_x_100000.tex}

As you can see, $\pi(x)$, which started as a very jagged and unpredictable function
becomes extremely well-behaved and almost smooth when you zoom out.
It seems that $x/\ln x$ is a tad too small.
However if you plot the same graph 
but for larger and larger values of $x$
(i.e. if you zoom out to see more and more of the graph)
you would notice that the gap narrows.

Around 1792, \href{https://en.wikipedia.org/wiki/Carl_Friedrich_Gauss}{Gauss},
when analyzing a table of primes up to 100,000, found that  
the function $f(x) = x/\ln x$ approximates $\pi(x)$ function.
However this was a numerical verification -- 
Gauss was not able to prove mathematically that his \lq\lq nice''
$f(x)$ does approximate $\pi(x)$ for all large $x$.

For a very long time, no one was able to prove that 
Gauss' $f(x)$ is very close to $\pi(x)$.
In 1850, \href{https://en.wikipedia.org/wiki/Pafnuty_Chebyshev}{Chebychev}
was able to proved that there are constants $C_1$ 
and $C_2$ such that for large values of $x$
\[
C_1 f(x) \leq \pi(x) \leq C_2 f(x)
\]
where $f(x)$ is Gauss's conjectured approximation.
By improving on proof techniques, mathematicians were
able to improve on the constants $C_1$ and $C_2$ to get
Gauss' $f(x) = x/\ln x$ closer and closer to $\pi(x)$.

Finally, in 1896,
\href{https://en.wikipedia.org/wiki/Jacques_Hadamard}{Hadamard}
and
\href{https://en.wikipedia.org/wiki/Charles_Jean_de_la_Vall%C3%A9e_Poussin}{de la Vall\'e Poussin}
  independently 
proven that the approximation $f(x) = x/\ln x$
is extremely good and tight. 
In fact:
\[
\lim_{x \rightarrow \infty} \frac{\pi(x)}{x/\ln x} = 1
\]
This is the famous Prime Number Theorem, PNT.
One can conclude from PNT that the $n$--th prime is \lq\lq roughly''
\[
n \ln n
\]

In the above statement: 
\[
\text{\lq\lq}C_1 f(x) \leq \pi(x) \leq C_2 f(x), \,\,\,\,\,
\text{ for large $x$''}
\]
the right half of the inequality:
\[
\text{\lq\lq}\pi(x) \leq C_2 f(x), \,\,\,\,\,
\text{ for large $x$''}
\]
is in fact our big-O:
\[
\pi(x) = O(f(x))
\]
The left half:
\[
\text{\lq\lq}C_1 f(x) \leq \pi(x) \,\,\,\,\,
\text{ for large $x$''}
\]
is the big-$\Omega$: $\pi(x) = \Omega(f(x))$.
And because $\pi(x)$ is big-O and big-$\Omega$ of $f(x)$,
we say $\pi(x) = \Theta(f(x))$.
The statement
\[
\lim_{x \rightarrow \infty} \frac{\pi(x)}{x \ln x} = 1
\]
is also another type of asymptotic statement.
In this case, we say that $\pi(x)$ and $x / \ln x$ are
\defone{asymptotically equivalent}
and the notation is $\pi(x) \sim x/\ln n$.

The key point to take away is that the asymptotic notations and concepts
(O, $\Omega$, $\Theta$, $\sim$) are for the comparison of function
when the $x$ is huge, which informally is the same as looking
the graphs of the functions \textit{when you zoom out}.

\href{https://en.wikipedia.org/wiki/Donald_Knuth}{D.~Knuth}
popularized the use of asymptotic
notation in computer science (around 1960s).
