\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{26. Pointers}

\textsc{Objectives}

\begin{itemize}

\item
  Declare pointers
\item
  Assign values to pointers, address-of operator, and NULL
\item
  Comparison of address values
\item
  Use pointers as function parameters modify values
\item
  Use pointers as function parameters speedup function call
\item
  Returning references and pointers
\item
  Dynamic memory manage for non-array values
\item
  Memory deallocation and preventing memory leaks
\end{itemize}

Now we come to a concept that is extremely useful and powerful in CS.

Drumroll \ldots{} \emph{pointers!!!}

Pointers are variables that can hold physical memory addresses. By pointing a pointer to any memory location in your computer, you can access the data at that point in your computer's memory. Both data and code are stored in your computer memory. Therefore with pointers you can access any data and code. This has several consequences.

\begin{itemize}
\item
  Pointers allow you to write a function that allows pass-by- reference.
  In fact your references should really be thought of as pointers under
  the hood.
\item
  In terms of data, pointers allow you to access a part of your
  program's memory called the free store or memory heap. This allows you
  to control memory usage and have better control over memory
  management, requesting memory only when you need it and releasing it
  back to the system when it's not needed.
\item
  By embedding pointer values into data, we can create very complex
  relations between data values by linking them together. This allows
  you to build data structures to model trees, graphs, etc. Graphs are
  probably one of the most important mathematical structures in CS. A
  road network is basically a graph. A computer network is basically a
  graph. Social graph that describes person-to-person relationship in
  social media is basically a graph. Etc., etc., etc., etc., etc.
  \ldots{} !!!
\item
  Since code also lives in your computer's memory, pointers can also
  point to code such as functions. With pointers, you can actually pass
  functions (technically speaking the memory location of functions) into
  functions, i.e., functions can become arguments.
\end{itemize}

\begin{itemize}
\item
  Nowadays, hardware devices are usually ``mapped'' to memory locations
  too. That means that pointers can also be used to access devices to
  perform I/O.
\item
  Etc!!! \ldots{} not to mention bizarre things like code that rewrites
  itself.
\end{itemize}

All the reasons have to do with using pointers to achieve some computational goals. Another really important reason why pointers are important is this: Pointers are not just some abstract idea of computations in the theoretical sense. Pointers are fundamental to actual computer architecture in the sense that pointers or memory addresses are understood by your hardware. For many programming languages, you cannot access pointers or memory addresses directly. Those languages will hide pointers away from programmers by providing some language feature(s) to achieve the same goal, bypassing pointers and memory addresses. This is no big deal if all you want to do is to achieve the high level computational goals. However, if you do need to dive into the guts of your computer, you will have to know pointers very well. This also means that people who are trained only in a language that does not provide means to access pointer or memory addresses will have problems doing low-level programming or will have problems fully understanding areas such as assembly language programming and computer architecture. And low level programming does happen in the real world for instance in the case of systems programming and game development.

Because the landscape of pointers is huge let me tell you what's the plan

\begin{itemize}

\item
  First I will talk about general concepts related to pointers: memory
  addresses, pointer values, pointer operations, etc. Another basic
  concept you need to know is pointer arithmetic. But I'll delay pointer
  arithmetic till the second set of notes on pointers so that we can go
  into some applications of pointers quickly
\item
  For the first application of pointers, we'll see that pointers allow
  us to directly access and manipulation data from anywhere. In
  particular, a function can access data that is in another function.
  You recall that this is exactly what references are used for. In fact
  your C++ compiler actually converts references to pointers. So
  understanding pointers is absolutely crucial to understanding
  references. And if you know pointers well, you don't even need
  references. In fact in the original C language -- the ancestor of C++
  -- there is no concept of references.
\item
  The second application is similar to the first: we'll call functions
  with pointers (we'll also look at passing in reference) and possibly
  returning pointers (we'll also look at returning reference). In this
  case, the goal is to speed up function call.
\item
  For the third application, we'll go into an area in your computer
  called the heap and perform dynamic memory management ourselves. (The
  heap is also called the free store.) This means that we can ask the
  heap to give us an \emph{int} value when we need it and give the
  \emph{int} value back to the heap when we're done with it. This is
  very different from using an \emph{int} value that belongs to an
  \emph{int} variable. For an \emph{int} variable, the value exists when
  you declare the \emph{int} variable. The value is given back to the
  computer when the variable goes out of scope. You don't have as much
  control over memory usage for your regular variables. To use the heap
  ... you must use pointers.
\end{itemize}

In the next set of notes on pointers, I'll talk about pointer arithmetic, the relationship between pointers and arrays, and dynamic memory management of (dynamic) arrays.

Here we go \ldots{} !!!

