\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{The free store or memory heap}

Our variables up to this point (except for pointer variables) have this
behavior:

\begin{itemize}

\item
  When you declare it in a block, the variable is created, the memory
  for the value is also created, and you have access to its name.
\item
  When you exit the block where the variable is declared, the variable
  name and its value are destroyed.
\end{itemize}

Note in particular that the scope and value of the variable goes hand in hand. (Remember: The scope of a variable is where you can access the variable's name.) Such variables are called \EMPHASIZE{automatic variables}. Because the memory used for their variables are destroyed and reclaimed automatically when the variable goes out of scope -- you don't have to manipulate the memory.

\begin{consolethree}[escapeinside=||]
if (i > 0)
{
    int x = 42;
    ...
}
// x dies when x goes out of scope. The memory
// used by x is reclaimed automatically by the
// the program.
\end{consolethree}
You can actually do this:

\begin{consolethree}[escapeinside=||]
auto int x = 42;
\end{consolethree}

but that's redundant \ldots C/C++ variables are automatic by default.

When you allocate memory for pointers, the values they point to are different. For instance you have already seen that
\begin{center}
\texttt{int * p = new int;}
\end{center}

creates an \texttt{int} value for \texttt{p} to point to. Note that

\begin{itemize}
\item
  \textbf{the value of }\texttt{\textbf{p}}\textbf{ (an address) is still
  automatic ... but }
\item
  \textbf{the value }\texttt{\textbf{p}}\textbf{ points to (an
  }\texttt{\textbf{int}}\textbf{) is not automatic}.
\end{itemize}

\begin{consolethree}[escapeinside=||]
if (i > 0)
{
    int * p = new int;
    ...
}

// p dies when p goes out of scope. The memory
// used by p is reclaimed automatically by the
// program. BUT the value p was pointing to is not
// reclaimed!!! 
\end{consolethree}
This is \EMPHASIZE{extremely important!!!}

This means that: When you leave the block where you declare a pointer variable \texttt{p} and called \texttt{new},

\begin{itemize}
\item
  Since \texttt{p} is automatic: the pointer \texttt{p} is destroyed and the
  memory used for the address in \texttt{p} reclaimed, but
\item
  Since *\texttt{p} is not automatic: the memory that \texttt{p} points to is not reclaimed by the computer. We'll talk about how to reclaim the memory later.
\end{itemize}

But anyway where exactly is the memory \texttt{p} points to? It's an an area called the \EMPHASIZE{free store}, also called the \EMPHASIZE{memory heap}. You basically request for memory space using the \texttt{new} command.
\begin{python}
from latextool_basic import *
from latexcircuit import POINT
p = Plot()

borderbox = Rect(0, 0, 6, 3, linewidth=0.1)
innerbox = Rect(1, 1, 4, 2, linewidth=0.05)
p += borderbox
p += innerbox

ptrbox = Rect(9, 2, 12, 3, linewidth=0.05, label='42')
p += ptrbox
p += Line(points=[(3.75, 1.5), (9, 2.5)], linewidth=0.075, linestyle='dashed', endstyle='>')

borderbox2 = Rect(7, 0, 13, 3, linewidth=0.1)
fstorelbl = Rect(8, 3.5, 8, 3.5, linecolor='white', label='Free store')

usemem = Rect(14, 2, 17, 3, linecolor='red', linewidth=0.1, innersep=0.1, label='Used memory', align='c')
unusedmem = Rect(14, 0, 17, 1, linecolor='red', linewidth=0.1, innersep=0.1, label='Used memory', align='c')
p += borderbox2
p += fstorelbl
p += usemem
p += unusedmem
p += Line(points=[(14, 2.5), (11, 2.5)], linecolor='red', linewidth=0.1, endstyle='>')
p += Line(points=[(14, 0.5), (11, 0.5)], linecolor='red', linewidth=0.1, endstyle='>')
print(p)
\end{python}

So our computational model now has the following pieces:
\begin{python}
from latextool_basic import *
p = Plot()

p += ellipse(-2, 4, 2, 7, linewidth=0.1)
p += Rect(0, 8, 0, 8, label='\\textbf{CPU}', linecolor='white')

p += Rect(3, 3, 7, 7, linewidth=0.1)
p += Rect(5, 8, 5, 8, label='\\textbf{Memory for functions}', linecolor='white')
p += Rect(3, 0, 7, 3, innersep=0.1, linecolor='white', s = 'For variables created in a function, including \\texttt{main()}', align='c')

p += Rect(8, 3, 12, 7, linewidth=0.1)
p += Rect(10, 8, 10, 8, label='\\textbf{Stack}', linecolor='white')
p += Rect(8, 0, 12, 3, innersep=0.1, linecolor='white', s = 'For values passed between functions and remembering point of return', align='c')

p += Rect(13, 3, 17, 7, linewidth=0.1)
p += Rect(15, 8, 15, 8, label='\\textbf{Free store or heap}', linecolor='white')
p += Rect(13, 0, 17, 3, innersep=0.1, linecolor='white', s = 'For values created using \\texttt{new}', align='c')

p += Rect(-2, -3, 2, -1, linewidth=0.1)
p += Rect(0, -0.5, 0, -0.5, label='\\textbf{Output}', linecolor='white')
print(p)
\end{python}
The free store will \EMPHASIZE{keep track of which part of its memory is already in use} via \texttt{new}. A piece of memory allocated to a pointer variable will be marked as used. That way, \EMPHASIZE{two calls to \texttt{new} will not give two pointers the same piece of memory}.

char * p = new char;

int * q = new int;

double * r = new double;

