\sectionthree{Some basic operations}
\begin{python0}
from solutions import *; clear()
\end{python0}

Here are some basic methods on a node.
Note that there are so many different variations of trees
that some of the methods might not apply for some cases.
So you have to understand the context.

The following are some basic queries (they don't modify the 
tree in any way):

\begin{console}
node.is_root()           true iff node is a root
node.is_leaf()           true iff node is a leaf
node.is_nonleaf()        true iff node is non-leaf

node.level()             level of node
node.height()            height of node
node.depth()             depth of node
node.size()              number of descendents + 1
\end{console}

Note that for instance in the case of \texttt{is\_root()},
the only way for this to be meaningful is when the node
has a parent pointer: a node is a root is the parent pointer
is \texttt{NULL}.
So if your node class does not have a parent pointer
member, then this method is not meaningful.

For querying ancestor information, you have the following:
\begin{console}
node.parent()            pointer/reference to parent
node.root()              pointer/reference to root 
\end{console}

The next two methods are related to siblings of a node.
This is usually only available if the siblings are
organized in an iterable manner such as for instance
if the siblings form a linked list.
\begin{console}
node.next()              pointer/reference to next 
                         sibling; NULL if there is no 
                         next sibling.
node.prev()              pointer/reference to previous 
                         sibling; NULL if there is no
                         previous sibling.
\end{console}
In the case where the siblings form a circular list, 
it's of course important that you don't run into an infinite loop!

For querying children related information, you might have the
following:
\begin{console}
node.num_children()      number of children
node.children()          pointer/reference to iterable
                         container with children
node.first_child()       first child (assuming children 
                         are ordered)
node.last_child()        pointer/reference to last child 
                         (assuming children are ordered)
node.child(i)            pointer/reference to the i-th 
                         child
\end{console}
Sometimes the following descendent queries are useful too:
\begin{console}
node.leftmost()          pointer/reference to leftmost
                         starting at node; for a binary
                         tree, keep following the left
                         pointer/reference
node.rightmost()         pointer/reference to rightmost
                         starting at node; for a binary
                         tree, keep following the right
                         pointer/reference
\end{console}

For insertion, look at this binary tree:

\input{stdout10.tex}

The most common case of insertion is as a child at a position that
is available.
For instance in the above, note that at node $l$ 
you can insert a node at the left or the right.
However note that at $h$, you can only insert at the left.
Note that the above also makes sense when the node that is
inserted is actually the root of a tree.
In the case of a $k$--ary tree, 
if child-$i$ is available, you can insert a node at that 
position.

Most tree API do not include inserting a node at a place that is not available.
Although that's entirely reasonable.
For instance there's no reason why you cannot do an insertion
right at the node $d$ with new data $o$ (i.e., it goes
between $d$ and $j$):

\input{stdout11.tex}

In this case, one has to make sure that the node $o$
has an available right position.
For instance suppose $o$ looks like this instead:

\input{stdout12.tex}

Then we can insert $o$ into the above tree (with root $a$) 
between $d$ and $j$ like this:

\input{stdout13.tex}

In the same manner, it shouldn't be too surprising that 
if we do an \lq\lq insert parent'' $o$ for $d$ in this tree:

\input{stdout14.tex}

that we get this:

\input{stdout15.tex}

The above also makes sense if node $o$ has a left child.

\begin{Verbatim}[frame=single]
node.insert_parent(node2) make node2 the parent of node
                          and node2 the child of node's
                          original parent
node.insert(i, data)      create a new node for data and 
                          attach the new node as the i-th 
                          child of node. If there's already 
                          an i-th child, an exception is 
                          thrown.
node.insert(i, node2)     similar to above
node.insert_left(data)    create a new node for data and 
                          attach the new node to the left 
                          of node (this is for binary 
                          tree). If there's already a left 
                          child, an exception is thrown.
node.insert_left(node2)   Similar to above
node.insert_right(node2)  Similar to above.
\end{Verbatim}

For removal, we can
remove a single node that is a leaf.
I'll call this \verb!remove_leaf!.
Look at this:

\input{stdout16.tex}

On removing $k$ we get

\input{stdout17.tex}

This is the usual node removal operation for a \textit{general} tree.

I want to add another node remove operation:
Suppose the node to be removed has only \textit{one} child,
then there's an obvious way to remove the node.
For instance look at

\input{stdout18.tex}

On removing $h$, we get

\input{stdout19.tex}
 
This is obviously the same even when $m$ is the root of a subtree.
Likewise, if I now remove $o$, I will get

\input{stdout20.tex}


\begin{Verbatim}[frame=single]
node.remove_if_leaf()    remove the node if it's leaf.
                         Otherwise an exception it 
                         thrown.
node.remove_if_leaf(i)   remove the i-th child if the
                         child is a leaf.
node.remove_left_leaf()  remove left child if it is a 
                         leaf; if the left child is not a 
                         leaf, an exception is thrown.
node.remove_right_leaf() remove right child if it is a 
                         leaf; if the right child is not 
                         a leaf, an exception is thrown.

\end{Verbatim}

We can also move the whole tree by specifying the root.
Of course this includes the case of removing a subtree by specifying
the root of a subtree.
