%-*-latex-*-
\sectionthree{Tree}
\begin{python0}
from solutions import *; clear()
\end{python0}

A tree is just a connected graph without cycles.
In this case, I'm talking about a tree as an undirected
graph.
You have to be careful, sometimes, a tree can also 
refer to a tree as a directed graph and acyclic.

Sometimes a special node is selected.
In this case, such a tree is sometimes called a rooted tree
(or a tree with a root).

Here's a tree (directed version):

\input{stdout00.tex}

\begin{myenum}

  \li
  $a$ is the \defone{root} of the tree.

  \li
  $a$ has 3 \defone{children}, i.e., $b,c,d$; 
  $a$ has a
  \sidebarskip{8pt}
  \defone{branching factor} 
  of 3.
  Sometimes, it's convenient to number the children.
  I'll call $b$ child $0$ of $a$; $d$ is child $2$ of $a$.

  \li
  $a$ is the \defone{parent} of $b$.

  \li
  $k$,$l$,$f$,$g$,$m$,$i$,$j$ are called
  \sidebarskip{-2pt}
  \defone{leaves}
  or
  \sidebarskip{12pt}\defone{non-internal} nodes because
  they don't have children.

  \li
  $a,b,c,d,e,h$ are called
  \sidebarskip{-2pt}
  \defone{non-leaf} nodes or 
  \sidebarskip{12pt}\defone{internal} nodes because 
  each have at least one child.

  \li
  $b$ has a \defone{depth} of $1$, i.e., 
  the length of the path from the root to $b$
  is 1. I will also say that $b$ is at level $1$.
  A root (see $a$ above) has depth 0.

  \li
  $b$ has a \defone{height} of $2$, i.e., 
  the length of the longest path from 
  $b$ to a leaf is 2.
  A leaf (see $k$ above) has height 0.

  \li
  The \defone{height} of a tree is the height of the root of the tree.
  The above tree has a height of 3.
  The height of an empty tree is -1.

  \li
  The root is at level 0. $b,c,d$ are at level 1.
  $e,f,g,h,i,j$ are at level 2. $k,l,m$ are at level 3.
  
  \li
  Two nodes $u$ and $v$ are \defone{siblings} if they have
  the same parent.
  For instance $b,c,d$ are siblings.
  
  \li
  $u$ has \defone{descendent} $v$ if there is a sequence of nodes 
  $u = v_0$, $v_1$, ..., $v_n = v$ where $v_i$ is the parent of $v_{i+1}$.
  $v$ is the \defone{ancestor} of $u$ if $u$ is a descendent of $v$.
  For instance in the above tree $b$ is an ancestor of $i$ and
  $i$ is a descendent of $l$.

\end{myenum}
Note that the above tree is a directed graph.
The following is also a tree (as in an \textit{undirected} tree).

\input{stdout01.tex}

In the case of implementation, you can think of the directed version
as having pointers from a parent to the children.
In the case of the undirected version, you can think of
each edge as allowing you to go both up and down -- 
nodes having pointers going to their children or parent.
Of course exceptions being that roots have no parents and leaves have no
children.

For each node of a tree $T$, if you consider all the descendents of the node, 
you have a tree. This is called a \defone{subtree} of $T$.
For instance

\input{stdout02.tex}

is a subtree of 

\input{stdout03.tex}

I will say that this is the subtree at $b$.

In the case of a node with two child, 
there are two subtrees, the subtree on the left is called the 
\defone{left subtree} and the one on the right is called the 
\defone{right subtree}.
(Duh.)
Look at the subtree at $b$:

\input{stdout04.tex}

The nodes $e,k,l$ form the left subtree of $b$
and the single node $f$ form the right subtree of $b$.


A tree is said to be \defone{$k$--ary} if every node has at most $k$ children;
that's the same as saying that every node has a maximum branching factor of
$k$.
In particular, a \defone{binary} tree is a tree with at most 2 children.

\input{exercises/tree-1/main.tex}

We include the empty graph as a tree, except of course such trees
don't have roots.
Consider a $k$--ary tree $T$.
\begin{myenum}

\li $T$ is \defone{full} if every node has exactly $k$ children, except for 
the leaves, i.e., if every node has
either $k$ or $0$ children.
Here's a full binary tree:

\input{stdout05.tex}

Here's another:

\input{stdout06.tex}

\li $T$ is \defone{perfect} if $T$ is full and the leaves are 
at the same level.
Here's a perfect binary tree:

\input{stdout07.tex}
 
\li $T$ is \defone{complete} if it's \lq\lq almost full'' 
except that the last level 
might not have all the nodes. 
Here's a complete binary tree that is not full:

\input{stdout08.tex}
      
\li $T$ is \defone{balanced} if at each node,
every pair of children have heights differ by 0 or 1.
\end{myenum}

\begin{ex}
If $T$ is a balanced tree, does it mean that $T$ is complete?
If $T$ is complete, does it mean that $T$ is balanced?
\qed
\end{ex}

Let $T$ be a $k$--ary tree with height $h$ and the total 
number of nodes is $n$.
At level $\ell$,
there are at most $k^\ell$ nodes.
(Don't forget that the first level that contains only the root
is called level $0$.)
In particular for a binary tree, at level $\ell$, 
there are $2^\ell$ nodes.

Therefore if the height is $h$, $T$ can have at most
\[
k^0 + k^1 + \cdots + k^h = \frac{k^{h+1} - 1}{k-1}
\]
nodes.
(Remember your geometric sum formula?)
In particular, for a binary tree, $T$ can have at most
\[
2^{h+1} - 1
\]
nodes.
Since $T$ has $n$ nodes, we must have the following relation
between $n$ and $h$ for a $k$--ary tree:
\[
h + 1 \leq n \leq \frac{k^{h+1} - 1}{k-1}
\]
And for the case of binary tree, we have
\[
h + 1 \leq n \leq 2^{h+1}
\]

