\begin{python0}
from solutions import *; clear()
\end{python0}
<<<<<<< HEAD
\input{../exercises/10-14-loops/question.tex}\sectionthree{Numerical Approximation}
=======
\sectionthree{Numerical Approximation}
>>>>>>> parent of 3043c95 (.)

Look at this sum:

\tab[6em]{\texttt{\ \ \ \ \ \ 1\ \ \ \ \ \ \ 1\ \ \ \ \ \ \ 1\ \ \ \ \ \ \ 1\ \ \ \ \ \ \ 1}}\\
\tab[6em]{\texttt{t = ----- + ----- + ----- + ----- + -----}}\\
\tab[6em]{\texttt{\ \ \ \ 1 * 1\ \ \ 2 * 2\ \ \ 3 * 3\ \ \ 4 * 4\ \ \ 5 * 5}}\\

This is a sum of 5 terms. It's easy to write a program
to compute the sum -- you have seen similar examples before.

<<<<<<< HEAD
\input{../exercises/11-14-loops/question.tex}%if you had a gun to my head and told me to figure out why this doesnt correctly tab before 1*1 but does afterwards or i die, i would pull the trigger for you
\input{../exercises/12-14-loops/question.tex}
\input{../exercises/13-14-loops/question.tex}\input{../exercises/14-14-loops/question.tex}
=======
\begin{ex} Write a program that computes t. Should you use a
for-loop or a while-loop? The value you get should be approximately
1.46361.
\end{ex}
%if you had a gun to my head and told me to figure out why this doesnt correctly tab before 1*1 but does afterwards or i die, i would pull the trigger for you
\begin{ex} Now modify the above program to compute

\texttt{\ \ \ \ \ \ 1\ \ \ \ \ \ \ 1\ \ \ \ \ \ \ 1\ \ \ \ \ \ \ 1\ \ \ \ \ \ \ 1\ \ \ \ \ \ \ \ 1}\\
\texttt{t = ----- + ----- + ----- + ----- + ----- + -----}\\
\texttt{\ \ \ \ 1 * 1\ \ \ 2 * 2\ \ \ 3 * 3\ \ \ 4 * 4\ \ \ 5 * 5\ \ \ 6 * 6}\\

The value should be 1.49139. (If you get 1 then you know
you're wrong.)
\end{ex}

\begin{ex} The first program of this section computes a sum of 5
terms. The second computes a sum of six terms. Now do the obvious:
Rewrite your program so that it prompts the user how many terms to use
in the sum. Let's say the variable for the number of
terms is n.
\end{ex}
\begin{ex} Modify your program to print the following values:
\begin{console}
  1
-----
1 * 1

  1       1
----- + -----
1 * 1   2 * 2

  1       1       1
----- + ----- + -----
1 * 1   2 * 2   3 * 3

1         1       1       1
----- + ----- + ----- + -----
1 * 1   2 * 2   3 * 3   4 * 4

  1       1       1       1       1
----- + ----- + ----- + ----- + -----
1 * 1   2 * 2   3 * 3   4 * 4   5 * 5

up to

  1       1       1              1
----- + ----- + ----- + ... + -------
1 * 1   2 * 2   3 * 3         10 * 10
\end{console}
You should get this:

\begin{console}
1
1.25
1.36111
1.42361
1.46361
1.49139
1.5118
1.52742
1.53977
1.54977 
\end{console}
\end{ex}

>>>>>>> parent of 3043c95 (.)
Now let the printout go up to a sum of 20 terms, then 30 terms, etc.
Keep going until the first 5 decimal places stops changing.

Basically, what you're doing in the previous exercises
is to compute the approximations to
\begin{console}
      1       1       1       1       1
s = ----- + ----- + ----- + ----- + ----- + ...
    1 * 1   2 * 2   3 * 3   4 * 4   5 * 5
\end{console}
Note that this goes on forever. The reason why it stops changing after a
while is that the term you're adding to the sum gets
smaller and smaller ``very quickly'' :
\tab[9em]{\texttt{
\ \ 1
-----
i * i
}}
If you want to know more, you should take Calc 2. This is not a Calc
class.

(To the math folks: If you keep going forever you get

\tab[9em]{$\frac{\pi^2}{6}$}

correct?)

Computations of approximations is important for many kinds of
applications including scientific, mathematical, and business
applications.

Now suppose you work for NASA and you need to compute the above sum
\begin{console}
      1       1       1       1       1
s = ----- + ----- + ----- + ----- + ----- + ...
    1 * 1   2 * 2   3 * 3   4 * 4   5 * 5
\end{console}
(which goes on forever) up to 10 decimal places. You can do what we did
early: Keep including more and more terms. You need to get C++ to reveal
more decimal places (use \texttt{std::setprecision()}). That would be
really time consuming.

Another way is to let the for-loop run without an upper limit for i (how
would you do that?) and when you see the first 10 decimal places not
changing you can write it down (if you can see it scrolling by) and stop
the program manually. But this might take a while and if you go out to
lunch, the for-loop keeps running even after the first 10 decimal places
stop changing. This is a waste of CPU time.

We need to figure out how to stop the loop AUTOMATICALLY.

By the way if you look at the term:\\
\tab[3em]{\texttt{
\ \ 1
-----
i * i
}}

obviously you want to avoid integer division. So in C++ you should do
\begin{center}
    \texttt{1.0 / (i * i)}
\end{center}
But besides fixing the division problem, you have another: recall that
an int value in C++ does not go on forever. If you want to represent
bigger numbers, you should do
\begin{center}
    \texttt{1.0 / (double(i) * i)}
\end{center}
i.e. even the denominator multiplication should be ``doubled'' -up.

<<<<<<< HEAD
\input{../exercises/15-14-loops/question.tex}\input{../exercises/16-14-loops/question.tex}\input{../exercises/17-14-loops/question.tex}\input{../exercises/18-14-loops/question.tex}
=======
\begin{ex}
Repeat the above example with this:
\begin{console}
      1       1       1       1       1
s = ----- + ----- + ----- + ----- + ----- + ...
    1*1*1   2*2*2   3*3*3   4*4*4   5*5*5
\end{console}
\end{ex}
\begin{ex}
Repeat the above example with this:
\begin{console}
      1       1       1 
s = ----- + ----- + ----- + ...
    1 * 1   3 * 3   5 * 5
\end{console}
\end{ex}
\begin{ex}
Repeat the above example with this:
\begin{console}
      1       1       1       1
s = ----- - ----- + ----- - ----- ...
    1 * 1   3 * 3   5 * 5   7 * 7
\end{console}
(WARNING: Note the alternating + and -- signs.)
\end{ex}
\begin{ex}
Repeat the above example with this:
\begin{console}
    1    1      1        1          1
s = - + --- + ----- + ------- + --------- + ...
    1   1*2   1*2*3   1*2*3*4   1*2*3*4*5
\end{console}
\end{ex}

>>>>>>> parent of 3043c95 (.)
