\begin{python0}
from solutions import *; clear()
\end{python0}
<<<<<<< HEAD
\input{../exercises/0-14-loops/question.tex}\input{../exercises/1-14-loops/question.tex}\sectionthree{Example: sentinel value}
=======
\sectionthree{Example: sentinel value}
>>>>>>> parent of 3043c95 (.)

Here's an easy example. Write a simple cash register program. The program prompts the cashier for the number of items and the unit price per line item (yes, it's a lousy point-of-sale program). There is a tax of 20\% (this is probably in New York).

\begin{console}
const double TAX = 0.2;
double total = 0.0;
double price = 0.0;
int num = 0;

while (1)
{     
      std::cout << "units: ";
      std::cin >> num;

      std::cout << "unit price: ";
      std::cin >> price;

      double cost = num * price * (1 + TAX);
      std::cout << "please pay " << cost
                << std::endl;
} 
\end{console}

The problem is that the program doesn't stop!!!

<<<<<<< HEAD
\input{../exercises/2-14-loops/question.tex}\input{../exercises/3-14-loops/question.tex}\input{../exercises/4-14-loops/question.tex}\input{../exercises/5-14-loops/question.tex}Of course for a real business application, the POS register would be
=======
\begin{ex}
Add code to this program so that when the cashier
enters 0 for number of items, the program gets out of the loop.
\end{ex}
\begin{ex} Add code to print the warning ``units cannot be
negative'' and \texttt{continue} when \texttt{num} is negative.
\end{ex}
\begin{ex} Add code to print the warning ``unit price cannot be
negative'' and \texttt{continue} when \texttt{price} is negative.
\end{ex}
\begin{ex} Add a running total so that the cashier can double
check the amount in the till at the end of the day.
\end{ex}
Of course for a real business application, the POS register would be
>>>>>>> parent of 3043c95 (.)
networked to a database server containing the product catalog and you
simply scan the UPC to retrieve the unit price. The data computed is
stored in a database directly. And of course you can have more than one
line item per sale.

Note that with multiple \texttt{continue}s, it makes the program harder to
read. This is similar to the situation where there are multiple breaks
in the body of the while-loop. Can you rewrite the above program so that
there are no continues?

