\begin{python0}
from solutions import *; clear()
\end{python0}
\input{temp/14-loops/exercises/6-14-loops/question.tex}\sectionthree{Error-checking for a sentinel while-loop}

The following is a structure for the while-loop that includes checks for
breaking and data validation:
\begin{align*}
\text{some initialization}\\
\text{while (1):}\\
&\text{get data}\\
&\text{if data is sentinel value:}\\
&\tab[3em]{\text{break}}\\
&\text{if data is not valid:}\\
&\tab[3em]{\text{print warning}}\\
&\tab[3em]{\text{continue}}\\
&\text{process data}\\
\end{align*}
OR
\begin{align*}
\text{some initialization}\\
\text{while (1):}\\
&\text{get data}\\
&\text{if data is sentinel value:}\\
&\tab[3em]{\text{break}}\\
&\text{if data is not valid:}\\
&\tab[3em]{\text{print warning}}\\
&\text{else:}\\
&\tab[3em]{\text{process data}}\\
\end{align*}
The second version removes the presence of \texttt{continue}. Make sure
you understand that they work the same!!! You can use either version in
your programs.

\input{temp/14-loops/exercises/7-14-loops/question.tex}
Although this is usually not mentioned in most textbooks, the above is a
very standard structure of the loop used in the industry, if you need to
use it at all. In general, alternative flow of execution must be bunched
up together in a spot which is easy to find. We won't
have time to talk too much about this topic, but in writing robust
software there are concepts such as assertions, exceptions,
pre-conditions, post-conditions, etc. which are all related to catching
or verification of error conditions. They are all related to flow of
control. These controls are also usually placed at the ``usual'' places,
near the top or the bottom of certain structures. I
won't elaborate on ``assertions, exceptions,
pre-conditions, post-conditions'' -- you will learn some of these in
future classes.

