\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Exceptions}

Exceptions are error control devices. Here's the big picture ...

Suppose you write a piece of code. Say function \texttt{f()} calls
function \texttt{g(),} and something terrible happens in \texttt{g()}.
\texttt{g()} can tell \texttt{f()} that something really bad happened in
different ways. For instance the return value of \texttt{g()} can be an
error code. Another way is to get \texttt{g()} to set a global variable to
some error code; \texttt{f()} can check this global variable after
execution has returned back to \texttt{f()}. In both cases, you control
the flow of execution to handle some error.

There's a better way. It involves \EMPHASIZE{exceptions}.

In the case of exceptions, to tell the calling function something has gone wrong, you \EMPHASIZE{throw} an object or value back to the calling function; this object/value is called the \EMPHASIZE{exception}. The calling function can choose to \EMPHASIZE{catch} the exception or not. But the interesting thing is that if \texttt{h()} calls \texttt{f()} and \texttt{f()} calls \texttt{g()} and \texttt{g()} throws an exception back, if \texttt{f()} does not catch it but \texttt{h()} does, then the flow of execution goes to the code in h() that catches the exception (or error).

That's the big picture. Now how do you try to catch the error? If \texttt{f()} wants to catch an exception, it must wrap the code where an exception might occur in a \EMPHASIZE{try-block}. The try-block is followed by a \EMPHASIZE{catch-block}. If \texttt{g()} throws an object or value of some type, say it's an integer, then your catch must specify that it's trying to catch an integer exception:

\begin{console}[commandchars=\~\@\$]
void f()
{    
     ~EMPHASIZE@try$
     ~EMPHASIZE@{$
         // code
         g();
         // more code
     ~EMPHASIZE@}$
     ~EMPHASIZE@catch (int i)$
     ~EMPHASIZE@{$
         // code to clean up the mess caused
         // by the error
     ~EMPHASIZE@}$
} 
\end{console}

What about on \texttt{g()}'s side? If something goes wrong
\texttt{g()} throws an integer back. This is achieved by:

\begin{console}
void g()
{    
     // ... some code ...
     throw 5;
     // ... some code ...
}
\end{console}

Your code can catch exceptions of different types, not necessarily an
integer. It can even be an object.

\begin{console}
void g()
{    
     std::cout << "g(): about to throw 5...\n";
     throw 5;
     std::cout << "g(): after throw 5 ...\n";

}

void f()
{    
     try
     {
           std::cout << "f(): about to call g()...\n";
           g();
           std::cout << "f(): after calling g()...\n";
     }
     catch (int i)
     {     
           std::cout << "f(): caught int " << i << '\n';
     }
}

int main()
{   
    f();
    return 0;
}
\end{console}

Run the following programs/experiments and \EMPHASIZE{trace their execution}. That is in fact the reason for exceptions: they alter the flow of normal program execution.

