\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Program 1. A Simple Example}

Run this program. Then uncomment the \texttt{throw} statement and run it
again.
\begin{console}[commandchars=\~\@\$]
#include <iostream>
#include <exception>
void f()
{    
     std::cout << "f ... 1\n";
     throw std::exception();
     
     //~EMPHASIZE@throw std::exception();$
     std::cout << "f ... 2\n";
}

int main()
{   
    ~EMPHASIZE@try$
    ~EMPHASIZE@{$   
        std::cout << "try ... 1\n";
        f();
        std::cout << "try ... 2\n";
    ~EMPHASIZE@}$
    ~EMPHASIZE@catch (std::exception & e)$
    ~EMPHASIZE@{$     
        std::cout << "caught exception ... \n";
        std::cout << "end\n";
    ~EMPHASIZE@}$
}
return 0;
\end{console}


WARNING: Depending on your compiler, you might need to write
\texttt{std::exception} instead of \texttt{exception}. You might or might
not need to \texttt{\#include} exception.

\texttt{\EMPHASIZE{exception}} is a class that comes with your compiler. In this case I'm throwing a standard \texttt{exception} object. You can actually throw any object or any value. If you prefer, you can create your own class for an exception object to throw, or you can just throw any value (\texttt{int}, \texttt{double}, \texttt{bool}, etc.)

This is really handy. Why? Suppose you're writing a program without exceptions and the code looks like the following, i.e., \texttt{f()} calls \texttt{g()}, \texttt{g()} calls \texttt{h()}, and \texttt{h()} calls \texttt{i()}:

\begin{console}
void i()
{    
     ...
}

void h()
{    
     ...
     i();
     ...
}

void g()
{    
     ...
     h();
     ...
}

void f()
{    
     ...
     g();
     ...
}
\end{console}

Here's a diagram that describes the function call
relationship:
\begin{console}
f → g → h → i
\end{console}
where the → means "calls".

After it's done you analyze your code and realize that
some errors might occur in \texttt{i()}

\begin{console}
void i()
{    
     ...
     z = x / y; // oops what if y is zero???
     ...
}
\end{console}

Suppose you want \texttt{f()} to handle this error. Then you might need to
pass an error code back to \texttt{f()} like this (we return 0 for no
error and 1 if there's an error). This means that the
error code must pass from \texttt{i} to \texttt{h}, \texttt{h} must pass the
error code to \texttt{g}, and \texttt{g} must pass the error code to
\texttt{f}:

\begin{console}[commandchars=\~\@\$]
~EMPHASIZE@int$ i()
{    
     ...
     // return 1 if there's an error
     ~EMPHASIZE@if (y == 0)$
         ~EMPHASIZE@return 1;$

     z = x / y; // oops what if y is zero???

     ...

     ~EMPHASIZE@return 0;$ // return 0 if there's no error
}

~EMPHASIZE@int$ h()
{    
     ...
     ~EMPHASIZE@if (i() != 0) return 1;$
     ...
     ~EMPHASIZE@return 0;$
}

~EMPHASIZE@int$ g()
{    
     ...
     ~EMPHASIZE@if (h() != 0) return 1;$
     ...
     ~EMPHASIZE@return 0;$
}

void f()
{    
     ...
     ~EMPHASIZE@if (g() != 0)$
     {
          ~EMPHASIZE@// ... do some error$
          ~EMPHASIZE@// handling ...$
     }
     ...
}
\end{console}

Boy ... what a pain!!! Look at the amount of changes you need to make!!!
Not to worry ... exceptions to the rescue ...

\begin{consolethree}[escapeinside=||]%[commandchars=\~\@\$]
void i()
{    
     ...
     |\EMPHASIZE{if (y == 0) throw exception();}|
     z = x / y; // oops what if y is zero???
     ...
}

void h()
{    
     ...
     i();
     ...
}

void g()
{    
     ...
     h();
     ...
}

void f()
{    
     ...
     |\EMPHASIZE{try}|
     |\EMPHASIZE{\{}|
          g();
     |\EMPHASIZE{\}}|
     |\EMPHASIZE{catch (exception \& e)}|
     |\EMPHASIZE{\{}|
          |\EMPHASIZE{// ... do some error}|
          |\EMPHASIZE{// handling ...}|
     |\EMPHASIZE{\}}|
     ...
}
\end{consolethree}

Much cleaner!!!

The important thing is this ...

You only need to modify the function code that might run into an error
and the function that should handle the error. In other words you only
need to change \texttt{i()} and \texttt{f()}. We say that \texttt{i()} is the
thrower and \texttt{f()} is the catcher.

Now, just imagine the nightmare if you have a chain of 20 function
calls: You would need to modify 20 functions from the thrower all the
way back to the catcher.

In a real piece of software, the exception handling code can for
instance print a message to the user to let them know that an email was
sent to the company and bug fixes will come soon. The program can (for
instance) shut down or maybe restart.

