\begin{python0}
from solutions import *; clear()
\end{python0}
\sectionthree{Other Scenarios}

Suppose you have an \texttt{LongInt} class for the modeling and
computation of integer values that are larger than what the \texttt{int}
provides. Then one possible place where you want to raise an error
(besides memory allocation error) if the division-by-zero error.
Therefore your \texttt{LongInt} class might look like this:

\begin{console}[commandchars=\~\@\$]
// File: LongInt.h

~textbf@class DivByZeroException {};$

class LongInt
{
...
       LongInt & operator/=(const LongInt & i)
       {
              ~textbf@if (i == 0) throw DivByZeroException();$
              ...
              return (*this);
       }
       
...
};
\end{console}

Your main program might then look like this:

\begin{console}
// File: main.cpp
#include <iostream>
#include <exception>
#include "LongInt.h"

int main()
{   
    std::cout << "predict the next-day stock "
    << "price of Google!!!\n"
    << "enter \n"
    << "1. current price\n"
    << "2. your weight\n"
    << "3. today's highest temperature\n"
    << "4. some random number\n";
    try
    {
        LongInt I, J, K, L;
        std::cin >> I >> J >> K >> L;
        std::cout << (I + L * J / K) << std::endl;
    }
    catch (std::bad_alloc & e)
    {
        std::cout << "out of mem!!! buy more"
                  << "ram!!!\n";
    }
    catch (DivByZeroException & e)
    {
        std::cout << "oops ... highest temp "
        << "cannot be zero\n";
    }
}
\end{console}

Another common thing that is done when something severe has occured is
to log the error, i.e. write some useful error/debugging message to a
file, and possible send the file to the author of the program.
You've seen this before when a program unexpectedly
crashes and a window pops up, informs you that an error has occured and
asks you if you want to send the error to Microsoft (or some other
company). That program was probably contains exception handling.

